---
title: "Command-Line Interface"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Command-Line Interface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

## Overview

scConvert includes a standalone C binary for streaming HDF5 conversions
between h5ad, h5Seurat, h5mu, and Loom -- no R or Python runtime required.
For formats that need R (RDS, Zarr, SCE, SOMA), the `scConvert_cli()` wrapper
falls back to the R hub automatically.

## Quick start

Load the bundled PBMC demo data (500 cells, 9 cell types) and save it as
h5Seurat -- the starting point for CLI conversion.

```{r load_demo, eval = TRUE, fig.width = 7, fig.height = 5}
obj <- readRDS(system.file("extdata", "pbmc_demo.rds", package = "scConvert"))
DimPlot(obj, reduction = "umap", group.by = "seurat_annotations") +
  ggplot2::ggtitle("PBMC demo (500 cells)")
```

```{r save_h5seurat, eval = TRUE}
h5s_path <- tempfile(fileext = ".h5Seurat")
writeH5Seurat(obj, h5s_path, overwrite = TRUE, verbose = FALSE)
cat("Saved:", h5s_path, "\n")
```

## Using the C binary from the shell

The binary auto-detects conversion direction from file extensions.

```{bash, eval = FALSE}
# Build (one time)
cd /path/to/scConvert/src && make

# Convert
scconvert data.h5seurat data.h5ad
scconvert data.h5ad data.h5seurat --assay RNA --gzip 6
scconvert multimodal.h5mu output.h5seurat
scconvert data.h5ad data.loom
```

**Options:**

| Flag | Description | Default |
|------|-------------|---------|
| `--assay <name>` | Assay/modality name | `RNA` |
| `--gzip <level>` | Compression level (0--9) | `1` |
| `--overwrite` | Overwrite existing output | off |
| `--quiet` | Suppress progress messages | off |

## Using `scConvert_cli()` from R

The R wrapper tries the C binary first, then falls back to R streaming or the
Seurat hub. It works even if the binary is not compiled.

```{r cli_r_wrapper, eval = TRUE}
h5ad_path <- tempfile(fileext = ".h5ad")
scConvert_cli(h5s_path, h5ad_path, verbose = FALSE)
cat("Converted to:", h5ad_path, "\n")
```

Verify the round-tripped data is intact:

```{r verify, eval = TRUE, fig.width = 7, fig.height = 5}
obj_rt <- readH5AD(h5ad_path, verbose = FALSE)
cat("Cells:", ncol(obj_rt), "| Genes:", nrow(obj_rt), "\n")
cat("Reductions:", paste(Reductions(obj_rt), collapse = ", "), "\n")

FeaturePlot(obj_rt, features = "LYZ", reduction = "umap") +
  ggtitle("LYZ expression after CLI round-trip")
```

## Batch conversion

Convert all h5ad files in a directory:

```{r batch, eval = FALSE}
h5ad_files <- list.files(".", pattern = "\\.h5ad$", full.names = TRUE)
for (f in h5ad_files) {
  out <- sub("\\.h5ad$", ".h5seurat", f)
  scConvert_cli(f, out)
}
```

For formats not supported by the C binary (RDS, Zarr, SCE), the same function
works via the R hub:

```{r other_formats, eval = FALSE}
scConvert_cli("data.h5ad", "data.rds")
scConvert_cli("data.rds", "data.zarr")
scConvert_cli("data.h5ad", "data.zarr")
```

## Performance

The C binary uses direct chunk copy and sparse zero-copy to avoid
decompressing data. Median wall-clock seconds on synthetic sparse h5ad
(20K genes, 5% density, Apple M4 Max):

| Cells | R read (readH5AD) | R write (writeH5AD) | CLI h5ad to h5seurat | CLI h5seurat to h5ad |
|------:|-------------------:|--------------------:|---------------------:|---------------------:|
| 1,000 | 0.28 s | 0.61 s | 0.02 s | 0.02 s |
| 10,000 | 0.49 s | 1.6 s | 0.04 s | 0.03 s |
| 50,000 | 1.4 s | 6.0 s | 0.13 s | 0.16 s |
| 100,000 | 2.9 s | 11.7 s | 0.29 s | 0.26 s |

The CLI is 10--50x faster because it never constructs a Seurat object. For
loading data into R for analysis, use `readH5AD()` or `readH5Seurat()`.

## Building the C binary

The binary is optional -- `scConvert_cli()` works without it.

```{bash, eval = FALSE}
# macOS (Homebrew)
brew install hdf5
cd /path/to/scConvert/src && make

# Ubuntu/Debian
sudo apt-get install libhdf5-dev
cd /path/to/scConvert/src && make

# Copy to PATH
cp src/scconvert ~/bin/
```

The binary links against libhdf5 and libz only. No other dependencies.

## Clean up

```{r cleanup, include = FALSE}
unlink(c(h5s_path, h5ad_path))
```
