---
title: "TileDB-SOMA and CELLxGENE Census"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TileDB-SOMA and CELLxGENE Census}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
has_soma <- requireNamespace("tiledbsoma", quietly = TRUE)
```

## Introduction

[TileDB-SOMA](https://github.com/single-cell-data/TileDB-SOMA) is a
cloud-native, array-based format for single-cell data. It is the storage
backend behind [CELLxGENE Census](https://chanzuckerberg.github.io/cellxgene-census/),
which hosts 61M+ cells across 900+ datasets. SOMA supports efficient slicing
by cells or features without loading the full dataset, making it ideal for
working with large atlases.
scConvert provides `writeSOMA()` and `readSOMA()` for Seurat interoperability.
Both functions require the
[tiledbsoma](https://github.com/single-cell-data/TileDB-SOMA) R package.

## Load demo data

We use a 500-cell PBMC dataset that ships with scConvert. It has 9 annotated
cell types, PCA/UMAP embeddings, and neighbor graphs.

```{r load-data, eval = TRUE}
pbmc <- readRDS(system.file("extdata", "pbmc_demo.rds", package = "scConvert"))
pbmc
```

```{r dimplot-original, eval = TRUE, fig.width = 7, fig.height = 5}
DimPlot(pbmc, reduction = "umap", group.by = "seurat_annotations",
        label = TRUE, pt.size = 0.8) +
  ggtitle("PBMC demo -- 9 cell types") + NoLegend()
```

## Write to SOMA

Save the Seurat object as a SOMA experiment. If tiledbsoma is not installed,
this section shows the code without running it.

```{r write-soma, eval = has_soma}
soma_uri <- file.path(tempdir(), "pbmc_demo.soma")

# Ensure graphs have a valid DefaultAssay (required by tiledbsoma)
for (gn in names(pbmc@graphs)) {
  if (length(DefaultAssay(pbmc@graphs[[gn]])) == 0) {
    DefaultAssay(pbmc@graphs[[gn]]) <- DefaultAssay(pbmc)
  }
}

writeSOMA(pbmc, uri = soma_uri, overwrite = TRUE)
cat("SOMA experiment written to:", soma_uri, "\n")
```

## Read back from SOMA

```{r read-soma, eval = has_soma}
pbmc_rt <- readSOMA(soma_uri)
cat("Cells:", ncol(pbmc_rt), "| Genes:", nrow(pbmc_rt), "\n")
cat("Metadata columns:", paste(colnames(pbmc_rt[[]]), collapse = ", "), "\n")
```

## Compare original and round-trip

Side-by-side UMAP plots confirm that cluster labels and coordinates survive
the SOMA round-trip.

```{r compare-dimplot, eval = has_soma, fig.width = 12, fig.height = 5}
library(patchwork)

p1 <- DimPlot(pbmc, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.8) +
  ggtitle("Original") + NoLegend()

p2 <- DimPlot(pbmc_rt, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.8) +
  ggtitle("After SOMA round-trip") + NoLegend()

p1 + p2
```

LYZ is a strong monocyte marker -- the violin plot below shows its expression
distribution across all 9 cell types loaded from the SOMA store.

```{r vlnplot, eval = has_soma, fig.width = 7, fig.height = 5}
Idents(pbmc_rt) <- "seurat_annotations"
VlnPlot(pbmc_rt, features = "LYZ", pt.size = 0) +
  ggtitle("LYZ expression by cell type (from SOMA)") + NoLegend()
```

### Quick fidelity check

```{r fidelity-check, eval = has_soma}
stopifnot(ncol(pbmc_rt) == ncol(pbmc))
stopifnot(nrow(pbmc_rt) == nrow(pbmc))
cat("Dimensions match:", ncol(pbmc_rt), "cells x", nrow(pbmc_rt), "genes\n")

shared_meta <- intersect(colnames(pbmc[[]]), colnames(pbmc_rt[[]]))
cat("Shared metadata columns:", length(shared_meta), "\n")
```

## CELLxGENE Census access

CELLxGENE Census stores the largest public atlas of single-cell data as a
single SOMA experiment. You can query it with `readSOMA()` and then convert
the result to any format.

```{r census, eval = FALSE}
library(cellxgene.census)

census <- open_soma(census_version = "stable")
human_uri <- census$get("census_data")$get("homo_sapiens")$uri

# Read a subset directly as a Seurat object
tcells <- readSOMA(
  uri = human_uri,
  measurement = "RNA",
  obs_query = "cell_type == 'T cell' & tissue_general == 'blood'"
)

# Save locally in any format
writeH5AD(tcells, "census_tcells.h5ad")
saveRDS(tcells, "census_tcells.rds")
```

## Pair converters

Direct conversion functions are available for all supported format pairs:

```{r pair-converters, eval = FALSE}
# SOMA <-> h5ad
H5ADToSOMA("data.h5ad", "data.soma")
SOMAToH5AD("data.soma", "data.h5ad")

# SOMA <-> Zarr
ZarrToSOMA("data.zarr", "data.soma")
SOMAToZarr("data.soma", "data.zarr")

# Or use the universal dispatcher
scConvert("data.h5ad", dest = "data.soma", overwrite = TRUE)
```

```{r cleanup, include = FALSE}
if (has_soma) {
  unlink(file.path(tempdir(), "pbmc_demo.soma"), recursive = TRUE)
}
```

## Session Info

```{r session-info, eval = TRUE}
sessionInfo()
```
