---
title: "Convert to Loom Format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convert to Loom Format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

The [Loom format](http://loompy.org/) is an HDF5-based file format used by
loompy and velocyto/scVelo for RNA velocity workflows. scConvert provides
`writeLoom()` and `readLoom()` for converting Seurat objects to and from Loom
with no external dependencies.

## Load the demo data

We use the shipped 500-cell PBMC demo dataset with PCA, UMAP, clusters, and
nine annotated cell types.

```{r load-data, eval = TRUE}
pbmc <- readRDS(system.file("extdata", "pbmc_demo.rds", package = "scConvert"))
pbmc
```

```{r dimplot-original, eval = TRUE}
DimPlot(pbmc, reduction = "umap", group.by = "seurat_annotations",
        label = TRUE, pt.size = 0.5) + NoLegend()
```

Let us also look at CD14, a marker for monocytes, before conversion:

```{r featureplot-original, eval = TRUE}
FeaturePlot(pbmc, features = "CD14", pt.size = 0.5)
```

## Write to Loom

`writeLoom()` saves the default assay's expression data as the main matrix, with
cell metadata as column attributes and gene metadata as row attributes.
Dimensional reductions are stored as additional column attributes.

```{r write-loom, eval = TRUE}
loom_path <- tempfile(fileext = ".loom")
writeLoom(pbmc, filename = loom_path, overwrite = TRUE)
cat("Loom file size:", round(file.size(loom_path) / 1e6, 1), "MB\n")
```

## Read it back

`readLoom()` loads the Loom file as a Seurat object. The `cells` and `features`
parameters control which column/row attributes are used for cell and gene names
(defaults: `"CellID"` and `"Gene"`).

```{r read-loom, eval = TRUE}
pbmc_rt <- readLoom(loom_path)
pbmc_rt
```

```{r verify-dims, eval = TRUE}
cat("Original:", ncol(pbmc), "cells,", nrow(pbmc), "genes\n")
cat("Loaded:  ", ncol(pbmc_rt), "cells,", nrow(pbmc_rt), "genes\n")
cat("Metadata preserved:", paste(
  intersect(colnames(pbmc[[]]), colnames(pbmc_rt[[]])), collapse = ", "), "\n")
```

## Compare expression

Expression values and cell metadata are preserved through the round-trip. Loom
stores dimensional reductions as column attributes, so UMAP and PCA coordinates
are available after loading.

```{r compare-dimplot, eval = TRUE, fig.width = 10, fig.height = 5}
library(patchwork)
p1 <- DimPlot(pbmc, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.5) + NoLegend() + ggtitle("Original")
p2 <- DimPlot(pbmc_rt, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.5) + NoLegend() + ggtitle("From Loom")
p1 + p2
```

The LYZ expression pattern (a monocyte marker) is identical before and after
conversion:

```{r featureplot-compare, eval = TRUE, fig.width = 10, fig.height = 5}
p1 <- FeaturePlot(pbmc, features = "LYZ", pt.size = 0.5) + ggtitle("LYZ - Original")
p2 <- FeaturePlot(pbmc_rt, features = "LYZ", pt.size = 0.5) + ggtitle("LYZ - From Loom")
p1 + p2
```

A violin plot confirms the per-cluster distribution is preserved:

```{r vlnplot, eval = TRUE}
VlnPlot(pbmc_rt, features = "CD14", group.by = "seurat_annotations", pt.size = 0) +
  NoLegend()
```

## What is preserved

Loom is a simpler format than h5ad or h5Seurat. Here is a summary of what
round-trips and what does not:

| Component | Preserved? | Notes |
|-----------|:----------:|-------|
| Expression matrix | Yes | Stored as `/matrix` |
| Raw counts | Yes | Stored in `/layers/counts` |
| Cell metadata | Yes | Each column becomes a `/col_attrs` entry |
| Gene metadata | Yes | Each column becomes a `/row_attrs` entry |
| PCA / UMAP embeddings | Yes | Stored as column attributes |
| Nearest-neighbor graphs | No | Not native to Loom; recompute with `FindNeighbors()` |

## Using scConvert() for format conversion

You can also convert Loom files to other formats using the `scConvert()`
dispatcher without loading into R:

```{r scconvert-loom, eval = TRUE}
h5ad_path <- tempfile(fileext = ".h5ad")
scConvert(loom_path, dest = h5ad_path, overwrite = TRUE)
cat("Converted Loom to h5ad:", round(file.size(h5ad_path) / 1e6, 1), "MB\n")
unlink(h5ad_path)
```

## Python interop (optional)

If you have Python with loompy installed, you can read the Loom file directly.

```{python loompy-read, eval = FALSE}
# Requires Python with loompy installed
import loompy

with loompy.connect("pbmc.loom") as ds:
    print(f"Shape: {ds.shape[0]} genes x {ds.shape[1]} cells")
    print(f"Row attributes: {list(ds.ra.keys())}")
    print(f"Column attributes: {list(ds.ca.keys())[:10]}")
    print(f"Layers: {list(ds.layers.keys())}")
```

You can also load Loom files into scanpy as AnnData objects:

```{python scanpy-loom, eval = FALSE}
# Requires Python with scanpy installed
import scanpy as sc

adata = sc.read_loom("pbmc.loom", sparse=True, cleanup=False)
print(adata)
```

## Clean up

```{r cleanup, eval = TRUE}
unlink(loom_path)
```
