---
title: "Converting Between Seurat and TileDB-SOMA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Converting Between Seurat and TileDB-SOMA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  out.width = "100%",
  purl = FALSE
)
has_tiledbsoma <- requireNamespace("tiledbsoma", quietly = TRUE)
```

## What is TileDB-SOMA?

[TileDB-SOMA](https://github.com/single-cell-data/TileDB-SOMA) is a
cloud-native, array-based format for storing and querying single-cell data at
scale. It is the storage backend behind
[CELLxGENE Census](https://chanzuckerberg.github.io/cellxgene-census/), the
largest public atlas of single-cell data (61M+ cells across 900+ datasets).

Key properties of SOMA:

- **Columnar storage** backed by TileDB arrays -- efficient slicing by cells
  or features without loading the full dataset
- **Cloud-native**: local paths, S3, GCS, and TileDB Cloud URIs
- **Query filtering**: select cells by metadata predicates at the storage level
- **Multi-measurement**: RNA, ATAC, and protein assays in one experiment

scConvert provides `readSOMA()` and `writeSOMA()` for Seurat interoperability,
plus pair converters for all supported formats. All SOMA functions require the
[tiledbsoma](https://github.com/single-cell-data/TileDB-SOMA) R package.

```{r packages, eval = has_tiledbsoma}
library(scConvert)
library(Seurat)
```

## Writing a Seurat object to SOMA

We start with the shipped `pbmc_small` dataset, write it to a SOMA experiment,
then read it back to verify the roundtrip.

```{r write_soma, eval = has_tiledbsoma}
# Load the shipped demo data
pbmc <- readRDS(system.file("testdata", "pbmc_small.rds", package = "scConvert"))
pbmc
```

Use `writeSOMA()` to save any Seurat object as a SOMA experiment:

```{r write_soma_run, eval = has_tiledbsoma}
soma_uri <- file.path(tempdir(), "pbmc_small.soma")

# Ensure graphs have a valid DefaultAssay (required by tiledbsoma)
for (gn in names(pbmc@graphs)) {
  if (length(DefaultAssay(pbmc@graphs[[gn]])) == 0) {
    DefaultAssay(pbmc@graphs[[gn]]) <- DefaultAssay(pbmc)
  }
}

writeSOMA(pbmc, uri = soma_uri, overwrite = TRUE)
cat("SOMA experiment written to:", soma_uri, "\n")
```

`writeSOMA()` delegates to `tiledbsoma::write_soma()`, which converts all
assays, metadata, embeddings, and graphs. Each Seurat assay becomes a separate
SOMA measurement.

## Reading from SOMA

Use `readSOMA()` to load a SOMA experiment as a Seurat object:

```{r read_soma, eval = has_tiledbsoma}
obj <- readSOMA(soma_uri)
obj
```

### Verify roundtrip fidelity

```{r verify_roundtrip, eval = has_tiledbsoma}
# Cell and gene counts match
cat("Original cells:", ncol(pbmc), " Roundtrip cells:", ncol(obj), "\n")
cat("Original genes:", nrow(pbmc), " Roundtrip genes:", nrow(obj), "\n")
stopifnot(ncol(obj) == ncol(pbmc))
stopifnot(nrow(obj) == nrow(pbmc))

# Metadata columns are preserved
orig_meta_cols <- sort(names(pbmc@meta.data))
rt_meta_cols <- sort(names(obj@meta.data))
shared_cols <- intersect(orig_meta_cols, rt_meta_cols)
cat("Shared metadata columns:", length(shared_cols), "/", length(orig_meta_cols), "\n")
cat("Columns:", paste(shared_cols, collapse = ", "), "\n")
```

### Filtering cells and features

`readSOMA()` supports server-side filtering via value filter queries, which is
particularly useful for large atlases:

```{r read_filtered, eval = FALSE}
# Filter by cell type and tissue
obj <- readSOMA(
  "path/to/experiment",
  obs_query = "cell_type == 'T cell' & tissue == 'blood' & disease == 'normal'",
  obs_column_names = c("cell_type", "tissue", "donor_id")
)

# Read a specific measurement (default is "RNA")
obj <- readSOMA("path/to/experiment", measurement = "ATAC")
```

### What gets read

`readSOMA()` maps the following SOMA components to Seurat:

| SOMA Source | Seurat Destination | Description |
|-------------|-------------------|-------------|
| `obs` | `meta.data` | Cell metadata (factors, numerics, strings) |
| `ms[measurement]/var` | Feature metadata | Gene-level annotations |
| `ms[measurement]/X/data` | Default assay counts | Expression matrix (sparse) |
| `ms[measurement]/obsm/*` | `reductions` | PCA, UMAP, etc. |
| `ms[measurement]/obsp/*` | `graphs` | Neighbor graphs (SNN, distances) |
| Other measurements | Additional assays | Multi-modal support |

The `soma_joinid` column (internal to TileDB-SOMA) is automatically removed
from cell and feature metadata.

## Direct pair converters

scConvert provides direct conversion functions between SOMA and all other
supported formats. All route through a Seurat intermediate.

### SOMA and h5ad

```{r pair_h5ad, eval = has_tiledbsoma}
h5ad_path <- file.path(tempdir(), "demo.h5ad")
soma_from_h5ad <- file.path(tempdir(), "from_h5ad.soma")
h5ad_from_soma <- file.path(tempdir(), "from_soma.h5ad")

# Write pbmc_small to h5ad first
writeH5AD(pbmc, filename = h5ad_path, overwrite = TRUE)

# h5ad -> SOMA
H5ADToSOMA(h5ad_path, soma_from_h5ad, overwrite = TRUE)

# SOMA -> h5ad
SOMAToH5AD(soma_from_h5ad, h5ad_from_soma, overwrite = TRUE)
```

Verify the h5ad roundtrip:

```{r verify_h5ad_roundtrip, eval = has_tiledbsoma}
rt <- readH5AD(h5ad_from_soma)
cat("h5ad -> SOMA -> h5ad roundtrip:\n")
cat("  Cells:", ncol(rt), " Genes:", nrow(rt), "\n")
stopifnot(ncol(rt) == ncol(pbmc))
stopifnot(nrow(rt) == nrow(pbmc))
```

### SOMA and h5Seurat

```{r pair_h5seurat, eval = has_tiledbsoma}
h5seurat_path <- file.path(tempdir(), "demo.h5seurat")
soma_from_h5s <- file.path(tempdir(), "from_h5seurat.soma")
h5s_from_soma <- file.path(tempdir(), "from_soma.h5seurat")

# Write to h5Seurat, then convert through SOMA
writeH5Seurat(pbmc, filename = h5seurat_path, overwrite = TRUE)
H5SeuratToSOMA(h5seurat_path, soma_from_h5s, overwrite = TRUE)
SOMAToH5Seurat(soma_from_h5s, h5s_from_soma, overwrite = TRUE)

rt2 <- readH5Seurat(h5s_from_soma)
cat("h5Seurat -> SOMA -> h5Seurat roundtrip:\n")
cat("  Cells:", ncol(rt2), " Genes:", nrow(rt2), "\n")
stopifnot(ncol(rt2) == ncol(pbmc))
```

### Other format pairs

Converters also exist for Loom, Zarr, and h5mu -- each follows the same
`{Format}ToSOMA()` / `SOMATo{Format}()` pattern:

```{r pair_others, eval = FALSE}
LoomToSOMA("data.loom", "data.soma")    # Loom -> SOMA
ZarrToSOMA("data.zarr", "data.soma")    # Zarr -> SOMA
H5MUToSOMA("data.h5mu", "data.soma")    # h5mu -> SOMA
SOMAToLoom("data.soma", "output.loom")   # SOMA -> Loom
SOMAToZarr("data.soma", "output.zarr")   # SOMA -> Zarr
SOMAToH5MU("data.soma", "output.h5mu")  # SOMA -> h5mu
```

## CLI and dispatcher usage

The C CLI binary handles HDF5-to-HDF5 conversions only, so SOMA is not
supported in the compiled binary. However, the R-level `scConvert_cli()` and
the universal `scConvert()` dispatcher auto-detect SOMA from the `.soma`
extension and route through the R hub:

```{r cli, eval = FALSE}
# scConvert_cli auto-dispatches
scConvert_cli("data.h5ad", "data.soma")
scConvert_cli("data.soma", "output.h5ad")

# Universal dispatcher
scConvert("data.h5ad", dest = "data.soma", overwrite = TRUE)
scConvert("data.soma", dest = "output.h5seurat", overwrite = TRUE)
```

## CELLxGENE Census access

[CELLxGENE Census](https://chanzuckerberg.github.io/cellxgene-census/) hosts
61M+ cells as a single SOMA experiment. Query it with `readSOMA()`, then
convert the result to any format:

```{r census, eval = FALSE}
library(cellxgene.census)

census <- open_soma(census_version = "stable")
human_uri <- census$get("census_data")$get("homo_sapiens")$uri

# Read a subset directly as a Seurat object
tcells <- readSOMA(
  uri = human_uri,
  measurement = "RNA",
  obs_query = "cell_type == 'T cell' & tissue_general == 'blood'",
  obs_column_names = c("cell_type", "tissue", "donor_id")
)

# Save locally in any format
writeH5AD(tcells, "census_tcells.h5ad")
writeSOMA(tcells, "census_tcells.soma")
```

See the [cellxgene-census documentation](https://chanzuckerberg.github.io/cellxgene-census/)
for more details on the Census API.

## Data mapping reference

| Seurat | SOMA | Direction | Notes |
|--------|------|-----------|-------|
| Default assay counts | `ms[RNA]/X/data` | Read/Write | Sparse matrix |
| `meta.data` | `obs` | Read/Write | Factors, numerics, strings |
| Feature metadata | `ms[RNA]/var` | Read/Write | Gene annotations |
| `Embeddings(, "pca")` | `ms[RNA]/obsm/X_pca` | Read/Write | Dense array |
| `Embeddings(, "umap")` | `ms[RNA]/obsm/X_umap` | Read/Write | Dense array |
| `Graphs(, "RNA_snn")` | `ms[RNA]/obsp/RNA_snn` | Read/Write | Sparse matrix |
| Additional assays | Other measurements | Read/Write | Multi-modal support |

## Limitations

- **Requires tiledbsoma**: install with `install.packages("tiledbsoma")`.
  `writeSOMA()` further requires `tiledbsoma::write_soma()` to be available.
- **No streaming conversion**: unlike h5ad/zarr converters, SOMA always routes
  through a Seurat intermediate (TileDB arrays differ fundamentally from HDF5).
- **No C binary path**: the compiled C CLI handles HDF5-to-HDF5 only.
- **scale.data** is not written to SOMA -- re-run `ScaleData()` after loading.
- **Active identity** (`Idents()`) is not stored -- set manually after reading.

## See also

- `vignette("convert-anndata")` -- Converting between Seurat and h5ad format
- `vignette("cli-usage")` -- Command-line interface for batch conversions
- `vignette("convert-zarr")` -- Converting between Seurat and Zarr format

```{r cleanup, include = FALSE, purl = FALSE}
# Remove temporary files created during this vignette
cleanup_paths <- c(
  file.path(tempdir(), "pbmc_small.soma"),
  file.path(tempdir(), "demo.h5ad"),
  file.path(tempdir(), "from_h5ad.soma"),
  file.path(tempdir(), "from_soma.h5ad"),
  file.path(tempdir(), "demo.h5seurat"),
  file.path(tempdir(), "from_h5seurat.soma"),
  file.path(tempdir(), "from_soma.h5seurat")
)
for (p in cleanup_paths) {
  if (dir.exists(p)) unlink(p, recursive = TRUE)
  if (file.exists(p)) file.remove(p)
}
```

## Session Info

```{r session, eval = TRUE, purl = FALSE}
sessionInfo()
```
