---
title: "Convert Between Seurat and AnnData (h5ad)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convert Between Seurat and AnnData (h5ad)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

The h5ad format is the standard file format for the Python single-cell ecosystem
(scanpy, scvi-tools, CellxGene). scConvert provides `writeH5AD()` and
`readH5AD()` for round-trip conversion between Seurat objects and h5ad files
with no Python dependency.

## Load the demo data

scConvert ships a 500-cell PBMC dataset with PCA, UMAP, clusters, and nine
annotated cell types. We will use this dataset throughout the vignette to
demonstrate the conversion workflow.

```{r load-data, eval = TRUE}
pbmc <- readRDS(system.file("extdata", "pbmc_demo.rds", package = "scConvert"))
pbmc
```

The dataset has been fully processed with clustering and dimensional reductions:

```{r dimplot-original, eval = TRUE}
DimPlot(pbmc, reduction = "umap", group.by = "seurat_annotations",
        label = TRUE, pt.size = 0.5) + NoLegend()
```

We can also look at a marker gene. LYZ is highly expressed in monocytes:

```{r featureplot-original, eval = TRUE}
FeaturePlot(pbmc, features = "LYZ", pt.size = 0.5)
```

## Write to h5ad

`writeH5AD()` saves the Seurat object as an h5ad file. The normalized data
matrix goes to `X`, raw counts go to `raw/X`, and all metadata, reductions, and
graphs are preserved.

```{r write-h5ad, eval = TRUE}
h5ad_path <- tempfile(fileext = ".h5ad")
writeH5AD(pbmc, h5ad_path, overwrite = TRUE)
cat("h5ad file size:", round(file.size(h5ad_path) / 1e6, 1), "MB\n")
```

## Read it back

`readH5AD()` loads the h5ad file directly into a Seurat object. No intermediate
h5Seurat file is needed.

```{r read-h5ad, eval = TRUE}
pbmc_rt <- readH5AD(h5ad_path)
pbmc_rt
```

We can confirm that the key components were preserved:

```{r verify-contents, eval = TRUE}
cat("Cells:", ncol(pbmc_rt), "\n")
cat("Genes:", nrow(pbmc_rt), "\n")
cat("Reductions:", paste(Reductions(pbmc_rt), collapse = ", "), "\n")
cat("Metadata cols:", paste(colnames(pbmc_rt[[]]), collapse = ", "), "\n")
```

## Compare original and round-trip

The UMAP coordinates, cluster labels, and expression values are preserved
through the conversion.

```{r compare-dimplot, eval = TRUE, fig.width = 10, fig.height = 5}
library(patchwork)
p1 <- DimPlot(pbmc, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.5) + NoLegend() + ggtitle("Original")
p2 <- DimPlot(pbmc_rt, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.5) + NoLegend() + ggtitle("Round-trip")
p1 + p2
```

The expression pattern for LYZ is identical before and after conversion:

```{r compare-featureplot, eval = TRUE, fig.width = 10, fig.height = 5}
p1 <- FeaturePlot(pbmc, features = "LYZ", pt.size = 0.5) + ggtitle("LYZ - Original")
p2 <- FeaturePlot(pbmc_rt, features = "LYZ", pt.size = 0.5) + ggtitle("LYZ - Round-trip")
p1 + p2
```

We can also confirm that the violin plot distribution matches:

```{r vlnplot, eval = TRUE}
VlnPlot(pbmc_rt, features = "LYZ", group.by = "seurat_annotations", pt.size = 0) +
  NoLegend()
```

## One-liner with scConvert()

For quick format conversion without loading into R, use the `scConvert()`
dispatcher. It handles any supported pair of formats automatically:

```{r scconvert-oneliner, eval = TRUE}
h5seurat_path <- tempfile(fileext = ".h5seurat")
scConvert(h5ad_path, dest = h5seurat_path, overwrite = TRUE)
cat("Converted to h5Seurat:", round(file.size(h5seurat_path) / 1e6, 1), "MB\n")
```

The dispatcher detects the source and destination formats from file extensions
and picks the most efficient conversion path.

## Layer mapping

During conversion, scConvert maps data between Seurat and h5ad as follows:

| Seurat | h5ad | Description |
|--------|------|-------------|
| `data` layer | `X` | Normalized expression matrix |
| `counts` layer | `raw/X` | Raw counts |
| `meta.data` | `obs` | Cell metadata (categoricals become factors) |
| Feature metadata | `var` | Gene-level annotations |
| `Reductions(obj)` | `obsm/X_pca`, `obsm/X_umap` | Dimensional reductions |
| `Graphs(obj)` | `obsp/connectivities`, `obsp/distances` | Neighbor graphs |
| `misc` | `uns` | Unstructured annotations |

## Python interop (optional)

If you have Python with scanpy installed, you can read the h5ad file directly.
The file produced by `writeH5AD()` is fully compatible with scanpy, scvi-tools,
and CellxGene.

```{python scanpy-read, eval = FALSE}
# Requires Python with scanpy installed
import scanpy as sc

adata = sc.read_h5ad("pbmc.h5ad")
print(adata)

# Visualize with cluster annotations
sc.pl.umap(adata, color="seurat_annotations")

# Expression patterns are preserved
sc.pl.umap(adata, color="LYZ", use_raw=False)
```

## Clean up

```{r cleanup, eval = TRUE}
unlink(h5ad_path)
unlink(h5seurat_path)
```
