---
title: "Working with SpatialData (.spatialdata.zarr)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with SpatialData (.spatialdata.zarr)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
)
```

## Overview

[SpatialData](https://spatialdata.scverse.org/) is the scverse standard for
representing spatial omics data. It uses a Zarr-based directory layout that
combines OME-NGFF images, geometric shapes, transcript-level point
annotations, segmentation labels, and anndata expression tables into a single
hierarchical store. Technologies such as Visium, MERFISH, Xenium, Slide-seq,
CODEX, and Stereo-seq all have SpatialData representations, making it a
unifying format across the spatial omics ecosystem.

scConvert provides native R support for reading and writing SpatialData stores
with no Python dependency, and includes direct pair converters for moving data
between SpatialData and other formats (h5ad, h5Seurat, Zarr).

```{r packages}
library(Seurat)
library(scConvert)
```

## SpatialData store layout

A `.spatialdata.zarr` directory follows this structure:

```
sample.spatialdata.zarr/
+-- .zattrs              # {"spatialdata_attrs": {"version": "0.2.0"}}
+-- .zgroup              # {"zarr_format": 2}
+-- images/              # OME-NGFF multiscale tissue images
+-- labels/              # Segmentation masks (cell/nucleus outlines)
+-- points/              # Point annotations (e.g., transcript coordinates)
+-- shapes/              # Geometric shapes (spot positions with radii)
+-- tables/              # AnnData expression tables (one or more)
    +-- table/           # Default table (standard anndata zarr)
```

The `tables/table/` subdirectory is a standard anndata zarr store, so
scConvert's existing Zarr infrastructure handles the expression data, metadata,
embeddings, and graphs. The SpatialData-specific elements (shapes, points,
images, labels) are handled by dedicated readers and writers.

## Reading SpatialData

Use `readSpatialData()` to load a SpatialData zarr store into a Seurat object:

```{r read_basic}
obj <- readSpatialData("visium.spatialdata.zarr")
obj
```

The reader extracts the following components:

| SpatialData Element | Seurat Destination | Notes |
|--------------------|--------------------|-------|
| `tables/table/` | Expression assay, metadata, reductions | Via `readZarr()` |
| `shapes/{region}/` | Spatial coordinates (FOV/image) | x, y, radius columns |
| `points/{region}/` | `misc$points_{name}` | Transcript-level data |
| `images/{name}/` | Image object (if coords present) | OME-NGFF, normalized to [0,1] |
| `labels/` | `misc$__spatialdata_labels__` | Names only (see Limitations) |

SpatialData attributes (`region`, `region_key`, `instance_key`) are stored in
`misc$__spatialdata_attrs__` and used during write-back to ensure round-trip
fidelity.

### Specifying a table

SpatialData stores can contain multiple tables. Use the `table` argument to
select which one to read:

```{r read_table}
# Read the default table
obj <- readSpatialData("multi_table.spatialdata.zarr", table = "table")

# Read a specific table
obj <- readSpatialData("multi_table.spatialdata.zarr", table = "rna_counts")
```

### Skipping images

For large stores where only the expression data and coordinates are needed,
set `images = FALSE` to skip OME-NGFF image reading:

```{r read_no_images}
obj <- readSpatialData("large_sample.spatialdata.zarr", images = FALSE)
```

## Writing SpatialData

Use `writeSpatialData()` to write a Seurat object as a SpatialData zarr store:

```{r write_basic}
writeSpatialData(obj, "output.spatialdata.zarr", overwrite = TRUE)
```

The writer creates the full SpatialData directory layout:

| Seurat Source | SpatialData Destination | Notes |
|--------------|------------------------|-------|
| Expression, metadata, reductions | `tables/table/` | Via `writeZarr()` |
| Tissue coordinates | `shapes/{region}/` | DataFrame with x, y, radius |
| `misc$points_*` | `points/{name}/` | Preserved from read |
| Image arrays | `images/{name}/` | OME-NGFF with multiscales metadata |

The region name is auto-detected from the Seurat image name, or you can set it
explicitly:

```{r write_region}
writeSpatialData(obj, "output.spatialdata.zarr",
                 region = "spots", overwrite = TRUE)
```

If the object has metadata columns `spatial_x` and `spatial_y` but no Seurat
image object, `writeSpatialData()` uses those as the shape coordinates.

## Direct pair converters

scConvert provides six direct conversion functions that combine reading and
writing in a single call, without requiring you to manage intermediate Seurat
objects:

```{r direct_converters}
# SpatialData <-> h5ad
SpatialDataToH5AD("sample.spatialdata.zarr", "sample.h5ad")
H5ADToSpatialData("sample.h5ad", "sample.spatialdata.zarr")

# SpatialData <-> h5Seurat
SpatialDataToH5Seurat("sample.spatialdata.zarr", "sample.h5seurat")
H5SeuratToSpatialData("sample.h5seurat", "sample.spatialdata.zarr")

# SpatialData <-> standard Zarr (anndata zarr without SpatialData wrapper)
SpatialDataToZarr("sample.spatialdata.zarr", "sample.zarr")
ZarrToSpatialData("sample.zarr", "sample.spatialdata.zarr")
```

All six functions accept `overwrite = TRUE` and `verbose = TRUE/FALSE`
arguments. The `SpatialDataToH5AD`, `SpatialDataToH5Seurat`, and
`SpatialDataToZarr` functions also accept a `table` argument to specify which
table to extract from the store.

These converters also work through the `scConvert()` dispatcher:

```{r scconvert_dispatch}
scConvert("sample.spatialdata.zarr", dest = "sample.h5ad", overwrite = TRUE)
scConvert("sample.h5ad", dest = "output.spatialdata.zarr", overwrite = TRUE)
```

## Python interoperability

SpatialData stores written by scConvert are readable by the Python
[spatialdata](https://spatialdata.scverse.org/) library and its ecosystem
(squidpy, napari-spatialdata):

```python
import spatialdata as sd

sdata = sd.read_zarr("output.spatialdata.zarr")
print(sdata)
print(sdata.tables)    # Expression tables
print(sdata.shapes)    # Spot/cell geometries

# Visualize with spatialdata-plot
sdata.pl.render_shapes().pl.show()
```

Conversely, SpatialData stores created by the Python `spatialdata` library
(e.g., from `squidpy.datasets` or `spatialdata-io` readers) can be loaded
directly with `readSpatialData()`.

## Known limitations

- **Labels/segmentation masks**: scConvert records the names of label layers
  found in `labels/` but does not read the full mask arrays. These are
  typically large (cell/nucleus segmentation at pixel resolution) and have no
  natural representation in Seurat. Label names are preserved in
  `misc$__spatialdata_labels__` for metadata purposes.

- **Multiple tables**: Only one table is read per call. If your SpatialData
  store contains multiple tables (e.g., separate RNA and protein assays), call
  `readSpatialData()` once per table and merge the resulting Seurat objects.

- **Large images**: OME-NGFF images larger than 100 million pixels are skipped
  to avoid memory issues. Use `images = FALSE` and work with coordinates only,
  or downsample the image externally before reading.

- **Coordinate transforms**: SpatialData supports affine and sequence
  coordinate transformations between elements. These are not currently applied
  during reading; coordinates are used as-is from the zarr arrays.

## See also

- [Spatial Transcriptomics: Visium](spatial-visium.html) -- Visium-specific
  workflows with tissue images and squidpy validation
- [Spatial Technologies](spatial-technologies.html) -- support for MERFISH,
  Slide-seq, Xenium, CODEX, and other platforms
- [Conversions: Seurat and Zarr](convert-zarr.html) -- general Zarr format
  details, streaming conversion, and Python interop

## Session Info

```{r session, eval = TRUE}
sessionInfo()
```
