---
title: "Convert Between Seurat and AnnData (h5ad)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convert Between Seurat and AnnData (h5ad)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

The h5ad format is the standard file format for the Python single-cell ecosystem
(scanpy, scvi-tools, CELLxGENE). scConvert reads and writes h5ad natively in R
with no Python dependency, making it easy to move data between Seurat and the
entire scverse stack.

## Read an h5ad file

scConvert ships a 500-cell PBMC dataset in h5ad format. `readH5AD()` loads it
directly into a Seurat object -- no intermediate conversion step is needed.

```{r read-h5ad, eval = TRUE}
h5ad_file <- system.file("extdata", "pbmc_demo.h5ad", package = "scConvert")
pbmc <- readH5AD(h5ad_file)
pbmc
```

## Inspect the loaded object

The reader reconstructs the full Seurat object from the h5ad contents,
including metadata, dimensional reductions, and neighbor graphs.

```{r inspect, eval = TRUE}
cat("Cells:", ncol(pbmc), "\n")
cat("Genes:", nrow(pbmc), "\n")
cat("Reductions:", paste(Reductions(pbmc), collapse = ", "), "\n")
cat("Metadata columns:", paste(colnames(pbmc[[]]), collapse = ", "), "\n")
```

The nine annotated cell types are preserved as a factor column:

```{r dimplot, eval = TRUE}
DimPlot(pbmc, reduction = "umap", group.by = "seurat_annotations",
        label = TRUE, pt.size = 0.5) + NoLegend()
```

We can also visualize a marker gene. LYZ is highly expressed in monocytes:

```{r featureplot, eval = TRUE}
FeaturePlot(pbmc, features = "LYZ", pt.size = 0.5)
```

## Write a Seurat object to h5ad

To go in the other direction, load a Seurat object and write it out as h5ad.
The normalized data matrix goes to `X`, raw counts go to `raw/X`, and all
metadata, reductions, and graphs are preserved.

```{r write-h5ad, eval = TRUE}
pbmc_seurat <- readRDS(system.file("extdata", "pbmc_demo.rds", package = "scConvert"))
h5ad_path <- tempfile(fileext = ".h5ad")
writeH5AD(pbmc_seurat, h5ad_path, overwrite = TRUE)
cat("h5ad file size:", round(file.size(h5ad_path) / 1e6, 1), "MB\n")
```

## Verify the round-trip

Read the newly written h5ad back and confirm that expression values, cluster
labels, and UMAP coordinates are preserved.

```{r roundtrip-read, eval = TRUE}
pbmc_rt <- readH5AD(h5ad_path)
```

```{r roundtrip-compare, eval = TRUE, fig.width = 10, fig.height = 5}
library(patchwork)
p1 <- DimPlot(pbmc_seurat, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.5) + NoLegend() + ggtitle("Original (.rds)")
p2 <- DimPlot(pbmc_rt, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.5) + NoLegend() + ggtitle("Round-trip (.h5ad)")
p1 + p2
```

## One-liner format conversion

For quick format conversion without loading data into R, use the `scConvert()`
dispatcher. It detects source and destination formats from file extensions and
picks the most efficient conversion path:

```{r scconvert-oneliner, eval = TRUE}
h5seurat_path <- tempfile(fileext = ".h5seurat")
scConvert(h5ad_path, dest = h5seurat_path, overwrite = TRUE)
cat("Converted h5ad to h5Seurat:", round(file.size(h5seurat_path) / 1e6, 1), "MB\n")
```

## Layer mapping reference

During conversion, scConvert maps data between Seurat and h5ad as follows:

| Seurat | h5ad | Description |
|--------|------|-------------|
| `data` layer | `X` | Normalized expression matrix |
| `counts` layer | `raw/X` | Raw counts |
| `meta.data` | `obs` | Cell metadata (categoricals become factors) |
| Feature metadata | `var` | Gene-level annotations |
| `Reductions(obj)` | `obsm/X_pca`, `obsm/X_umap` | Dimensional reductions |
| `Graphs(obj)` | `obsp/connectivities`, `obsp/distances` | Neighbor graphs |
| `misc` | `uns` | Unstructured annotations |

## Python interop (optional)

The h5ad files produced by `writeH5AD()` are fully compatible with scanpy,
scvi-tools, and CELLxGENE. If you have Python with scanpy installed:

```{python scanpy-read, eval = FALSE}
import scanpy as sc

adata = sc.read_h5ad("pbmc_demo.h5ad")
print(adata)

# Visualize with cluster annotations
sc.pl.umap(adata, color="seurat_annotations")

# Expression patterns are preserved
sc.pl.umap(adata, color="LYZ", use_raw=False)
```

## Clean up

```{r cleanup, eval = TRUE}
unlink(h5ad_path)
unlink(h5seurat_path)
```
