---
title: "SpatialData Zarr Stores"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SpatialData Zarr Stores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

## Introduction

[SpatialData](https://spatialdata.scverse.org/) is the scverse standard for
spatial omics data. A `.spatialdata.zarr` store combines expression tables,
spatial coordinates, shapes, and tissue images in a single Zarr directory.
Technologies like Visium, MERFISH, Xenium, Slide-seq, and CODEX all have
SpatialData representations, making it a unifying format for the spatial
transcriptomics ecosystem.

scConvert provides `readSpatialData()` and `writeSpatialData()` to move
between SpatialData stores and Seurat objects, with no Python dependency.

## Load spatial demo data

We use a 400-spot Visium mouse brain dataset that ships with scConvert. It
includes a tissue image, PCA/UMAP embeddings, and 15 clusters.

```{r load-data, eval = TRUE}
brain <- readRDS(system.file("extdata", "spatial_demo.rds", package = "scConvert"))
brain
```

```{r spatial-plot-original, eval = TRUE, fig.width = 7, fig.height = 5}
SpatialFeaturePlot(brain, features = "Hpca") +
  ggtitle("Hpca expression -- Original Visium data")
```

## Write to SpatialData

Save the Seurat object as a SpatialData Zarr store. The expression data,
metadata, embeddings, spatial coordinates, and shapes are all written.

```{r write-spatialdata, eval = TRUE}
sd_path <- file.path(tempdir(), "brain.spatialdata.zarr")
writeSpatialData(brain, sd_path, overwrite = TRUE, verbose = FALSE)
cat("SpatialData store written to:", sd_path, "\n")
```

## Read back from SpatialData

Load the store back into a Seurat object.

```{r read-spatialdata, eval = TRUE}
brain_rt <- readSpatialData(sd_path, verbose = FALSE)
cat("Cells:", ncol(brain_rt), "| Genes:", nrow(brain_rt), "\n")
cat("Reductions:", paste(Reductions(brain_rt), collapse = ", "), "\n")
cat("Images:", paste(Images(brain_rt), collapse = ", "), "\n")
```

## Compare original and round-trip

Side-by-side comparisons confirm that cluster assignments and gene expression
patterns survive the SpatialData round-trip.

### Cluster UMAP

```{r compare-dimplot, eval = TRUE, fig.width = 12, fig.height = 5}
library(patchwork)

p1 <- DimPlot(brain, reduction = "umap", label = TRUE, pt.size = 1) +
  ggtitle("Original") + NoLegend()

p2 <- DimPlot(brain_rt, reduction = "umap", label = TRUE, pt.size = 1) +
  ggtitle("After SpatialData round-trip") + NoLegend()

p1 + p2
```

### Gene expression

Ttr (transthyretin) marks the choroid plexus in mouse brain. Expression
patterns are identical before and after the round-trip.

```{r compare-feature, eval = TRUE, fig.width = 12, fig.height = 5}
p1 <- FeaturePlot(brain, features = "Ttr", pt.size = 1) +
  ggtitle("Ttr -- Original")
p2 <- FeaturePlot(brain_rt, features = "Ttr", pt.size = 1) +
  ggtitle("Ttr -- SpatialData round-trip")
p1 + p2
```

### Quick fidelity check

```{r fidelity-check, eval = TRUE}
stopifnot(ncol(brain_rt) == ncol(brain))
stopifnot(nrow(brain_rt) == nrow(brain))
stopifnot(identical(sort(colnames(brain_rt)), sort(colnames(brain))))
cat("All checks passed:", ncol(brain_rt), "spots x", nrow(brain_rt), "genes.\n")
```

## Direct pair converters

Convert between SpatialData and other formats in a single call:

```{r pair-converters, eval = FALSE}
# SpatialData <-> h5ad
SpatialDataToH5AD("sample.spatialdata.zarr", "sample.h5ad")
H5ADToSpatialData("sample.h5ad", "sample.spatialdata.zarr")

# SpatialData <-> h5Seurat
SpatialDataToH5Seurat("sample.spatialdata.zarr", "sample.h5seurat")
H5SeuratToSpatialData("sample.h5seurat", "sample.spatialdata.zarr")

# Or use the universal dispatcher
scConvert("sample.spatialdata.zarr", dest = "sample.h5ad", overwrite = TRUE)
```

## Limitations

- **Images**: Large OME-NGFF images (>100M pixels) are skipped to avoid memory
  issues. Use `images = FALSE` to read coordinates only.
- **Coordinate transforms**: SpatialData supports affine transformations
  between elements. These are not currently applied during reading; coordinates
  are used as-is.
- **Labels/segmentation**: Cell and nucleus segmentation masks are recorded by
  name but not loaded (they have no natural Seurat representation).
- **Multiple tables**: Only one table is read per call. Use the `table`
  argument to select which one.

## Python interoperability

SpatialData stores written by scConvert are readable by the Python
[spatialdata](https://spatialdata.scverse.org/) library. Requires Python
with spatialdata installed.

```{python python-read, eval = FALSE}
import spatialdata as sd

sdata = sd.read_zarr("brain.spatialdata.zarr")
print(sdata)
sdata.pl.render_shapes().pl.show()
```

```{r cleanup, include = FALSE}
unlink(sd_path, recursive = TRUE)
```

## Session Info

```{r session-info, eval = TRUE}
sessionInfo()
```
