---
title: "Convert to Zarr Format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convert to Zarr Format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

## Introduction

[Zarr](https://zarr.dev/) is a directory-based format for chunked, compressed
arrays. It is widely used in cloud-native single-cell workflows -- including
[CELLxGENE Census](https://chanzuckerberg.github.io/cellxgene-census/) and
[SpatialData](https://spatialdata.scverse.org/) -- because each chunk is an
independent file that can be read in parallel from object stores like S3 or
GCS. scConvert reads and writes Zarr v2 stores following the AnnData on-disk
specification, with no Python dependency.

## Load demo data

We start with a 500-cell PBMC dataset that ships with scConvert. It has PCA,
UMAP, cluster labels, and neighbor graphs already computed.

```{r load-data, eval = TRUE}
pbmc <- readRDS(system.file("extdata", "pbmc_demo.rds", package = "scConvert"))
pbmc
```

```{r dimplot-original, eval = TRUE, fig.width = 7, fig.height = 5}
DimPlot(pbmc, reduction = "umap", group.by = "seurat_annotations",
        label = TRUE, pt.size = 0.8) +
  ggtitle("Original PBMC data") + NoLegend()
```

## Write to Zarr

Save the Seurat object as a Zarr store with `writeZarr()`. Expression data,
metadata, embeddings, and graphs are all preserved.

```{r write-zarr, eval = TRUE}
zarr_path <- file.path(tempdir(), "pbmc.zarr")
writeZarr(pbmc, zarr_path, overwrite = TRUE, verbose = FALSE)
cat("Zarr store written to:", zarr_path, "\n")
```

## Read back from Zarr

Load the Zarr store back into a Seurat object with `readZarr()`.

```{r read-zarr, eval = TRUE}
pbmc_rt <- readZarr(zarr_path, verbose = FALSE)
cat("Cells:", ncol(pbmc_rt), "| Genes:", nrow(pbmc_rt), "\n")
cat("Reductions:", paste(Reductions(pbmc_rt), collapse = ", "), "\n")
```

## Compare original and round-trip

The UMAP coordinates, cluster labels, and expression values are preserved
exactly through the Zarr round-trip. We verify this with side-by-side
visualizations.

### Cluster labels

```{r compare-dimplot, eval = TRUE, fig.width = 12, fig.height = 5}
library(patchwork)

p1 <- DimPlot(pbmc, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.8) +
  ggtitle("Original") + NoLegend()

p2 <- DimPlot(pbmc_rt, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.8) +
  ggtitle("After Zarr round-trip") + NoLegend()

p1 + p2
```

### Gene expression

Expression patterns are also preserved -- here we compare LYZ (a monocyte
marker) across the original and round-tripped objects.

```{r compare-featureplot, eval = TRUE, fig.width = 12, fig.height = 5}
p1 <- FeaturePlot(pbmc, features = "LYZ", pt.size = 0.8) +
  ggtitle("LYZ -- Original")
p2 <- FeaturePlot(pbmc_rt, features = "LYZ", pt.size = 0.8) +
  ggtitle("LYZ -- Zarr round-trip")
p1 + p2
```

### Quick fidelity check

```{r fidelity-check, eval = TRUE}
# Dimensions match
stopifnot(ncol(pbmc_rt) == ncol(pbmc))
stopifnot(nrow(pbmc_rt) == nrow(pbmc))

# Barcodes and features match
stopifnot(identical(sort(colnames(pbmc_rt)), sort(colnames(pbmc))))
stopifnot(identical(sort(rownames(pbmc_rt)), sort(rownames(pbmc))))

# PCA coordinates are exact
pca_diff <- max(abs(
  Embeddings(pbmc, "pca")[colnames(pbmc_rt), ] -
  Embeddings(pbmc_rt, "pca")
))
stopifnot(pca_diff < 1e-10)
cat("All checks passed: dimensions, barcodes, features, and PCA are exact.\n")
```

## Streaming converters

For converting between file formats without loading data into R, scConvert
provides streaming converters that copy fields directly between backends:

```{r streaming, eval = FALSE}
# h5ad <-> Zarr (no Seurat intermediate)
H5ADToZarr("data.h5ad", "data.zarr")
ZarrToH5AD("data.zarr", "data.h5ad")

# h5Seurat <-> Zarr
H5SeuratToZarr("data.h5seurat", "data.zarr")
ZarrToH5Seurat("data.zarr", "data.h5seurat")

# Or use the universal dispatcher
scConvert("data.h5ad", dest = "data.zarr", overwrite = TRUE)
```

These are particularly useful for large datasets where materializing the full
Seurat object would be expensive.

## Python interoperability

Zarr stores produced by `writeZarr()` are directly readable by Python's
`anndata.read_zarr()` and scanpy. Requires Python with anndata installed.

```{python python-read, eval = FALSE}
import anndata as ad
import scanpy as sc

adata = ad.read_zarr("pbmc.zarr")
print(adata)
sc.pl.umap(adata, color="seurat_annotations")
```

```{r cleanup, include = FALSE}
unlink(zarr_path, recursive = TRUE)
```

## Session Info

```{r session-info, eval = TRUE}
sessionInfo()
```
