---
title: "Convert Between Seurat and SingleCellExperiment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convert Between Seurat and SingleCellExperiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
has_sce <- requireNamespace("SingleCellExperiment", quietly = TRUE) &&
  requireNamespace("SummarizedExperiment", quietly = TRUE)
library(scConvert)
library(Seurat)
```

Many Bioconductor workflows -- scran normalization, scater quality control,
slingshot trajectory inference -- operate on the SingleCellExperiment (SCE)
container. Seurat provides built-in converters (`as.SingleCellExperiment()` and
`as.Seurat()`) that make it straightforward to move data between the two
ecosystems. scConvert extends this by letting you route SCE objects to any file
format (h5ad, h5Seurat, loom, zarr, rds) through its hub architecture.

## Load demo data

scConvert ships a small PBMC dataset. We start by reading it into a Seurat
object.

```{r load-data}
pbmc <- readH5AD(system.file("extdata", "pbmc_demo.h5ad", package = "scConvert"))
pbmc
```

## Seurat to SingleCellExperiment

Convert the Seurat object to SCE. Counts, normalized data, dimensional
reductions, and cell metadata are all transferred.

```{r seurat-to-sce, eval = has_sce}
library(SingleCellExperiment)

sce <- as.SingleCellExperiment(pbmc)
sce
```

Inspect what was carried over:

```{r inspect-sce, eval = has_sce}
cat("Assays:", paste(assayNames(sce), collapse = ", "), "\n")
cat("Reduced dims:", paste(reducedDimNames(sce), collapse = ", "), "\n")
cat("colData columns:", paste(colnames(colData(sce)), collapse = ", "), "\n")
```

## SingleCellExperiment to Seurat

Convert the SCE back to a Seurat object with `as.Seurat()`. The converter
detects which assays exist (counts, logcounts) and maps them to the appropriate
Seurat layers.

```{r sce-to-seurat, eval = has_sce}
seurat_rt <- as.Seurat(sce, counts = "counts", data = "logcounts")
seurat_rt
```

## Verify the round-trip

Confirm that expression values, cell barcodes, and embeddings survive the
Seurat -> SCE -> Seurat round-trip.

```{r roundtrip-check, eval = has_sce}
# Cell barcodes
cat("Barcodes match:", identical(Cells(pbmc), Cells(seurat_rt)), "\n")

# Gene names
cat("Features match:", identical(rownames(pbmc), rownames(seurat_rt)), "\n")

# Count matrix values
orig_counts <- GetAssayData(pbmc, layer = "counts")
rt_counts <- GetAssayData(seurat_rt, layer = "counts")
cat("Counts identical:", identical(orig_counts, rt_counts), "\n")

# Dimensionality reductions (SCE stores these in reducedDims)
cat("Reductions in original:", paste(Reductions(pbmc), collapse = ", "), "\n")
cat("Reductions after round-trip:", paste(Reductions(seurat_rt), collapse = ", "), "\n")
```

## SCE to h5ad via scConvert

To share an SCE dataset with Python collaborators, pass it directly to
`scConvert()`. The SCE is converted to Seurat internally, then written to the
target format -- no manual intermediate step needed.

```{r sce-to-h5ad, eval = has_sce}
h5ad_path <- tempfile(fileext = ".h5ad")
scConvert(sce, dest = h5ad_path, overwrite = TRUE)
cat("h5ad file size:", round(file.size(h5ad_path) / 1e6, 1), "MB\n")
```

Read it back to confirm the data survived the full SCE -> Seurat -> h5ad
-> Seurat chain:

```{r h5ad-roundtrip, eval = has_sce}
pbmc_h5ad <- readH5AD(h5ad_path)
cat("Cells:", ncol(pbmc_h5ad), "\n")
cat("Genes:", nrow(pbmc_h5ad), "\n")
cat("Reductions:", paste(Reductions(pbmc_h5ad), collapse = ", "), "\n")
```

## Seurat to SCE via scConvert

The reverse also works. Use `dest = "sce"` to get a SingleCellExperiment
directly from any Seurat object:

```{r seurat-to-sce-scconvert, eval = has_sce}
sce2 <- scConvert(pbmc, dest = "sce")
cat("Class:", class(sce2), "\n")
cat("Dimensions:", nrow(sce2), "x", ncol(sce2), "\n")
```

## Practical example: Bioconductor QC on SCE, then back to Seurat

A common workflow is to use Bioconductor tools for quality control or
normalization, then bring the result back into Seurat for downstream analysis.
Here we compute log-library-size and detected-feature counts using base R
(the same quantities that `scater::addPerCellQCMetrics` would compute).

```{r practical-example, eval = has_sce}
# Compute QC metrics on the SCE
sce_qc <- as.SingleCellExperiment(pbmc)
counts_mat <- assay(sce_qc, "counts")
colData(sce_qc)$log_library_size <- log1p(Matrix::colSums(counts_mat))
colData(sce_qc)$detected_features <- Matrix::colSums(counts_mat > 0)

cat("Added QC columns:",
    paste(c("log_library_size", "detected_features"), collapse = ", "), "\n")

# Convert back to Seurat -- the new colData columns become metadata
seurat_qc <- as.Seurat(sce_qc, counts = "counts", data = "logcounts")
cat("Seurat metadata columns:",
    paste(colnames(seurat_qc[[]]), collapse = ", "), "\n")
```

The new QC columns (`log_library_size`, `detected_features`) are now available
as Seurat metadata for filtering or visualization.

## Clean up

```{r cleanup, eval = has_sce}
unlink(h5ad_path)
```
