---
title: "Converting Between Seurat and TileDB-SOMA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Converting Between Seurat and TileDB-SOMA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  out.width = "100%",
  purl = FALSE
)
```

## What is TileDB-SOMA?

[TileDB-SOMA](https://github.com/single-cell-data/TileDB-SOMA) is a
cloud-native, array-based format for storing and querying single-cell data at
scale. It is the storage backend behind
[CELLxGENE Census](https://chanzuckerberg.github.io/cellxgene-census/), the
largest public atlas of single-cell data (61M+ cells across 900+ datasets).

Key properties of SOMA:

- **Columnar storage** backed by TileDB arrays -- efficient slicing by cells
  or features without loading the full dataset
- **Cloud-native**: local paths, S3, GCS, and TileDB Cloud URIs
- **Query filtering**: select cells by metadata predicates at the storage level
- **Multi-measurement**: RNA, ATAC, and protein assays in one experiment

scConvert provides `readSOMA()` and `writeSOMA()` for Seurat interoperability,
plus pair converters for all supported formats. All SOMA functions require the
[tiledbsoma](https://github.com/single-cell-data/TileDB-SOMA) R package.

```{r install_deps}
# Install tiledbsoma (required for SOMA support)
install.packages("tiledbsoma")

library(scConvert)
library(Seurat)
```

## Reading from SOMA

Use `readSOMA()` to load a SOMA experiment as a Seurat object:

```{r read_basic}
# Read a local SOMA experiment
obj <- readSOMA("path/to/experiment")

# Read a specific measurement (default is "RNA")
obj <- readSOMA("path/to/experiment", measurement = "ATAC")
```

### Filtering cells and features

`readSOMA()` supports server-side filtering via value filter queries, which is
particularly useful for large atlases:

```{r read_filtered}
# Filter by cell type and tissue
obj <- readSOMA(
  "path/to/experiment",
  obs_query = "cell_type == 'T cell' & tissue == 'blood' & disease == 'normal'",
  obs_column_names = c("cell_type", "tissue", "donor_id")
)
```

### What gets read

`readSOMA()` maps the following SOMA components to Seurat:

| SOMA Source | Seurat Destination | Description |
|-------------|-------------------|-------------|
| `obs` | `meta.data` | Cell metadata (factors, numerics, strings) |
| `ms[measurement]/var` | Feature metadata | Gene-level annotations |
| `ms[measurement]/X/data` | Default assay counts | Expression matrix (sparse) |
| `ms[measurement]/obsm/*` | `reductions` | PCA, UMAP, etc. |
| `ms[measurement]/obsp/*` | `graphs` | Neighbor graphs (SNN, distances) |
| Other measurements | Additional assays | Multi-modal data (ATAC, protein) |

The `soma_joinid` column (internal to TileDB-SOMA) is automatically removed
from cell and feature metadata.

## Writing to SOMA

Use `writeSOMA()` to save a Seurat object as a SOMA experiment:

```{r write_basic}
# Write to a local SOMA experiment
writeSOMA(seurat_obj, uri = "output.soma")

# Overwrite an existing experiment
writeSOMA(seurat_obj, uri = "output.soma", overwrite = TRUE)
```

`writeSOMA()` delegates to `tiledbsoma::write_soma()`, which converts all
assays, metadata, embeddings, and graphs. Each Seurat assay becomes a separate
SOMA measurement.

## Direct pair converters

scConvert provides direct conversion functions between SOMA and all other
supported formats. All route through a Seurat intermediate.

### SOMA and h5ad

```{r pair_h5ad}
# h5ad -> SOMA
H5ADToSOMA("data.h5ad", "data.soma")

# SOMA -> h5ad
SOMAToH5AD("data.soma", "output.h5ad")
```

### SOMA and h5Seurat

```{r pair_h5seurat}
# h5Seurat -> SOMA
H5SeuratToSOMA("data.h5seurat", "data.soma")

# SOMA -> h5Seurat
SOMAToH5Seurat("data.soma", "output.h5seurat")
```

### Other format pairs

Converters also exist for Loom, Zarr, and h5mu -- each follows the same
`{Format}ToSOMA()` / `SOMATo{Format}()` pattern:

```{r pair_others}
LoomToSOMA("data.loom", "data.soma")    # Loom -> SOMA
ZarrToSOMA("data.zarr", "data.soma")    # Zarr -> SOMA
H5MUToSOMA("data.h5mu", "data.soma")    # h5mu -> SOMA
SOMAToLoom("data.soma", "output.loom")   # SOMA -> Loom
SOMAToZarr("data.soma", "output.zarr")   # SOMA -> Zarr
SOMAToH5MU("data.soma", "output.h5mu")  # SOMA -> h5mu
```

## CLI and dispatcher usage

Both `scConvert_cli()` and the universal `scConvert()` dispatcher auto-detect
SOMA from the `.soma` extension:

```{r cli}
# scConvert_cli auto-dispatches
scConvert_cli("data.h5ad", "data.soma")
scConvert_cli("data.soma", "output.h5ad")

# Universal dispatcher
scConvert("data.h5ad", dest = "data.soma", overwrite = TRUE)
scConvert("data.soma", dest = "output.h5seurat", overwrite = TRUE)
```

## CELLxGENE Census access

[CELLxGENE Census](https://chanzuckerberg.github.io/cellxgene-census/) hosts
61M+ cells as a single SOMA experiment. Query it with `readSOMA()`, then
convert the result to any format:

```{r census}
library(cellxgene.census)

census <- open_soma(census_version = "stable")
human_uri <- census$get("census_data")$get("homo_sapiens")$uri

# Read a subset directly as a Seurat object
tcells <- readSOMA(
  uri = human_uri,
  measurement = "RNA",
  obs_query = "cell_type == 'T cell' & tissue_general == 'blood'",
  obs_column_names = c("cell_type", "tissue", "donor_id")
)

# Save locally in any format
writeH5AD(tcells, "census_tcells.h5ad")
writeSOMA(tcells, "census_tcells.soma")
```

See the [cellxgene-census documentation](https://chanzuckerberg.github.io/cellxgene-census/)
for more details on the Census API.

## Data mapping reference

| Seurat | SOMA | Direction | Notes |
|--------|------|-----------|-------|
| Default assay counts | `ms[RNA]/X/data` | Read/Write | Sparse matrix |
| `meta.data` | `obs` | Read/Write | Factors, numerics, strings |
| Feature metadata | `ms[RNA]/var` | Read/Write | Gene annotations |
| `Embeddings(, "pca")` | `ms[RNA]/obsm/X_pca` | Read/Write | Dense array |
| `Embeddings(, "umap")` | `ms[RNA]/obsm/X_umap` | Read/Write | Dense array |
| `Graphs(, "RNA_snn")` | `ms[RNA]/obsp/RNA_snn` | Read/Write | Sparse matrix |
| Additional assays | Other measurements | Read/Write | Multi-modal support |

## Limitations

- **Requires tiledbsoma**: install with `install.packages("tiledbsoma")`.
  `writeSOMA()` further requires `tiledbsoma::write_soma()` to be available.
- **No streaming conversion**: unlike h5ad/zarr converters, SOMA always routes
  through a Seurat intermediate (TileDB arrays differ fundamentally from HDF5).
- **No C binary path**: the compiled C CLI handles HDF5-to-HDF5 only.
- **scale.data** is not written to SOMA -- re-run `ScaleData()` after loading.
- **Active identity** (`Idents()`) is not stored -- set manually after reading.

## See also

- `vignette("convert-anndata")` -- Converting between Seurat and h5ad format
- `vignette("cli-usage")` -- Command-line interface for batch conversions
- `vignette("convert-zarr")` -- Converting between Seurat and Zarr format

## Session Info

```{r session, eval = TRUE, purl = FALSE}
sessionInfo()
```
