---
title: "Convert to Zarr Format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convert to Zarr Format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

## Introduction

[Zarr](https://zarr.dev/) is a directory-based format for chunked, compressed
arrays. It is widely used in cloud-native single-cell workflows -- including
[CELLxGENE Census](https://chanzuckerberg.github.io/cellxgene-census/) and
[SpatialData](https://spatialdata.scverse.org/) -- because each chunk is an
independent file that can be read in parallel from object stores like S3 or
GCS. scConvert reads and writes Zarr v2 stores following the AnnData on-disk
specification, with no Python dependency.

## Read a Zarr store

We start with the shipped PBMC h5ad demo and convert it to Zarr so we can
demonstrate the Zarr read/write workflow. This dataset has 230 cells, 220
genes, PCA/UMAP embeddings, neighbor graphs, and 9 annotated cell types.

```{r create-zarr, eval = TRUE}
h5ad_path <- system.file("testdata", "pbmc_small.h5ad", package = "scConvert")
zarr_path <- file.path(tempdir(), "pbmc_demo.zarr")
scConvert(h5ad_path, zarr_path, overwrite = TRUE, verbose = FALSE)
```

```{r read-zarr, eval = TRUE}
pbmc <- readZarr(zarr_path, verbose = FALSE)
pbmc
```

```{r dimplot-zarr, eval = TRUE, fig.width = 7, fig.height = 5}
DimPlot(pbmc, reduction = "umap", group.by = "seurat_annotations",
        label = TRUE, pt.size = 0.8) +
  ggtitle("PBMC data read from Zarr") + NoLegend()
```

LYZ is a strong monocyte marker. We can verify that expression values were
loaded correctly from the Zarr store.

```{r featureplot-lyz, eval = TRUE, fig.width = 7, fig.height = 5}
FeaturePlot(pbmc, features = "LYZ", pt.size = 0.8) +
  ggtitle("LYZ expression (from Zarr)")
```

## Write to Zarr and read back

Now we write the Seurat object to a new Zarr store and read it back to verify
round-trip fidelity.

```{r write-roundtrip, eval = TRUE}
rt_path <- file.path(tempdir(), "pbmc_roundtrip.zarr")
writeZarr(pbmc, rt_path, overwrite = TRUE, verbose = FALSE)
pbmc_rt <- readZarr(rt_path, verbose = FALSE)
cat("Cells:", ncol(pbmc_rt), "| Genes:", nrow(pbmc_rt), "\n")
cat("Reductions:", paste(names(pbmc_rt@reductions), collapse = ", "), "\n")
```

### Verify the round-trip

```{r fidelity-check, eval = TRUE}
cat("Barcodes match:", identical(sort(Cells(pbmc)), sort(Cells(pbmc_rt))), "\n")
cat("Features match:", identical(sort(rownames(pbmc)), sort(rownames(pbmc_rt))), "\n")
cat("Dimensions match:", identical(dim(pbmc), dim(pbmc_rt)), "\n")
```

## Streaming converters

For converting between file formats without loading data into R, scConvert
provides streaming converters that copy fields directly between backends:

```{r streaming, eval = FALSE}
# h5ad <-> Zarr (no Seurat intermediate)
H5ADToZarr("data.h5ad", "data.zarr")
ZarrToH5AD("data.zarr", "data.h5ad")

# h5Seurat <-> Zarr
H5SeuratToZarr("data.h5seurat", "data.zarr")
ZarrToH5Seurat("data.zarr", "data.h5seurat")

# Or use the universal dispatcher
scConvert("data.h5ad", dest = "data.zarr", overwrite = TRUE)
```

These are particularly useful for large datasets where materializing the full
Seurat object would be expensive.

## Python interoperability

Zarr stores produced by `writeZarr()` are directly readable by Python's
`anndata.read_zarr()` and scanpy. Requires Python with anndata installed.

```{python python-read, eval = FALSE}
import anndata as ad
import scanpy as sc

adata = ad.read_zarr("pbmc.zarr")
print(adata)
sc.pl.umap(adata, color="seurat_annotations")
```

```{r cleanup, include = FALSE}
unlink(zarr_path, recursive = TRUE)
unlink(rt_path, recursive = TRUE)
```

## Session Info

```{r session-info, eval = TRUE}
sessionInfo()
```
