---
title: 'Conversions: h5Seurat and AnnData'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conversions: h5Seurat and AnnData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  out.width = "100%",
  dev = "png",
  purl = FALSE
)
has_SeuratData <- requireNamespace("SeuratData", quietly = TRUE) &&
  tryCatch({
    library(SeuratData)
    "pbmc3k.SeuratData" %in% rownames(InstalledData())
  }, error = function(e) FALSE)
# Python ecosystem detection via reticulate
has_scanpy <- FALSE
has_squidpy <- FALSE
if (requireNamespace("reticulate", quietly = TRUE)) {
  scverse_path <- "/opt/homebrew/Caskroom/miniconda/base/envs/scverse"
  if (dir.exists(scverse_path)) {
    reticulate::use_condaenv("scverse", required = FALSE)
    Sys.setenv(
      OMP_NUM_THREADS = "1", OPENBLAS_NUM_THREADS = "1",
      MKL_NUM_THREADS = "1", NUMBA_THREADING_LAYER = "tbb"
    )
    has_scanpy <- reticulate::py_module_available("scanpy")
    has_squidpy <- reticulate::py_module_available("squidpy")
    if (has_scanpy) {
      reticulate::py_run_string("import matplotlib; matplotlib.use('Agg')")
    }
  }
}
```

```{r cleanup, echo=FALSE, results='hide', purl = FALSE}
outs <- c(
  "pbmc3k.h5Seurat", "pbmc3k.h5ad",
  "stxBrain.h5Seurat", "stxBrain.h5ad", "stxBrain_direct.h5ad",
  "crc_sample.h5ad", "crc_normalized.h5ad", "crc_normalized.h5seurat",
  "visium_colon.h5ad", "visium_normalized.h5ad", "visium_normalized.h5seurat",
  "cbmc_rna.h5seurat", "cbmc_rna.h5ad", "cbmc_adt.h5seurat", "cbmc_adt.h5ad"
)
existing_files <- outs[file.exists(outs)]
if (length(existing_files) > 0) file.remove(existing_files)
```

This vignette showcases how to convert between `Seurat` objects and AnnData files via h5Seurat files. This allows interoperability between Seurat and [Scanpy](https://scanpy.readthedocs.io/).

```{r packages, eval = has_SeuratData, purl = FALSE}
library(Seurat)
library(scConvert)
library(ggplot2)
library(patchwork)
```

## Converting from Seurat to AnnData via h5Seurat

To demonstrate conversion from a `Seurat` object to an AnnData file, we'll use
the `pbmc3k.final` dataset from SeuratData - a processed PBMC dataset with
clustering and UMAP. If SeuratData is not available, you can download an h5ad
directly from [CZ CellxGene Discover](https://cellxgene.cziscience.com/collections):

```{r get_data_cxg, eval = FALSE}
# Alternative: download from CellxGene (Tabula Microcebus lung, ~1.7K cells)
cxg_url <- "https://datasets.cellxgene.cziscience.com/582ffa61-1c54-4492-8763-2ecfa9efa070.h5ad"
download.file(cxg_url, destfile = "cxg_demo.h5ad", mode = "wb")
pbmc <- readH5AD("cxg_demo.h5ad")
```

```{r get_data, message=FALSE, warning=FALSE, eval = has_SeuratData, purl = FALSE}
library(SeuratData)
if (!"pbmc3k.final" %in% rownames(InstalledData())) {
  InstallData("pbmc3k")
}

data("pbmc3k.final", package = "pbmc3k.SeuratData")
pbmc <- UpdateSeuratObject(pbmc3k.final)
pbmc
```

This is a fully processed Seurat object with clustering and dimensional reductions:

```{r plot_pbmc, eval = has_SeuratData, purl = FALSE}
DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```

```{r plot_pbmc_cd14, eval = has_SeuratData, purl = FALSE}
FeaturePlot(pbmc, features = "CD14", pt.size = 0.5)
```

Converting the `Seurat` object to an AnnData file is a two-step process:

1. Save the `Seurat` object as an h5Seurat file using `writeH5Seurat()`
2. Convert to AnnData using `scConvert()`

```{r convert_seurat, eval = has_SeuratData, purl = FALSE}
cat("Seurat layers:", paste(Layers(pbmc), collapse = ", "), "\n")
writeH5Seurat(pbmc, filename = "pbmc3k.h5Seurat", overwrite = TRUE)
scConvert("pbmc3k.h5Seurat", dest = "h5ad", overwrite = TRUE)
```

We can view the AnnData file in Scanpy:

```{python load_adata, eval = has_scanpy && has_SeuratData}
import scanpy as sc
adata = sc.read_h5ad("pbmc3k.h5ad")
print(adata)
```

And visualize with cluster annotations:

```{python plot_adata, eval = has_scanpy && has_SeuratData}
sc.pl.umap(adata, color='seurat_annotations', legend_loc='on data', legend_fontsize=8)
```

```{python plot_adata_cd14, eval = has_scanpy && has_SeuratData}
sc.pl.umap(adata, color='CD14', use_raw=False)
```

The conversion preserves expression patterns - CD14 shows consistent distribution in both tools.

## Direct Loading with readH5AD

In addition to the two-step `scConvert()` + `readH5Seurat()` workflow shown above, scConvert provides `readH5AD()` for loading h5ad files directly into Seurat objects without creating an intermediate h5Seurat file.

```{r direct_load, eval = has_SeuratData, purl = FALSE}
# One-step: load h5ad directly into a Seurat object
pbmc_direct <- readH5AD("pbmc3k.h5ad", verbose = TRUE)

# Compare with the two-step approach
cat("Direct cells:", ncol(pbmc_direct), "| Two-step cells:", ncol(pbmc), "\n")
cat("Direct reductions:", paste(names(pbmc_direct@reductions), collapse = ", "), "\n")
```

`readH5AD()` reads the following from h5ad files:

| h5ad Location | Seurat Destination | Description |
|--------------|-------------------|-------------|
| `X` | Default assay `data` layer | Expression matrix (sparse or dense) |
| `raw/X` | `counts` layer | Raw counts if present |
| `layers/*` | Additional layers | Named layers mapped to Seurat slots |
| `obs` | `meta.data` | Cell metadata (categorical preserved as factors) |
| `var` | Feature metadata | Gene-level annotations |
| `var['highly_variable']` | `VariableFeatures()` | Variable feature selection |
| `obsm/X_umap` | `reductions$umap` | UMAP coordinates |
| `obsm/X_pca` | `reductions$pca` | PCA embeddings |
| `obsm/X_tsne` | `reductions$tsne` | tSNE coordinates |
| `obsm/spatial` | Spatial coordinates | Via `H5ADSpatialToSeurat()` |
| `obsp/connectivities` | `graphs$RNA_snn` | SNN graph |
| `obsp/distances` | `graphs$RNA_nn` | Distance graph |
| `uns/*` | `misc` | Unstructured annotations |

**When to use each approach:**

| Scenario | Recommended |
|----------|------------|
| Quick exploration of an h5ad file | `readH5AD()` |
| Round-trip editing (load, modify, re-export) | `scConvert()` + `readH5Seurat()` |
| Need h5Seurat for other tools | `scConvert()` |
| Loading scanpy-processed data for Seurat analysis | `readH5AD()` |
| Working with spatial h5ad from CellxGene | `readH5AD()` |

## Converting from AnnData to Seurat via h5Seurat

To demonstrate conversion from AnnData to Seurat, we'll use a colorectal cancer sample from [CellxGene](https://cellxgene.cziscience.com).

```{r get_h5ad, eval = has_SeuratData, purl = FALSE}
h5ad_path <- system.file("testdata", "crc_sample.h5ad", package = "scConvert")
if (file.exists(h5ad_path)) {
  file.copy(h5ad_path, "crc_sample.h5ad", overwrite = TRUE)
} else {
  download.file(
    "https://datasets.cellxgene.cziscience.com/91cf9a95-0b9a-4ece-b8eb-7b9e3409a0d3.h5ad",
    "crc_sample.h5ad", mode = "wb"
  )
}
```

View the h5ad file in Scanpy:

```{python view_crc, eval = has_scanpy}
import scanpy as sc
adata_crc = sc.read_h5ad("crc_sample.h5ad")
print(adata_crc)
```

CellxGene datasets use Ensembl IDs as `var_names` by default. The `feature_name` column contains gene symbols. During conversion, scConvert automatically uses gene symbols when `feature_name` is available.

> **Note on data layers**: During conversion, AnnData's `X` matrix (typically log-normalized data indicated by `log1p_total_counts` in obs) is stored in Seurat's `data` layer. If `raw/X` exists, it becomes the `counts` layer.

Visualize with scanpy before conversion:

```{python plot_crc_tissue, eval = has_scanpy && has_SeuratData}
sc.pl.umap(adata_crc, color='tissue')
```

```{python normalize_crc, eval = has_scanpy && has_SeuratData}
import pandas as pd

sc.pp.normalize_total(adata_crc, target_sum=1e4)
sc.pp.log1p(adata_crc)

# Set gene symbols as var_names (force string conversion from Categorical)
adata_crc.var_names = pd.Index(adata_crc.var['feature_name'].astype(str).values)
adata_crc.var_names_make_unique()

adata_crc.write_h5ad("crc_normalized.h5ad")

sc.pl.umap(adata_crc, color='EPCAM', use_raw=False, title='EPCAM')
```

Convert to Seurat:

```{r convert_crc, eval = has_scanpy, purl = FALSE}
scConvert("crc_normalized.h5ad", dest = "h5seurat", overwrite = TRUE)
crc <- readH5Seurat("crc_normalized.h5seurat")

# Verify layer mapping: X -> data (log-normalized), raw/X -> counts (if exists)
cat("Layers:", paste(Layers(crc), collapse = ", "), "\n")
cat("Data layer range:", round(range(GetAssayData(crc, layer = "data")[1:100, 1:10]), 2), "\n")
crc
```

### Visualize Converted Data

The UMAP coordinates and normalized expression from scanpy are preserved:

```{r plot_crc, eval = has_scanpy, purl = FALSE}
DimPlot(crc, reduction = "umap", group.by = "tissue", pt.size = 0.5)
```

```{r plot_crc_epcam, eval = has_scanpy, purl = FALSE}
FeaturePlot(crc, features = "EPCAM", pt.size = 0.5)
```

The conversion preserves expression patterns - EPCAM shows consistent distribution in both tools.

## Visium Spatial Data Conversion

For spatial transcriptomics data, we use the stxBrain dataset from SeuratData (Visium v2 format):

```{r get_stxBrain, eval = has_SeuratData, purl = FALSE}
library(SeuratData)
if (!"stxBrain" %in% rownames(InstalledData())) {
  InstallData("stxBrain")
}

brain <- UpdateSeuratObject(LoadData("stxBrain", type = "anterior1"))
brain <- NormalizeData(brain)
cat("Layers:", paste(Layers(brain), collapse = ", "), "\n")

SpatialFeaturePlot(brain, features = "Hpca")
```

Convert to h5ad using the direct pipeline:

```{r convert_stxBrain, eval = has_SeuratData, purl = FALSE}
writeH5AD(brain, "stxBrain.h5ad", overwrite = TRUE)
```

View in Python with Squidpy:

```{python load_stxBrain, eval = has_SeuratData && has_squidpy, error = TRUE}
import squidpy as sq
import scanpy as sc

adata_spatial = sc.read_h5ad("stxBrain.h5ad")
print(adata_spatial)

sq.pl.spatial_scatter(adata_spatial, color="Hpca", library_id="anterior1",
                      img_res_key="lowres", size=1, alpha=0.5, use_raw=False)
```

## Multi-assay Conversion (CITE-seq)

For multi-modal data like CITE-seq, each assay must be converted separately since h5ad format only supports a single matrix per file.

**Conversion behavior:**

- **No assay specified**: Only the default assay is converted
- **Single assay specified**: That specific assay is converted
- **Multiple assays**: Must call the function multiple times, once per assay

This example uses the `cbmc` dataset from SeuratData:

```{r cite_seq_conversion, eval = has_SeuratData, purl = FALSE}
library(SeuratData)
if (!"cbmc" %in% rownames(InstalledData())) {
  InstallData("cbmc")
}
data("cbmc", package = "cbmc.SeuratData")
cbmc <- UpdateSeuratObject(cbmc)

cat("Assays:", paste(Assays(cbmc), collapse = ", "), "\n")

# Process RNA for visualization
DefaultAssay(cbmc) <- "RNA"
cbmc <- NormalizeData(cbmc, verbose = FALSE)
cbmc <- FindVariableFeatures(cbmc, verbose = FALSE)
cbmc <- ScaleData(cbmc, verbose = FALSE)
cbmc <- RunPCA(cbmc, verbose = FALSE)
cbmc <- RunUMAP(cbmc, dims = 1:30, verbose = FALSE)

# Normalize ADT with CLR
cbmc <- NormalizeData(cbmc, assay = "ADT", normalization.method = "CLR", verbose = FALSE)
```

Visualize RNA and ADT markers before conversion:

```{r cite_seq_featureplot, eval = has_SeuratData, purl = FALSE, fig.width = 14, fig.height = 6}
library(patchwork)
p1 <- FeaturePlot(cbmc, features = "CD3D", pt.size = 0.3) + ggtitle("CD3D (RNA)")
DefaultAssay(cbmc) <- "ADT"
p2 <- FeaturePlot(cbmc, features = "adt_CD3", pt.size = 0.3) + ggtitle("CD3 protein (ADT)")
p1 + p2
```

Convert RNA assay to h5ad:

```{r cite_seq_convert, eval = has_SeuratData, purl = FALSE}
DefaultAssay(cbmc) <- "RNA"
writeH5Seurat(cbmc, "cbmc_rna.h5seurat", overwrite = TRUE)
scConvert("cbmc_rna.h5seurat", dest = "h5ad", overwrite = TRUE)
```

Verify and visualize the converted file in scanpy:

```{python verify_cbmc_h5ad, eval = has_scanpy && has_SeuratData}
import scanpy as sc
adata_rna = sc.read_h5ad("cbmc_rna.h5ad")
print(adata_rna)

sc.pl.umap(adata_rna, color='CD3D', use_raw=False, title='CD3D (RNA, from h5ad)')
```

The h5ad file preserves the full structure: expression matrix (X), cell metadata (obs), gene info (var), and UMAP coordinates.

Convert the ADT assay separately:

```{r cite_seq_adt, eval = has_SeuratData, purl = FALSE}
# Convert ADT assay
DefaultAssay(cbmc) <- "ADT"
writeH5Seurat(cbmc, "cbmc_adt.h5seurat", overwrite = TRUE)
scConvert("cbmc_adt.h5seurat", dest = "h5ad", overwrite = TRUE)
```

```{python verify_adt_h5ad, eval = has_scanpy && has_SeuratData}
import scanpy as sc
adata_adt = sc.read_h5ad("cbmc_adt.h5ad")
print(adata_adt)
print("ADT features:", list(adata_adt.var_names)[:10])
```

Each assay produces a separate h5ad file with the same cell metadata but different features.

> **Note on multi-modal formats**: For true multi-modal interoperability, consider the [MuData/h5mu format](https://muon.scverse.org/) from the scverse ecosystem. scConvert provides native `writeH5MU()` and `readH5MU()` functions for reading and writing h5mu files with no external dependencies. See `vignette("multimodal-h5mu")` for details.

## Spatial h5ad to Seurat

Converting native spatial h5ad files to Seurat is fully supported. We use a Visium colon sample from [CellxGene](https://cellxgene.cziscience.com) that was processed with scanpy/squidpy standard workflows:

```{r get_spatial_h5ad, eval = has_scanpy, purl = FALSE}
cache_dir <- tools::R_user_dir("scConvert", which = "cache")
cache_path <- file.path(cache_dir, "visium_colon.h5ad")

if (!file.exists(cache_path)) {
  dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
  message("Downloading Visium colon dataset (~1.7GB)...")
  download.file(
    "https://datasets.cellxgene.cziscience.com/ab9f4860-c0e3-444b-a982-38c13f0be6f5.h5ad",
    cache_path, mode = "wb"
  )
}
file.copy(cache_path, "visium_colon.h5ad", overwrite = TRUE)
```

View the native spatial h5ad in Python:

```{python view_spatial_h5ad, eval = has_scanpy}
import scanpy as sc
adata_spatial = sc.read_h5ad("visium_colon.h5ad")
print(adata_spatial)

lib_id = list(adata_spatial.uns['spatial'].keys())[0]
spatial_data = adata_spatial.uns['spatial'][lib_id]
print("\nSpatial library:", lib_id)
print("Image keys:", list(spatial_data.get('images', {}).keys()))
print("in_tissue column:", 'in_tissue' in adata_spatial.obs.columns)
```

Normalize, set gene symbols, and save for conversion:

```{python squidpy_spatial_ACTC1, eval = has_squidpy}
import squidpy as sq
import scanpy as sc
import pandas as pd

adata_spatial = sc.read_h5ad("visium_colon.h5ad")

# Filter to tissue spots only
adata_spatial = adata_spatial[adata_spatial.obs['in_tissue'] == 1].copy()

sc.pp.normalize_total(adata_spatial, target_sum=1e4)
sc.pp.log1p(adata_spatial)

adata_spatial.var_names = pd.Index(adata_spatial.var['feature_name'].astype(str).values)
adata_spatial.var_names_make_unique()

adata_spatial.write_h5ad("visium_normalized.h5ad")

lib_id = list(adata_spatial.uns['spatial'].keys())[0]
sq.pl.spatial_scatter(adata_spatial, color='ACTC1', library_id=lib_id,
                      size=1, alpha=0.5, use_raw=False, title='ACTC1')
```

Convert to Seurat:

```{r spatial_h5ad_to_seurat, eval = has_scanpy, purl = FALSE}
scConvert("visium_normalized.h5ad", dest = "h5seurat", overwrite = TRUE)
visium <- readH5Seurat("visium_normalized.h5seurat")

# Verify layer mapping: X -> data (log-normalized)
cat("Layers:", paste(Layers(visium), collapse = ", "), "\n")
visium
```

Verify spatial data was preserved:

```{r plot_spatial_native, eval = has_scanpy, fig.width=10, fig.height=8, purl = FALSE}
SpatialFeaturePlot(visium, features = "ACTC1", pt.size.factor = 3, alpha = 1)
```

> **Note**: Spatial images and coordinates are preserved during conversion. Some scanpy-specific structures (like neighbor graphs in `obsp`) may need to be recomputed in Seurat using `FindNeighbors()`.

## Data Mapping Reference

This section provides comprehensive mapping tables showing how data is converted between Seurat and AnnData formats.

### Core Data Slots

**Layer mapping during Seurat -> h5ad conversion (via h5Seurat):**

| Seurat Slot | h5ad Destination | Notes |
|-------------|------------------|-------|
| `data` (normalized) | `X` | All genes; used as primary matrix |
| `counts` (raw) | `raw/X` | All genes; if available |
| `scale.data` | *(skipped)* | Only ~2,000 variable features; recompute with `sc.pp.scale()` |
| `VariableFeatures()` | `var['highly_variable']` | Boolean column in `var` |

> **Note**: scConvert prioritizes `data` (all genes) over `scale.data` (variable features only) when writing `X`. This ensures `var` contains the full gene set. Variable features are preserved as a boolean `highly_variable` column in `var`. `scale.data` is only used as a fallback if neither `data` nor `counts` are available.

**Layer mapping during h5ad -> Seurat conversion:**

| Seurat Slot | h5ad Source | Condition |
|-------------|-------------|-----------|
| `data` | `X` | Always mapped |
| `counts` | `raw/X` | If `raw` group exists |
| `counts` | `X` | Fallback if no `raw` group |
| `scale.data` | N/A | Not stored in h5ad; recompute with `ScaleData()` |

> **Note**: The `data` slot always receives `X` (which typically contains log-normalized values in scanpy workflows). If the h5ad file has a `raw` group, its `X` matrix becomes `counts`. This matches the scanpy convention where `adata.X` holds processed data and `adata.raw.X` holds raw counts. Scaled data (z-scores for ~2000 variable features) is not stored in standard h5ad files and should be recomputed in Seurat using `ScaleData()` after conversion.

**Other data structures:**

| Data Type | Seurat Location | AnnData Location |
|-----------|-----------------|------------------|
| Cell metadata | `meta.data` | `obs` |
| Feature metadata | `meta.features` | `var` |
| UMAP coords | `reductions$umap` | `obsm['X_umap']` |
| PCA coords | `reductions$pca` | `obsm['X_pca']` |
| Variable features | `VariableFeatures()` | `var['highly_variable']` |
| Spatial coords | `GetTissueCoordinates()` | `obsm['spatial']` |
| Spatial images | `Images()` | `uns['spatial'][lib]['images']` |
| Neighbor graphs | `Graphs()` | `obsp['distances'/'connectivities']` |

### Metadata Column Mapping

Common column name conventions differ between Seurat and scanpy workflows:

| Seurat (`meta.data`) | AnnData (`obs`) | Description |
|---------------------|-----------------|-------------|
| `seurat_clusters` | `leiden` / `louvain` | Cluster assignments |
| `orig.ident` | `batch` / `sample` | Sample identifier |
| `nCount_RNA` | `n_counts` / `total_counts` | Total UMI per cell |
| `nFeature_RNA` | `n_genes` / `n_genes_by_counts` | Genes detected per cell |
| `percent.mt` | `pct_counts_mt` / `percent_mito` | Mitochondrial fraction |
| `cell_type` | `cell_type` / `celltype` | Cell type annotations |
| `Phase` | `phase` / `cell_cycle_phase` | Cell cycle phase |

### Column Name Standardization Option

By default, column names are preserved exactly. Use `standardize = TRUE` for automatic name conversion to scanpy conventions:

| Seurat Name | scanpy Name (with `standardize=TRUE`) |
|-------------|---------------------------------------|
| `seurat_clusters` | `clusters` |
| `nCount_RNA` | `n_counts` |
| `nFeature_RNA` | `n_genes` |
| `percent.mt` | `percent_mito` |

```r
# Preserve original names (default)
scConvert("data.h5Seurat", dest = "h5ad")

# Convert to scanpy naming conventions
scConvert("data.h5Seurat", dest = "h5ad", standardize = TRUE)
```

### Expression Scale Handling

**Automatic Layer Detection**: scConvert maps layers based on h5ad structure:

| h5ad Source | Seurat Destination | Condition |
|-------------|-------------------|-----------|
| `X` | `data` | Always |
| `raw/X` | `counts` | If `raw` group exists |
| `X` | `counts` | Fallback if no `raw` |

**When Scales Match**: If h5ad follows scanpy conventions (`X` = log-normalized, `raw.X` = counts), no additional processing needed after conversion.

**When Scales Differ**: If `X` contains raw counts instead of normalized data:

```r
# Normalize after conversion
seurat_obj <- NormalizeData(seurat_obj)
```

> **Warning**: Do NOT normalize in both Python and R - this double-normalizes data.

### Indexing Conventions

Python uses 0-based indexing; R uses 1-based. scConvert handles this automatically:

| Data Type | h5ad (Python) | Seurat (R) |
|-----------|---------------|------------|
| Categorical codes | 0-indexed | 1-indexed factors |
| Cluster labels | Unchanged | Unchanged |
| Sparse matrix indices | 0-based (CSR/CSC) | 0-based (dgCMatrix) |

**Example**: Cluster 0 in scanpy remains labeled "0" in Seurat, but internally stored as factor level 1.

### Structure Verification

After conversion, verify data integrity:

**h5ad -> Seurat:**

```r
# Check layers and dimensions
cat("Layers:", paste(Layers(seurat_obj), collapse = ", "), "\n")
cat("Cells:", ncol(seurat_obj), "Genes:", nrow(seurat_obj), "\n")

# Verify data ranges (log-normalized data typically 0-6)
cat("Data range:", range(GetAssayData(seurat_obj, layer = "data")[1:100, 1:10]), "\n")

# Check metadata preserved
head(seurat_obj[[]])
```

**Seurat -> h5ad:**

```python
import scanpy as sc
adata = sc.read_h5ad("converted.h5ad")
print(adata)
print("obs columns:", list(adata.obs.columns))
print("X range:", adata.X.min(), "-", adata.X.max())
```

## Session Info

```{r sessionInfo, purl = FALSE}
sessionInfo()
```

```{r cleanup-final, echo=FALSE, results='hide', purl = FALSE}
outs <- c(
  "pbmc3k.h5Seurat", "pbmc3k.h5ad",
  "stxBrain.h5Seurat", "stxBrain.h5ad", "stxBrain_direct.h5ad",
  "crc_sample.h5ad", "crc_normalized.h5ad", "crc_normalized.h5seurat",
  "visium_colon.h5ad", "visium_normalized.h5ad", "visium_normalized.h5seurat",
  "cbmc_rna.h5seurat", "cbmc_rna.h5ad", "cbmc_adt.h5seurat", "cbmc_adt.h5ad"
)
existing_files <- outs[file.exists(outs)]
if (length(existing_files) > 0) file.remove(existing_files)
```
