---
title: "Working with SpatialData (.spatialdata.zarr)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with SpatialData (.spatialdata.zarr)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  dev = "png",
  fig.width = 8,
  fig.height = 6,
  purl = FALSE
)
# SpatialData zarr I/O is pure R — always available
has_spatialdata <- TRUE
has_scanpy <- FALSE
if (requireNamespace("reticulate", quietly = TRUE)) {
  scverse_path <- "/opt/homebrew/Caskroom/miniconda/base/envs/scverse"
  if (dir.exists(scverse_path)) {
    reticulate::use_condaenv("scverse", required = FALSE)
    Sys.setenv(
      OMP_NUM_THREADS = "1", OPENBLAS_NUM_THREADS = "1",
      MKL_NUM_THREADS = "1", NUMBA_THREADING_LAYER = "tbb"
    )
    has_scanpy <- reticulate::py_module_available("scanpy")
    if (has_scanpy) {
      reticulate::py_run_string("import matplotlib; matplotlib.use('Agg')")
    }
  }
}
```

## Overview

[SpatialData](https://spatialdata.scverse.org/) is the scverse standard for
representing spatial omics data. It uses a Zarr-based directory layout that
combines OME-NGFF images, geometric shapes, transcript-level point
annotations, segmentation labels, and anndata expression tables into a single
hierarchical store. Technologies such as Visium, MERFISH, Xenium, Slide-seq,
CODEX, and Stereo-seq all have SpatialData representations, making it a
unifying format across the spatial omics ecosystem.

scConvert provides native R support for reading and writing SpatialData stores
with no Python dependency, and includes direct pair converters for moving data
between SpatialData and other formats (h5ad, h5Seurat, Zarr).

```{r packages, eval = has_spatialdata}
library(Seurat)
library(scConvert)
```

## SpatialData store layout

A `.spatialdata.zarr` directory follows this structure:

```
sample.spatialdata.zarr/
+-- .zattrs              # {"spatialdata_attrs": {"version": "0.2.0"}}
+-- .zgroup              # {"zarr_format": 2}
+-- images/              # OME-NGFF multiscale tissue images
+-- labels/              # Segmentation masks (cell/nucleus outlines)
+-- points/              # Point annotations (e.g., transcript coordinates)
+-- shapes/              # Geometric shapes (spot positions with radii)
+-- tables/              # AnnData expression tables (one or more)
    +-- table/           # Default table (standard anndata zarr)
```

The `tables/table/` subdirectory is a standard anndata zarr store, so
scConvert's existing Zarr infrastructure handles the expression data, metadata,
embeddings, and graphs. The SpatialData-specific elements (shapes, points,
images, labels) are handled by dedicated readers and writers.

## Create a demo SpatialData store

We create a Seurat object with spatial coordinates and write it as a
SpatialData store using `writeSpatialData()`:

```{r create_demo, eval = has_spatialdata}
library(Seurat)
set.seed(42)

# Synthetic spatial expression data: 100 genes × 200 cells
counts <- Matrix::rsparsematrix(100, 200, density = 0.05, rand.x = function(n) rpois(n, 5) + 1)
dimnames(counts) <- list(paste0("Gene", seq_len(100)), paste0("Cell", seq_len(200)))
demo <- CreateSeuratObject(counts = counts, min.cells = 0, min.features = 0)

# Add spatial coordinates and metadata
demo$spatial_x <- runif(200, 0, 1000)
demo$spatial_y <- runif(200, 0, 1000)
demo$cell_type <- factor(sample(c("Epithelial", "Stromal", "Immune"), 200, replace = TRUE))
demo$region <- "tissue_section"

# Write as SpatialData
writeSpatialData(demo, "demo.spatialdata.zarr", overwrite = TRUE, verbose = TRUE)
```

## Reading SpatialData

Use `readSpatialData()` to load the store into a Seurat object:

```{r read_basic, eval = has_spatialdata}
obj <- readSpatialData("demo.spatialdata.zarr", verbose = TRUE)
obj
```

### What gets preserved

The reader extracts the following components:

| SpatialData Element | Seurat Destination | Notes |
|--------------------|--------------------|-------|
| `tables/table/` | Expression assay, metadata, reductions | Via `readZarr()` |
| `shapes/{region}/` | Spatial coordinates (FOV/image) | x, y, radius columns |
| `points/{region}/` | `misc$points_{name}` | Transcript-level data |
| `images/{name}/` | Image object (if coords present) | OME-NGFF, normalized to [0,1] |
| `labels/` | `misc$__spatialdata_labels__` | Names only (see Limitations) |

SpatialData attributes (`region`, `region_key`, `instance_key`) are stored in
`misc$__spatialdata_attrs__` and used during write-back to ensure round-trip
fidelity.

### Inspect the Seurat object

```{r inspect, eval = has_spatialdata}
cat("Cells:    ", ncol(obj), "\n")
cat("Features: ", nrow(obj), "\n")
cat("Layers:   ", paste(Layers(obj), collapse = ", "), "\n")
cat("Images:   ", paste(Images(obj), collapse = ", "), "\n")

# SpatialData metadata preserved for roundtrip
sd_attrs <- obj@misc[["__spatialdata_attrs__"]]
if (!is.null(sd_attrs)) {
  cat("Region:       ", paste(sd_attrs$region, collapse = ", "), "\n")
  cat("Region key:   ", sd_attrs$region_key, "\n")
  cat("Instance key: ", sd_attrs$instance_key, "\n")
}

# Points stored in misc
point_keys <- grep("^points_", names(obj@misc), value = TRUE)
if (length(point_keys) > 0) {
  for (pk in point_keys) {
    cat("Points (", pk, "): ", nrow(obj@misc[[pk]]), " entries\n", sep = "")
  }
}

# Label names
label_names <- obj@misc[["__spatialdata_labels__"]]
if (!is.null(label_names)) {
  cat("Labels:   ", paste(label_names, collapse = ", "), "\n")
}
```

### Specifying a table

SpatialData stores can contain multiple tables. Use the `table` argument to
select which one to read:

```{r read_table, eval = FALSE}
# Read the default table
obj <- readSpatialData("multi_table.spatialdata.zarr", table = "table")

# Read a specific table
obj <- readSpatialData("multi_table.spatialdata.zarr", table = "rna_counts")
```

### Skipping images

For large stores where only the expression data and coordinates are needed,
set `images = FALSE` to skip OME-NGFF image reading:

```{r read_no_images, eval = FALSE}
obj <- readSpatialData("large_sample.spatialdata.zarr", images = FALSE)
```

## Writing SpatialData

Use `writeSpatialData()` to write the Seurat object back as a SpatialData zarr
store, completing the round trip:

```{r write_basic, eval = has_spatialdata}
writeSpatialData(obj, "demo_roundtrip.spatialdata.zarr", overwrite = TRUE)
cat("Written to: demo_roundtrip.spatialdata.zarr\n")
cat("Store size: ",
    round(sum(file.info(
      list.files("demo_roundtrip.spatialdata.zarr",
                 recursive = TRUE, full.names = TRUE)
    )$size) / 1024^2, 1), " MB\n")
```

The writer creates the full SpatialData directory layout:

| Seurat Source | SpatialData Destination | Notes |
|--------------|------------------------|-------|
| Expression, metadata, reductions | `tables/table/` | Via `writeZarr()` |
| Tissue coordinates | `shapes/{region}/` | DataFrame with x, y, radius |
| `misc$points_*` | `points/{name}/` | Preserved from read |
| Image arrays | `images/{name}/` | OME-NGFF with multiscales metadata |

### Verify the roundtrip

```{r verify_roundtrip, eval = has_spatialdata}
obj_rt <- readSpatialData("demo_roundtrip.spatialdata.zarr", verbose = FALSE)

cat("Original cells:   ", ncol(obj), "\n")
cat("Roundtrip cells:  ", ncol(obj_rt), "\n")
cat("Original features:", nrow(obj), "\n")
cat("Roundtrip features:", nrow(obj_rt), "\n")
cat("Dimensions match: ", ncol(obj) == ncol(obj_rt) && nrow(obj) == nrow(obj_rt), "\n")
```

## Direct pair converters

scConvert provides six direct conversion functions that combine reading and
writing in a single call, without requiring you to manage intermediate Seurat
objects.

### SpatialData to h5ad

```{r direct_to_h5ad, eval = has_spatialdata}
SpatialDataToH5AD("demo.spatialdata.zarr", "demo_spatial.h5ad", overwrite = TRUE)
cat("h5ad size:", round(file.size("demo_spatial.h5ad") / 1024^2, 1), "MB\n")
```

### All six converters

```{r direct_converters, eval = FALSE}
# SpatialData <-> h5ad
SpatialDataToH5AD("sample.spatialdata.zarr", "sample.h5ad")
H5ADToSpatialData("sample.h5ad", "sample.spatialdata.zarr")

# SpatialData <-> h5Seurat
SpatialDataToH5Seurat("sample.spatialdata.zarr", "sample.h5seurat")
H5SeuratToSpatialData("sample.h5seurat", "sample.spatialdata.zarr")

# SpatialData <-> standard Zarr (anndata zarr without SpatialData wrapper)
SpatialDataToZarr("sample.spatialdata.zarr", "sample.zarr")
ZarrToSpatialData("sample.zarr", "sample.spatialdata.zarr")
```

All six functions accept `overwrite = TRUE` and `verbose = TRUE/FALSE`
arguments. The `SpatialDataToH5AD`, `SpatialDataToH5Seurat`, and
`SpatialDataToZarr` functions also accept a `table` argument to specify which
table to extract from the store.

These converters also work through the `scConvert()` dispatcher:

```{r scconvert_dispatch, eval = FALSE}
scConvert("sample.spatialdata.zarr", dest = "sample.h5ad", overwrite = TRUE)
scConvert("sample.h5ad", dest = "output.spatialdata.zarr", overwrite = TRUE)
```

## Python interoperability

SpatialData stores written by scConvert are readable by the Python
[spatialdata](https://spatialdata.scverse.org/) library and its ecosystem
(squidpy, napari-spatialdata).

### Read the h5ad output in Python

The h5ad produced by `SpatialDataToH5AD()` is a standard AnnData file readable
by scanpy:

```{python verify_h5ad_py, eval = has_scanpy}
import scanpy as sc

adata = sc.read_h5ad("demo_spatial.h5ad")
print(adata)
print(f"\nobs columns: {list(adata.obs.columns)}")
print(f"Cells: {adata.n_obs}, Genes: {adata.n_vars}")
```

## Known limitations

- **Labels/segmentation masks**: scConvert records the names of label layers
  found in `labels/` but does not read the full mask arrays. These are
  typically large (cell/nucleus segmentation at pixel resolution) and have no
  natural representation in Seurat. Label names are preserved in
  `misc$__spatialdata_labels__` for metadata purposes.

- **Multiple tables**: Only one table is read per call. If your SpatialData
  store contains multiple tables (e.g., separate RNA and protein assays), call
  `readSpatialData()` once per table and merge the resulting Seurat objects.

- **Large images**: OME-NGFF images larger than 100 million pixels are skipped
  to avoid memory issues. Use `images = FALSE` and work with coordinates only,
  or downsample the image externally before reading.

- **Coordinate transforms**: SpatialData supports affine and sequence
  coordinate transformations between elements. These are not currently applied
  during reading; coordinates are used as-is from the zarr arrays.

## See also

- [Spatial Transcriptomics: Visium](spatial-visium.html) -- Visium-specific
  workflows with tissue images and squidpy validation
- [Spatial Technologies](spatial-technologies.html) -- support for MERFISH,
  Slide-seq, Xenium, CODEX, and other platforms
- [Conversions: Seurat and Zarr](convert-zarr.html) -- general Zarr format
  details, streaming conversion, and Python interop

## Session Info

```{r session, eval = TRUE}
sessionInfo()
```

```{r cleanup, echo = FALSE, results = "hide"}
unlink("demo.spatialdata.zarr", recursive = TRUE)
unlink("demo_roundtrip.spatialdata.zarr", recursive = TRUE)
unlink("demo_spatial.h5ad")
```
