---
title: "In-Memory vs On-Disk Conversion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{In-Memory vs On-Disk Conversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

```{r libraries, message = FALSE, warning = FALSE}
library(scConvert)
library(Seurat)
library(ggplot2)
```

scConvert provides three conversion modes with different speed and memory
trade-offs. This vignette explains when to use each, and compares Seurat
serialization formats for disk space.

## Three conversion modes

```{r load-demo, eval = TRUE}
obj <- readRDS(system.file("extdata", "pbmc_demo.rds", package = "scConvert"))
DimPlot(obj, group.by = "seurat_annotations", label = TRUE, pt.size = 1) +
  ggtitle("500 PBMCs — demo dataset") + NoLegend()
```

### 1. In-memory (hub path)

The default mode loads data into a Seurat object, then saves to the target
format. This supports all format pairs but requires the full dataset in RAM.

```{r hub-path, eval = TRUE}
# Seurat object -> h5ad (in-memory)
h5ad_path <- file.path(tempdir(), "pbmc_hub.h5ad")
t1 <- system.time(writeH5AD(obj, h5ad_path, verbose = FALSE))
cat("In-memory write:", round(t1["elapsed"], 2), "s\n")
```

Use in-memory conversion when you need to manipulate the data in R, or when
the source/destination formats don't share the same on-disk layout.

### 2. On-disk streaming (R)

For Zarr conversions, scConvert can copy data field-by-field between files
without ever constructing a Seurat object. This keeps memory usage constant
regardless of dataset size.

```{r streaming, eval = TRUE}
# h5ad -> zarr (streaming, no Seurat object in memory)
zarr_path <- file.path(tempdir(), "pbmc_stream.zarr")
t2 <- system.time(H5ADToZarr(h5ad_path, zarr_path, stream = TRUE, verbose = FALSE))
cat("Streaming h5ad -> zarr:", round(t2["elapsed"], 2), "s\n")
```

Available streaming converters:

| Function | Direction |
|----------|-----------|
| `H5ADToZarr()` | h5ad → zarr |
| `ZarrToH5AD()` | zarr → h5ad |
| `H5SeuratToZarr()` | h5Seurat → zarr |
| `ZarrToH5Seurat()` | zarr → h5Seurat |

All accept `stream = TRUE` (the default) to bypass the Seurat intermediate.

### 3. C binary (on-disk, fastest)

For HDF5-based format pairs (h5ad, h5Seurat, h5mu, Loom), the compiled C
binary copies datasets directly at the HDF5 level. It uses constant memory
and is typically 10--50x faster than the R path.

```{r cli-convert, eval = TRUE}
h5s_path <- file.path(tempdir(), "pbmc_cli.h5seurat")
writeH5Seurat(obj, h5s_path, overwrite = TRUE, verbose = FALSE)

h5ad_cli <- file.path(tempdir(), "pbmc_cli.h5ad")
t3 <- system.time(scConvert_cli(h5s_path, h5ad_cli, verbose = FALSE))
cat("C binary h5seurat -> h5ad:", round(t3["elapsed"], 2), "s\n")
```

Build the binary with `cd src && make` (requires HDF5 headers).

### Performance at scale

On synthetic sparse h5ad files (20K genes, 5% density), median of 3 runs on
Apple M4 Max:

| Operation | 100K cells | 500K cells |
|-----------|-----------|-----------|
| **C binary** (h5ad ↔ h5seurat) | 0.19 s | 0.63 s |
| **Streaming** (h5ad → zarr) | ~1 s | ~5 s |
| **In-memory** read h5ad | 2.56 s | 12.68 s |
| **In-memory** write h5ad (gzip=0) | 0.61 s | 3.29 s |

The C binary stays under 1 second even at 500K cells because it never
decompresses the expression matrix — it copies HDF5 chunks directly.

## Seurat serialization formats

A Seurat object can be saved in several R-native formats. All preserve the
full object structure (assays, reductions, graphs, metadata, images).

```{r save-formats, eval = TRUE}
rds_path <- file.path(tempdir(), "pbmc.rds")
h5s_path2 <- file.path(tempdir(), "pbmc.h5seurat")

saveRDS(obj, rds_path)
writeH5Seurat(obj, h5s_path2, overwrite = TRUE, verbose = FALSE)

sizes <- data.frame(
  Format = character(), Size_KB = numeric(), stringsAsFactors = FALSE
)
sizes <- rbind(sizes, data.frame(Format = "RDS (.rds)", Size_KB = file.size(rds_path) / 1024))
sizes <- rbind(sizes, data.frame(Format = "h5Seurat (.h5seurat)", Size_KB = file.size(h5s_path2) / 1024))

if (requireNamespace("qs2", quietly = TRUE)) {
  qs2_path <- file.path(tempdir(), "pbmc.qs2")
  qs2::qs_save(obj, qs2_path)
  sizes <- rbind(sizes, data.frame(Format = "qs2 (.qs2)", Size_KB = file.size(qs2_path) / 1024))
}

# Also save as RData
rdata_path <- file.path(tempdir(), "pbmc.RData")
save(obj, file = rdata_path)
sizes <- rbind(sizes, data.frame(Format = "RData (.RData)", Size_KB = file.size(rdata_path) / 1024))

sizes$Size_MB <- round(sizes$Size_KB / 1024, 2)
sizes$Size_KB <- round(sizes$Size_KB, 0)
knitr::kable(sizes[, c("Format", "Size_MB")], col.names = c("Format", "Size (MB)"))
```

### Format comparison

| Format | Compression | Random access | Language |
|--------|------------|---------------|----------|
| **RDS** | gzip | No (full load) | R only |
| **RData** | gzip | No (full load) | R only |
| **qs2** | zstd | No (full load) | R only |
| **h5Seurat** | gzip (HDF5) | Yes (selective load) | R, C |

Key differences:

- **qs2** is among the fastest R serialization formats (2--5x faster than
  `saveRDS`), with similar or better compression. Use it for local caching.
  The original **qs** package was archived from CRAN in 2026 and superseded
  by **qs2**.
- **h5Seurat** is the only format that supports **selective loading** — read
  just one assay or one reduction without loading the entire object.
- **RDS / RData** are universally available but slower for large objects.

### Exchange formats

For sharing data with Python or other tools, use cross-language formats:

```{r exchange-formats, eval = TRUE}
h5ad_path2 <- file.path(tempdir(), "pbmc_exchange.h5ad")
writeH5AD(obj, h5ad_path2, verbose = FALSE)

loom_path <- file.path(tempdir(), "pbmc_exchange.loom")
writeLoom(obj, loom_path, verbose = FALSE)

zarr_path2 <- file.path(tempdir(), "pbmc_exchange.zarr")
writeZarr(obj, zarr_path2, verbose = FALSE)

exchange <- data.frame(
  Format = c("h5ad (AnnData)", "Loom", "Zarr"),
  Size_MB = round(c(
    file.size(h5ad_path2),
    file.size(loom_path),
    sum(file.info(list.files(zarr_path2, recursive = TRUE, full.names = TRUE))$size)
  ) / 1024^2, 2),
  Ecosystem = c("scanpy, CELLxGENE", "loompy, velocyto", "cloud / AnnData")
)
knitr::kable(exchange)
```

## When to use what

| Goal | Recommended |
|------|------------|
| Fast local save/load in R | `qs2::qs_save()` / `qs2::qs_read()` |
| Selective loading (big data) | `writeH5Seurat()` / `readH5Seurat()` |
| Share with Python | `writeH5AD()` or C binary |
| Batch convert many files | C binary (`scconvert`) |
| Cloud storage | `writeZarr()` |
| Convert without loading | `scConvert_cli()` or streaming converters |

## Clean up

```{r cleanup, include = FALSE}
unlink(c(h5ad_path, h5ad_cli, h5s_path, h5s_path2, rds_path, h5ad_path2, loom_path), recursive = FALSE)
unlink(c(zarr_path, zarr_path2), recursive = TRUE)
if (exists("qs_path")) unlink(qs_path)
if (exists("qs2_path")) unlink(qs2_path)
unlink(rdata_path)
```
