---
title: "Convert to Loom Format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convert to Loom Format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

The [Loom format](http://loompy.org/) is an HDF5-based file format used by
loompy and RNA velocity tools (velocyto, scVelo). scConvert provides
`readLoom()` and `writeLoom()` for round-trip conversion between Seurat objects
and Loom files with no external dependencies.

## Read a Loom file

scConvert ships a 500-cell PBMC dataset in Loom format. `readLoom()` loads it
directly into a Seurat object.

```{r read-loom, eval = TRUE}
loom_file <- system.file("extdata", "pbmc_demo.loom", package = "scConvert")
pbmc <- readLoom(loom_file)
pbmc
```

The reader reconstructs cell metadata, gene metadata, and dimensional
reductions from the Loom column and row attributes:

```{r inspect, eval = TRUE}
cat("Cells:", ncol(pbmc), "\n")
cat("Genes:", nrow(pbmc), "\n")
cat("Reductions:", paste(Reductions(pbmc), collapse = ", "), "\n")
cat("Metadata columns:", paste(colnames(pbmc[[]]), collapse = ", "), "\n")
```

The nine annotated cell types are preserved:

```{r dimplot, eval = TRUE}
DimPlot(pbmc, reduction = "umap", group.by = "seurat_annotations",
        label = TRUE, pt.size = 0.5) + NoLegend()
```

Expression data is fully available. LYZ is a monocyte marker:

```{r featureplot, eval = TRUE}
FeaturePlot(pbmc, features = "LYZ", pt.size = 0.5)
```

## Write a Seurat object to Loom

`writeLoom()` saves the default assay's expression data as the main matrix,
with cell metadata as column attributes and gene metadata as row attributes.
Dimensional reductions are stored as additional column attributes.

```{r write-loom, eval = TRUE}
pbmc_seurat <- readRDS(system.file("extdata", "pbmc_demo.rds", package = "scConvert"))
loom_path <- tempfile(fileext = ".loom")
writeLoom(pbmc_seurat, filename = loom_path, overwrite = TRUE)
cat("Loom file size:", round(file.size(loom_path) / 1e6, 1), "MB\n")
```

## Verify the round-trip

Read the written Loom file back and compare the UMAP projections.

```{r roundtrip-read, eval = TRUE}
pbmc_rt <- readLoom(loom_path)
```

```{r roundtrip-compare, eval = TRUE, fig.width = 10, fig.height = 5}
library(patchwork)
p1 <- DimPlot(pbmc_seurat, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.5) + NoLegend() + ggtitle("Original (.rds)")
p2 <- DimPlot(pbmc_rt, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.5) + NoLegend() + ggtitle("Round-trip (.loom)")
p1 + p2
```

## What is preserved

Loom is a simpler format than h5ad or h5Seurat. Here is a summary of what
round-trips and what does not:

| Component | Preserved? | Notes |
|-----------|:----------:|-------|
| Expression matrix | Yes | Stored as `/matrix` |
| Raw counts | Yes | Stored in `/layers/counts` |
| Cell metadata | Yes | Each column becomes a `/col_attrs` entry |
| Gene metadata | Yes | Each column becomes a `/row_attrs` entry |
| PCA / UMAP embeddings | Yes | Stored as column attributes |
| Nearest-neighbor graphs | No | Not native to Loom; recompute with `FindNeighbors()` |

## Per-cluster expression

A violin plot confirms that per-cluster expression distributions are preserved
through Loom conversion:

```{r vlnplot, eval = TRUE}
VlnPlot(pbmc_rt, features = "LYZ", group.by = "seurat_annotations", pt.size = 0) +
  NoLegend()
```

## Python interop (optional)

The Loom files produced by `writeLoom()` are compatible with loompy, scanpy,
and scVelo.

```{python loompy-read, eval = FALSE}
import loompy

with loompy.connect("pbmc_demo.loom") as ds:
    print(f"Shape: {ds.shape[0]} genes x {ds.shape[1]} cells")
    print(f"Row attributes: {list(ds.ra.keys())}")
    print(f"Column attributes: {list(ds.ca.keys())[:10]}")
    print(f"Layers: {list(ds.layers.keys())}")
```

```{python scanpy-loom, eval = FALSE}
import scanpy as sc

adata = sc.read_loom("pbmc_demo.loom", sparse=True, cleanup=False)
print(adata)
```

## Clean up

```{r cleanup, eval = TRUE}
unlink(loom_path)
```
