---
title: "Getting Started with scConvert"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with scConvert}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  message   = FALSE,
  warning   = FALSE
)
```

```{r libraries}
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
})
set.seed(42L)
```

scConvert converts between single-cell data formats entirely in R, with no
Python dependency. This vignette walks through the core API in under five
minutes. For real-data demos, Python interop, and CLI performance benchmarks
see the [full articles](https://mianaz.github.io/scConvert/).

## Supported formats

| Format | Extension | Ecosystem | Read | Write |
|--------|-----------|-----------|:----:|:-----:|
| AnnData | `.h5ad` | scanpy, CELLxGENE | yes | yes |
| h5Seurat | `.h5seurat` | Seurat | yes | yes |
| MuData | `.h5mu` | muon (multimodal) | yes | yes |
| Loom | `.loom` | loompy, scVelo | yes | yes |
| Zarr | `.zarr` | cloud AnnData | yes | yes |
| TileDB-SOMA | `soma://` | CELLxGENE Census | yes | yes |
| SpatialData | `.zarr` | scverse spatial | yes | yes |
| RDS | `.rds` | R native | yes | yes |
| SingleCellExperiment | in-memory | Bioconductor | yes | -- |

## Quick conversion with scConvert()

`scConvert()` is a universal dispatcher: give it a source and a destination
path and it picks the fastest conversion path automatically.

```{r quick-convert}
h5ad_file   <- system.file("testdata", "pbmc_small.h5ad",   package = "scConvert")
h5seurat_out <- file.path(tempdir(), "pbmc.h5seurat")
zarr_out     <- file.path(tempdir(), "pbmc.zarr")

t0 <- proc.time()
scConvert(h5ad_file, dest = h5seurat_out, overwrite = TRUE)
cat(sprintf("h5ad -> h5seurat: %.2fs\n", (proc.time() - t0)[["elapsed"]]))

t0 <- proc.time()
scConvert(h5ad_file, dest = zarr_out, overwrite = TRUE)
cat(sprintf("h5ad -> zarr:     %.2fs\n", (proc.time() - t0)[["elapsed"]]))
```

## Loading files into Seurat

Each reader returns a standard Seurat object.

```{r read-h5ad}
obj <- readH5AD(h5ad_file)
cat(sprintf("Loaded: %d cells x %d genes\n", ncol(obj), nrow(obj)))
cat(sprintf("Reductions: %s\n", paste(names(obj@reductions), collapse = ", ")))
```

```{r read-h5seurat}
obj2 <- readH5Seurat(h5seurat_out)
cat(sprintf("h5seurat: %d cells x %d genes\n", ncol(obj2), nrow(obj2)))
```

```{r read-zarr}
obj3 <- readZarr(zarr_out, verbose = FALSE)
cat(sprintf("zarr:     %d cells x %d genes\n", ncol(obj3), nrow(obj3)))
```

## Writing files from Seurat

```{r write-files}
h5ad_out  <- file.path(tempdir(), "output.h5ad")
h5s_out   <- file.path(tempdir(), "output.h5seurat")
zarr_out2 <- file.path(tempdir(), "output.zarr")

t0 <- proc.time(); writeH5AD(obj, h5ad_out, verbose = FALSE)
cat(sprintf("writeH5AD:    %.2fs\n", (proc.time() - t0)[["elapsed"]]))

t0 <- proc.time(); writeH5Seurat(obj, h5s_out, overwrite = TRUE, verbose = FALSE)
cat(sprintf("writeH5Seurat: %.2fs\n", (proc.time() - t0)[["elapsed"]]))
```

```{r file-sizes}
sizes <- data.frame(
  Format   = c("h5ad", "h5Seurat"),
  Size_MB  = round(c(file.size(h5ad_out), file.size(h5s_out)) / 1024^2, 2)
)
knitr::kable(sizes, col.names = c("Format", "Size (MB)"))
```

## Next steps

- **[Convert Between Seurat and AnnData](articles/convert-anndata.html)** --
  real PBMC3k data, full Seurat pipeline, and live Python scanpy validation.
- **[CLI & Atlas-Scale Conversion](articles/cli-usage.html)** -- the C binary
  converts 592K cells in under a second.
- **[Zarr Format](articles/convert-zarr.html)** -- cloud-native storage with
  live `zarr.open()` Python validation.
- **[Spatial Visium](articles/spatial-visium.html)** -- real mouse brain data,
  Seurat spatial plots, and squidpy validation.
- **[Compare Tools](articles/compare-tools.html)** -- benchmark vs anndataR,
  schard, zellkonverter, scNT.

```{r cleanup, include = FALSE}
unlink(c(h5seurat_out, h5ad_out, h5s_out))
unlink(c(zarr_out, zarr_out2), recursive = TRUE)
```
