---
title: "Getting Started with scConvert"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with scConvert}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

```{r libraries}
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
})
```

scConvert converts between single-cell data formats entirely in R, with no
Python dependency. This vignette walks through installation, basic conversion,
and file I/O in under five minutes.

## Installation

Install scConvert from GitHub:

```{r install, eval = FALSE}
remotes::install_github("mianaz/scConvert")
```

## Quick conversion with scConvert()

The `scConvert()` function is a universal dispatcher: give it a source and a
destination, and it picks the fastest conversion path automatically. File
formats are detected from extensions.

```{r quick-convert, eval = TRUE}
h5ad_file <- system.file("testdata", "pbmc_small.h5ad", package = "scConvert")

h5seurat_out <- file.path(tempdir(), "pbmc.h5seurat")
scConvert(h5ad_file, dest = h5seurat_out, overwrite = TRUE)

zarr_out <- file.path(tempdir(), "pbmc.zarr")
scConvert(h5ad_file, dest = zarr_out, overwrite = TRUE)
```

`scConvert()` works for any supported format pair -- h5ad, h5Seurat, Zarr,
Loom, RDS, and more.

## Loading files into Seurat

When you need to work with the data in R, use the format-specific readers.
Each returns a standard Seurat object.

### From h5ad

```{r read-h5ad, eval = TRUE}
obj <- readH5AD(h5ad_file)
obj
```

```{r inspect, eval = TRUE}
dim(obj)
head(obj[[]], n = 4)
```

### From h5Seurat

```{r read-h5seurat, eval = TRUE}
obj2 <- readH5Seurat(h5seurat_out)
obj2
```

### From Zarr

```{r read-zarr, eval = TRUE}
obj3 <- readZarr(zarr_out)
obj3
```

## Writing files from Seurat

Starting from any Seurat object, write to the format your collaborators or
downstream tools need.

```{r write-files, eval = TRUE}
h5ad_out <- file.path(tempdir(), "output.h5ad")
writeH5AD(obj, h5ad_out, verbose = FALSE)

h5s_out <- file.path(tempdir(), "output.h5seurat")
writeH5Seurat(obj, h5s_out, overwrite = TRUE, verbose = FALSE)

zarr_out2 <- file.path(tempdir(), "output.zarr")
writeZarr(obj, zarr_out2, verbose = FALSE)
```

```{r file-sizes, eval = TRUE}
sizes <- data.frame(
  Format = c("h5ad", "h5Seurat", "Zarr"),
  Size_MB = round(c(
    file.size(h5ad_out),
    file.size(h5s_out),
    sum(file.info(list.files(zarr_out2, recursive = TRUE, full.names = TRUE))$size)
  ) / 1024^2, 2)
)
knitr::kable(sizes, col.names = c("Format", "Size (MB)"))
```

## Supported formats

| Format | Extension | Ecosystem | Read | Write |
|--------|-----------|-----------|:----:|:-----:|
| AnnData | `.h5ad` | scanpy, CELLxGENE | yes | yes |
| h5Seurat | `.h5seurat` | Seurat | yes | yes |
| MuData | `.h5mu` | muon (multimodal) | yes | yes |
| Loom | `.loom` | loompy, HCA | yes | yes |
| Zarr | `.zarr` | cloud AnnData | yes | yes |
| TileDB-SOMA | `soma://` | CELLxGENE Census | yes | yes |
| SpatialData | `.zarr` | scverse spatial | yes | yes |
| RDS | `.rds` | R native | yes | yes |
| SingleCellExperiment | in-memory | Bioconductor | yes | -- |

**When to use which format:**

- **h5ad** -- sharing with Python users or submitting to CELLxGENE.
- **h5Seurat** -- archiving a Seurat object with selective loading support.
- **Zarr** -- cloud-friendly, chunk-based access (e.g., S3 / GCS).
- **Loom** -- interoperability with loompy, velocyto, or legacy pipelines.
- **RDS** -- quick local save/load within R (no HDF5 dependency).

## Verifying a conversion

After converting, a quick sanity check confirms that dimensions and cell
identifiers are preserved.

```{r verify, eval = TRUE}
original <- readH5AD(h5ad_file)
converted <- readH5Seurat(h5seurat_out)

stopifnot(identical(dim(original), dim(converted)))
stopifnot(identical(sort(colnames(original)), sort(colnames(converted))))

cat("Dimensions match:", paste(dim(original), collapse = " x "), "\n")
cat("Cell names match:", length(colnames(original)), "cells verified\n")
```

## Next steps

- **[Convert Between Seurat and AnnData](convert-anndata.html)** -- layer
  mapping, Python interop, and round-trip verification.
- **[In-Memory vs On-Disk Conversion](conversion-modes.html)** -- hub path,
  streaming, and the C binary for large datasets.
- **[Direct H5AD Loading](h5Seurat-load.html)** -- selective loading and
  BPCells for atlas-scale data.
- **[Multimodal H5MU](multimodal-h5mu.html)** -- CITE-seq and ATAC+RNA via
  MuData.
- **[Zarr Format](convert-zarr.html)** -- cloud-native storage and streaming
  converters.
- **[Spatial Technologies](spatial-technologies.html)** -- Visium, MERFISH,
  and SpatialData support.
- **[CLI Usage](cli-usage.html)** -- the standalone C binary for batch
  conversion without R.

## Clean up

```{r cleanup, include = FALSE}
unlink(c(h5seurat_out, h5ad_out, h5s_out), recursive = FALSE)
unlink(c(zarr_out, zarr_out2), recursive = TRUE)
```
