---
title: "Working with h5Seurat Files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with h5Seurat Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

The h5Seurat format is an HDF5-based file format designed for Seurat objects. It
stores all assays, reductions, graphs, and metadata in a single file and
supports selective loading so you can read only the components you need.

## Save a Seurat object

We start with the shipped 500-cell PBMC demo dataset, which includes PCA, UMAP,
neighbor graphs, and nine annotated cell types.

```{r load-data, eval = TRUE}
pbmc <- readRDS(system.file("extdata", "pbmc_demo.rds", package = "scConvert"))
pbmc
```

```{r dimplot-original, eval = TRUE}
DimPlot(pbmc, reduction = "umap", group.by = "seurat_annotations",
        label = TRUE, pt.size = 0.5) + NoLegend()
```

Write the object to an h5Seurat file with `writeH5Seurat()`. All assays,
dimensional reductions, graphs, and cell metadata are saved automatically.

```{r write-h5seurat, eval = TRUE}
h5seurat_path <- tempfile(fileext = ".h5Seurat")
writeH5Seurat(pbmc, filename = h5seurat_path, overwrite = TRUE)
cat("File size:", round(file.size(h5seurat_path) / 1e6, 1), "MB\n")
```

## Load it back

`readH5Seurat()` reads the full object back into memory by default:

```{r read-full, eval = TRUE}
pbmc2 <- readH5Seurat(h5seurat_path)
pbmc2
```

The loaded object preserves all components -- clusters, reductions, and
expression data look identical to the original:

```{r compare-dimplot, eval = TRUE, fig.width = 10, fig.height = 5}
library(patchwork)
p1 <- DimPlot(pbmc, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.5) + NoLegend() + ggtitle("Original")
p2 <- DimPlot(pbmc2, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.5) + NoLegend() + ggtitle("From h5Seurat")
p1 + p2
```

## Selective loading

For large datasets, you can load only what you need. This is particularly useful
when you have a multi-gigabyte h5Seurat file and only want to make a quick UMAP
plot. The `readH5Seurat()` function accepts four parameters for controlling what
gets loaded:

- `assays`: which assays and layers to load
- `reductions`: which dimensional reductions to load
- `graphs`: which nearest-neighbor graphs to load
- `images`: which spatial images to load

Load only the RNA assay with UMAP, skipping PCA and all graphs:

```{r selective-load, eval = TRUE}
pbmc_light <- readH5Seurat(h5seurat_path, assays = "RNA",
                           reductions = "umap", graphs = FALSE)
pbmc_light
```

The loaded object still supports visualization since UMAP coordinates are
present:

```{r selective-dimplot, eval = TRUE}
DimPlot(pbmc_light, reduction = "umap", group.by = "seurat_annotations",
        label = TRUE, pt.size = 0.5) + NoLegend()
```

You can also control which layers of an assay to load. For example, to load
only the normalized data (skipping raw counts and scaled data):

```{r selective-layers, eval = TRUE}
pbmc_data_only <- readH5Seurat(h5seurat_path,
                               assays = list("RNA" = "data"),
                               reductions = "umap", graphs = FALSE)
cat("Layers loaded:", paste(Layers(pbmc_data_only), collapse = ", "), "\n")
```

## Query file contents with scConnect()

Before loading, you can inspect what is stored in an h5Seurat file without
reading any data into memory. `scConnect()` opens a lightweight connection to
the file and `.index()` summarizes its contents, showing which assays,
reductions, graphs, and images are available.

```{r scconnect, eval = TRUE}
hfile <- scConnect(h5seurat_path)
hfile$index()
hfile$close_all()
```

This is especially useful for large files where you want to decide what to load
before committing memory.

## Compare expression before and after

The expression values are preserved exactly through the h5Seurat round-trip.
Here we compare the LYZ gene (a monocyte marker) in the original and reloaded
objects:

```{r featureplot-compare, eval = TRUE, fig.width = 10, fig.height = 5}
p1 <- FeaturePlot(pbmc, features = "LYZ", pt.size = 0.5) + ggtitle("LYZ - Original")
p2 <- FeaturePlot(pbmc2, features = "LYZ", pt.size = 0.5) + ggtitle("LYZ - From h5Seurat")
p1 + p2
```

A violin plot confirms the per-cluster expression distribution is identical:

```{r vlnplot, eval = TRUE}
VlnPlot(pbmc2, features = "LYZ", group.by = "seurat_annotations", pt.size = 0) +
  NoLegend()
```

## Incremental loading with scAppendData()

If you loaded a partial object and later need additional components, use
`scAppendData()` to add them without reloading from scratch:

```{r append-data, eval = TRUE}
# Start with a minimal load (data layer + UMAP only)
pbmc_partial <- readH5Seurat(h5seurat_path, assays = list("RNA" = "data"),
                             reductions = "umap", graphs = FALSE)
cat("Before append - Reductions:", paste(Reductions(pbmc_partial), collapse = ", "), "\n")

# Now add PCA
pbmc_partial <- scAppendData(h5seurat_path, pbmc_partial,
                             assays = FALSE, reductions = "pca",
                             graphs = FALSE, images = FALSE)
cat("After append  - Reductions:", paste(Reductions(pbmc_partial), collapse = ", "), "\n")
```

## Clean up

```{r cleanup, eval = TRUE}
unlink(h5seurat_path)
```
