---
title: "Working with h5Seurat Files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with h5Seurat Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

The h5Seurat format is an HDF5-based file format designed specifically for
Seurat objects. It stores all assays, reductions, graphs, and metadata in a
single file and supports selective loading so you can read only the components
you need.

## Read an h5Seurat file

scConvert ships a 500-cell PBMC dataset in h5Seurat format. `readH5Seurat()`
loads it directly into a fully functional Seurat object.

```{r read-h5seurat, eval = TRUE}
h5seurat_file <- system.file("extdata", "pbmc_demo.h5seurat", package = "scConvert")
pbmc <- readH5Seurat(h5seurat_file)
pbmc
```

The loaded object contains all nine annotated cell types with PCA, UMAP, and
neighbor graphs:

```{r dimplot, eval = TRUE}
DimPlot(pbmc, reduction = "umap", group.by = "seurat_annotations",
        label = TRUE, pt.size = 0.5) + NoLegend()
```

Expression data is fully available. LYZ is a monocyte marker:

```{r featureplot, eval = TRUE}
FeaturePlot(pbmc, features = "LYZ", pt.size = 0.5)
```

## Write a Seurat object to h5Seurat

`writeH5Seurat()` saves any Seurat object to h5Seurat format, preserving all
assays, dimensional reductions, graphs, and cell metadata.

```{r write-h5seurat, eval = TRUE}
pbmc_seurat <- readRDS(system.file("extdata", "pbmc_demo.rds", package = "scConvert"))
h5seurat_path <- tempfile(fileext = ".h5Seurat")
writeH5Seurat(pbmc_seurat, filename = h5seurat_path, overwrite = TRUE)
cat("File size:", round(file.size(h5seurat_path) / 1e6, 1), "MB\n")
```

## Verify the round-trip

Read the written file back and confirm that the UMAP coordinates, cluster
labels, and expression values are identical to the original.

```{r roundtrip-read, eval = TRUE}
pbmc_rt <- readH5Seurat(h5seurat_path)
```

```{r roundtrip-compare, eval = TRUE, fig.width = 10, fig.height = 5}
library(patchwork)
p1 <- DimPlot(pbmc_seurat, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.5) + NoLegend() + ggtitle("Original (.rds)")
p2 <- DimPlot(pbmc_rt, reduction = "umap", group.by = "seurat_annotations",
              label = TRUE, pt.size = 0.5) + NoLegend() + ggtitle("Round-trip (.h5Seurat)")
p1 + p2
```

## Selective loading

For large datasets, you can load only what you need. This is useful when you
have a multi-gigabyte h5Seurat file and only want to make a quick UMAP plot.

Load only the RNA assay with UMAP, skipping PCA and all graphs:

```{r selective-load, eval = TRUE}
pbmc_light <- readH5Seurat(h5seurat_file, assays = "RNA",
                           reductions = "umap", graphs = FALSE)
pbmc_light
```

The light object still supports visualization since UMAP coordinates are
present:

```{r selective-dimplot, eval = TRUE}
DimPlot(pbmc_light, reduction = "umap", group.by = "seurat_annotations",
        label = TRUE, pt.size = 0.5) + NoLegend()
```

## Query file contents with scConnect()

Before loading any data, you can inspect what is stored in an h5Seurat file
using `scConnect()`. This opens a lightweight connection and `.index()`
summarizes the contents -- which assays, reductions, graphs, and images are
available.

```{r scconnect, eval = TRUE}
hfile <- scConnect(h5seurat_file)
hfile$index()
hfile$close_all()
```

This is especially useful for large files where you want to decide what to load
before committing memory.

## Per-cluster expression

A violin plot shows the LYZ expression distribution across all nine cell types:

```{r vlnplot, eval = TRUE}
VlnPlot(pbmc, features = "LYZ", group.by = "seurat_annotations", pt.size = 0) +
  NoLegend()
```

## Clean up

```{r cleanup, eval = TRUE}
unlink(h5seurat_path)
```
