---
title: "h5Seurat Format Specification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{h5Seurat Format Specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

## Overview

h5Seurat is the native HDF5 format for Seurat objects. It stores all
components -- expression matrices, metadata, reductions, graphs, and spatial
data -- in a single `.h5Seurat` file. This vignette provides a quick example
followed by the format specification.

## Quick example

```{r demo, eval = TRUE, fig.width = 7, fig.height = 5}
obj <- readRDS(system.file("extdata", "pbmc_demo.rds", package = "scConvert"))

h5s_path <- tempfile(fileext = ".h5Seurat")
writeH5Seurat(obj, h5s_path, overwrite = TRUE, verbose = FALSE)

obj_loaded <- readH5Seurat(h5s_path, verbose = FALSE)
cat("Cells:", ncol(obj_loaded), "| Genes:", nrow(obj_loaded), "\n")
cat("Assays:", paste(names(obj_loaded@assays), collapse = ", "), "\n")
cat("Reductions:", paste(names(obj_loaded@reductions), collapse = ", "), "\n")

DimPlot(obj_loaded, reduction = "umap", group.by = "seurat_annotations") +
  ggplot2::ggtitle("Loaded from h5Seurat")
```

## Top-level structure

Every h5Seurat file has the following layout:

| Entry | Type | Required | Description |
|-------|------|:--------:|-------------|
| `cell.names` | Dataset (string) | Yes | Cell barcodes, length = n_cells |
| `meta.data` | Group or dataset | Yes | Cell-level metadata (data frame) |
| `assays/` | Group | Yes | One sub-group per assay |
| `reductions/` | Group | Yes | One sub-group per reduction |
| `graphs/` | Group | Yes | One sub-group per neighbor graph |
| `images/` | Group | No | Spatial image data (Visium, etc.) |
| `misc/` | Group | Yes | Miscellaneous data (list) |
| `tools/` | Group | Yes | Tool-specific results (list) |
| `commands/` | Group | No | Command log |

**Required attributes** on the root group:

| Attribute | Type | Description |
|-----------|------|-------------|
| `project` | String | Seurat project name |
| `active.assay` | String | Default assay name (must exist in `assays/`) |
| `version` | String | Seurat version |

## Assay layout

Each assay is stored under `assays/{name}/`:

| Entry | Type | Required | Description |
|-------|------|:--------:|-------------|
| `features` | Dataset (string) | Yes | Gene/feature names |
| `data` | Sparse group or dense dataset | Yes | Log-normalized expression (genes x cells) |
| `counts` | Sparse group or dense dataset | No | Raw UMI counts |
| `scale.data` | Dense dataset | No | Scaled expression (variable features x cells) |
| `scaled.features` | Dataset (string) | No | Feature names for scale.data |
| `variable.features` | Dataset (string) | No | Highly variable feature names |
| `meta.features` | Group or dataset | No | Feature-level metadata (data frame) |
| `misc` | Group | No | Additional assay data |

**Required attribute:** `key` (string) -- the assay key prefix (e.g., `"rna_"`).

Seurat v5 `Assay5` objects are automatically converted to this layout, with
layers stored as separate matrices.

## Reductions layout

Each reduction is stored under `reductions/{name}/`:

| Entry | Type | Required | Description |
|-------|------|:--------:|-------------|
| `cell.embeddings` | Dense dataset | Yes | Embedding matrix (n_cells x n_components) |
| `feature.loadings` | Dense dataset | No | Gene loadings (PCA only) |
| `feature.loadings.projected` | Dense dataset | No | Projected loadings |
| `misc` | Group | No | Additional data |

**Required attributes:** `active.assay` (string), `key` (string),
`global` (logical).

## Graphs layout

Each graph is stored under `graphs/{name}/` as a sparse matrix group:

| Entry | Type | Description |
|-------|------|-------------|
| `data` | Dataset | Non-zero edge weights |
| `indices` | Dataset (int) | 0-based row indices |
| `indptr` | Dataset (int) | Column pointers (length = n_cells + 1) |

**Required attribute:** `dims` (2 integers) -- matrix dimensions.
**Optional attribute:** `assay.used` (string).

## Common data types

### Sparse matrices

Stored as CSC (compressed sparse column) with three datasets: `data`,
`indices`, `indptr`. This matches the `dgCMatrix` layout in R. The `dims`
attribute records (n_rows, n_cols).

### Factors

Stored as a group with two datasets:

- `levels` -- string dataset of unique levels
- `values` -- integer dataset of 1-based level indices

This avoids HDF5 enum types, which are not supported by all HDF5
implementations.

### Data frames

Stored as either:

- **Dataset** -- compound HDF5 dataset (when no factors are present)
- **Group** -- one dataset per column, with factors as sub-groups

Group-based data frames may include a `colnames` attribute for column ordering
and a `row.names` dataset.

### Logicals

Encoded as integers: `FALSE` = 0, `TRUE` = 1, `NA` = 2. An optional
`s3class = "logical"` attribute enforces type on read.

### Strings

All strings are variable-length UTF-8 encoded, for both datasets and
attributes.

## Clean up

```{r cleanup, include = FALSE}
unlink(h5s_path)
```
