---
title: "Immune Repertoire (TCR/BCR) Data with scConvert"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Immune Repertoire (TCR/BCR) Data with scConvert}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  out.width = "100%",
  dev = "png",
  purl = FALSE
)
# Python ecosystem detection via reticulate
has_scanpy <- FALSE
if (requireNamespace("reticulate", quietly = TRUE)) {
  scverse_path <- "/opt/homebrew/Caskroom/miniconda/base/envs/scverse"
  if (dir.exists(scverse_path)) {
    reticulate::use_condaenv("scverse", required = FALSE)
    Sys.setenv(
      OMP_NUM_THREADS = "1", OPENBLAS_NUM_THREADS = "1",
      MKL_NUM_THREADS = "1", NUMBA_THREADING_LAYER = "tbb"
    )
    has_scanpy <- reticulate::py_module_available("scanpy")
    if (has_scanpy) {
      reticulate::py_run_string("import matplotlib; matplotlib.use('Agg')")
    }
  }
}
```

```{r cleanup, echo=FALSE, results='hide', purl = FALSE}
outs <- c(
  "tcr_annotated.h5ad", "tcr_annotated.h5seurat", "tcr_annotated.h5mu",
  "tcr_annotated.loom", "bcr_annotated.h5ad"
)
existing_files <- outs[file.exists(outs)]
if (length(existing_files) > 0) file.remove(existing_files)
```

## Introduction

Single-cell immune repertoire profiling (TCR/BCR sequencing) is often paired with gene expression (GEX) in multimodal experiments such as 10x Genomics 5' GEX + VDJ. This vignette demonstrates how scConvert preserves immune repertoire metadata during format conversion using a synthetic example built on the shipped `pbmc_small` dataset.

### Key concepts

- **TCR/BCR clonotype data** is stored as cell-level metadata (e.g., `clonotype_id`, `CTgene`, `CTaa`, `CTstrict`)
- In scanpy/scirpy workflows, TCR/BCR annotations live in `obs` (cell metadata) and `obsm` (alignment matrices)
- In Seurat, clonotype data lives in `meta.data` after tools like [scRepertoire](https://github.com/ncborcherding/scRepertoire) or [Immcantation](https://immcantation.readthedocs.io/)
- scConvert preserves all cell metadata during conversion, so TCR/BCR annotations survive roundtrip

```{r packages}
library(Seurat)
library(scConvert)
```

## Creating synthetic TCR metadata

In a real workflow, you would add TCR annotations to your Seurat object using scRepertoire:

```{r real_workflow, eval = FALSE}
# Real workflow (requires Cell Ranger VDJ output):
library(scRepertoire)
tcr <- combineTCR(
  read.csv("filtered_contig_annotations.csv"),
  samples = "Sample1"
)
obj <- combineExpression(tcr, obj, cloneCall = "aa", group.by = "sample")
```

For this vignette, we create synthetic TCR metadata that mimics what `combineExpression()` would produce. This lets us demonstrate format conversion without requiring real VDJ data files.

```{r synthetic_tcr}
obj <- readRDS(system.file("testdata", "pbmc_small.rds", package = "scConvert"))
n_cells <- ncol(obj)

set.seed(42)

# Define 10 synthetic clonotypes with realistic gene/sequence annotations
clonotype_pool <- data.frame(
  clonotype_id = paste0("clonotype", 1:10),
  CTgene = c(
    "TRAV1-2.TRAJ33.TRAC_TRBV6-1.TRBJ2-1.TRBC2",
    "TRAV12-1.TRAJ39.TRAC_TRBV28.TRBJ1-1.TRBC1",
    "TRAV38-2.TRAJ28.TRAC_TRBV5-1.TRBJ2-7.TRBC2",
    "TRAV21.TRAJ12.TRAC_TRBV7-2.TRBJ2-3.TRBC2",
    "TRAV8-6.TRAJ45.TRAC_TRBV19.TRBJ1-2.TRBC1",
    "TRAV26-1.TRAJ37.TRAC_TRBV12-3.TRBJ2-5.TRBC2",
    "TRAV13-1.TRAJ44.TRAC_TRBV4-1.TRBJ1-5.TRBC1",
    "TRAV29.TRAJ20.TRAC_TRBV20-1.TRBJ1-6.TRBC1",
    "TRAV17.TRAJ31.TRAC_TRBV11-2.TRBJ2-2.TRBC2",
    "TRAV3.TRAJ48.TRAC_TRBV9.TRBJ2-4.TRBC2"
  ),
  CTaa = c(
    "CAVMDSNYQLIW_CASSETSGSTDTQYF",
    "CVVNMGGKLIF_CASSQETQYF",
    "CAFMKTSYDKVIF_CASSLGQAYEQYF",
    "CAVRDSNYQLIW_CASSFSTCSANYGYTF",
    "CAVSDYGGSQGNLIF_CASSIRSSYEQYF",
    "CIVRVGNTGKLIF_CASSLEETQYF",
    "CAASRGSTLGRLYF_CASRPGLAGDNEQFF",
    "CAASADYKLSF_CSARDPGQGASYEQYF",
    "CATDAGGTSYGKLTF_CASSLAGGTDTQYF",
    "CAVRDNYGQNFVF_CASSVAGGREQYF"
  ),
  CTnt = c(
    "TGTGCAGTAATGGATTCAAATTACCAGTTAATCTGG_TGTGCCAGCAGTGAAACTAGCGGGTCCACTGACACCCAGTACTTT",
    "TGTGTAGTAAATATGATGGGGGAAAGCTGATATTT_TGTGCCAGCAGCCAAGAAACTCAGTACTTC",
    "TGTGCATTTATGAAGACATCATATGATAAGGTGATTTTT_TGTGCCAGCAGTTTGGGACAGGCGTATGAGCAGTACTTC",
    "TGTGCAGTACGGGATTCAAACTACCAGTTAATCTGG_TGTGCCAGCAGCTTCTCAACCTGCAGTGCAAATTATGGTTATACTTTT",
    "TGTGCTGTAAGTGATTATGGAGGTAGCCAAGGAAATCTCATCTTT_TGTGCCAGCAGCATAAGGAGCAGCTATGAGCAGTACTTC",
    "TGTATTGTGAGGGTGGGTAACACCGGAAAACTGATATTT_TGTGCCAGCAGCTTGGAGGAAACTCAGTACTTC",
    "TGTGCAGCAAGCAGGGGCTCAACCCTGGGAAGGCTGTATTTT_TGTGCCAGCAGACCGGGACTGGCGGGAGACAACGAGCAGTTCTTC",
    "TGTGCAGCAAGTGCTGATTATAAGTTGAGCTTT_TGTGCCAGCGCTCGGGATCCCGGACAAGGAGCCTCATATGAGCAGTACTTC",
    "TGTGCCACTGATGCAGGTGGAACTAGCTATGGAAAGCTGACATTT_TGTGCCAGCAGTTTGGCAGGGGGCACTGACACCCAGTACTTT",
    "TGTGCAGTAAGAGATAACTATGGTCAGAATTTTGTCTTT_TGTGCCAGCAGCGTTGCAGGGGGCAGAGAGCAGTACTTC"
  ),
  CTstrict = c(
    "TRAV1-2.TRAJ33.TRAC_CAVMDSNYQLIW_TRBV6-1.TRBJ2-1.TRBC2_CASSETSGSTDTQYF",
    "TRAV12-1.TRAJ39.TRAC_CVVNMGGKLIF_TRBV28.TRBJ1-1.TRBC1_CASSQETQYF",
    "TRAV38-2.TRAJ28.TRAC_CAFMKTSYDKVIF_TRBV5-1.TRBJ2-7.TRBC2_CASSLGQAYEQYF",
    "TRAV21.TRAJ12.TRAC_CAVRDSNYQLIW_TRBV7-2.TRBJ2-3.TRBC2_CASSFSTCSANYGYTF",
    "TRAV8-6.TRAJ45.TRAC_CAVSDYGGSQGNLIF_TRBV19.TRBJ1-2.TRBC1_CASSIRSSYEQYF",
    "TRAV26-1.TRAJ37.TRAC_CIVRVGNTGKLIF_TRBV12-3.TRBJ2-5.TRBC2_CASSLEETQYF",
    "TRAV13-1.TRAJ44.TRAC_CAASRGSTLGRLYF_TRBV4-1.TRBJ1-5.TRBC1_CASRPGLAGDNEQFF",
    "TRAV29.TRAJ20.TRAC_CAASADYKLSF_TRBV20-1.TRBJ1-6.TRBC1_CSARDPGQGASYEQYF",
    "TRAV17.TRAJ31.TRAC_CATDAGGTSYGKLTF_TRBV11-2.TRBJ2-2.TRBC2_CASSLAGGTDTQYF",
    "TRAV3.TRAJ48.TRAC_CAVRDNYGQNFVF_TRBV9.TRBJ2-4.TRBC2_CASSVAGGREQYF"
  ),
  stringsAsFactors = FALSE
)

# Assign clonotypes to cells with a realistic frequency distribution
# (a few dominant clonotypes, many rare ones -- power-law-like)
weights <- c(30, 25, 20, 15, 10, 8, 6, 4, 3, 2)
assignments <- sample(1:10, n_cells, replace = TRUE, prob = weights)

# Add TCR metadata columns to the Seurat object
obj$clonotype_id <- factor(
  clonotype_pool$clonotype_id[assignments],
  levels = clonotype_pool$clonotype_id
)
obj$CTgene    <- clonotype_pool$CTgene[assignments]
obj$CTnt      <- clonotype_pool$CTnt[assignments]
obj$CTaa      <- clonotype_pool$CTaa[assignments]
obj$CTstrict  <- clonotype_pool$CTstrict[assignments]

# Compute frequency and proportion per clonotype
freq_table <- table(assignments)
obj$frequency  <- as.integer(freq_table[as.character(assignments)])
obj$proportion <- as.numeric(obj$frequency / n_cells)
```

Let's inspect the TCR metadata we just added:

```{r inspect_tcr}
tcr_cols <- c("clonotype_id", "CTgene", "CTnt", "CTaa",
              "CTstrict", "frequency", "proportion")
head(obj@meta.data[, tcr_cols], 8)
```

```{r clonotype_distribution}
table(obj$clonotype_id)
```

## Writing to h5ad and reading back

The core test: write the TCR-annotated Seurat object to h5ad format and verify that all immune repertoire columns survive the roundtrip.

```{r write_h5ad}
writeH5AD(obj, "tcr_annotated.h5ad")
```

```{r read_h5ad}
obj_h5ad <- readH5AD("tcr_annotated.h5ad")
```

### Verify TCR metadata preservation

```{r verify_h5ad}
# Check all TCR columns are present
tcr_cols_loaded <- intersect(tcr_cols, colnames(obj_h5ad@meta.data))
cat("TCR columns found:", paste(tcr_cols_loaded, collapse = ", "), "\n")
cat("All TCR columns preserved:",
    all(tcr_cols %in% colnames(obj_h5ad@meta.data)), "\n")
```

```{r verify_values}
# Exact match of clonotype assignments
cat("clonotype_id match:",
    all(as.character(obj$clonotype_id) == as.character(obj_h5ad$clonotype_id)),
    "\n")

# Exact match of amino acid sequences
cat("CTaa match:", all(obj$CTaa == obj_h5ad$CTaa), "\n")

# Exact match of gene annotations
cat("CTgene match:", all(obj$CTgene == obj_h5ad$CTgene), "\n")

# Numeric columns
cat("frequency match:", all(obj$frequency == obj_h5ad$frequency), "\n")
cat("proportion match:",
    all(abs(obj$proportion - obj_h5ad$proportion) < 1e-10), "\n")
```

### What gets preserved

When converting a Seurat object with TCR/BCR annotations to h5ad:

| Seurat Location | h5ad Location | Example Columns |
|-----------------|---------------|-----------------|
| `meta.data` | `obs` | `clonotype_id`, `CTgene`, `CTnt`, `CTaa`, `CTstrict` |
| `meta.data` | `obs` | `frequency`, `proportion` |
| `meta.data` | `obs` | Factor columns encoded as HDF5 categoricals |
| Reductions | `obsm` | PCA, UMAP embeddings |
| Graphs | `obsp` | SNN/KNN neighbor graphs |

All string and numeric columns in `meta.data` are written to the h5ad `obs` dataframe. Factor columns are encoded as HDF5 categorical (enum) types, which scirpy and scanpy read natively.

## Python interop: scanpy reads TCR metadata

```{r python_scanpy, eval = has_scanpy}
library(reticulate)
```

```{python scanpy_read, eval = has_scanpy}
import scanpy as sc

adata = sc.read_h5ad("tcr_annotated.h5ad")
print(f"Shape: {adata.shape}")
print(f"\nobs columns: {list(adata.obs.columns)}")

# Access TCR metadata
print(f"\nClonotype distribution:")
print(adata.obs["clonotype_id"].value_counts())

print(f"\nFirst 5 CDR3 amino acid sequences:")
print(adata.obs["CTaa"].head())
```

## Multi-format preservation

scConvert can write to many formats. Here we verify TCR metadata preservation across them.

### h5Seurat

```{r h5seurat}
scConvert(obj, dest = "tcr_annotated.h5seurat")
obj_h5s <- readH5Seurat("tcr_annotated.h5seurat")

cat("h5Seurat -- clonotype_id preserved:",
    all(as.character(obj$clonotype_id) == as.character(obj_h5s$clonotype_id)),
    "\n")
cat("h5Seurat -- CTaa preserved:",
    all(obj$CTaa == obj_h5s$CTaa), "\n")
```

### h5mu

```{r h5mu}
# h5mu is designed for multi-modal data; for single-assay objects
# with TCR metadata, h5ad or h5seurat is more appropriate.
# Here we demonstrate the conversion when the object has multiple assays:
if (length(Assays(obj)) > 1) {
  writeH5MU(obj, "tcr_annotated.h5mu")
  obj_h5mu <- readH5MU("tcr_annotated.h5mu")
  cat("h5mu -- clonotype_id preserved:",
      all(as.character(obj$clonotype_id) == as.character(obj_h5mu$clonotype_id)),
      "\n")
} else {
  cat("h5mu -- Skipped (single-assay object; use h5ad instead)\n")
}
```

### Loom

```{r loom}
writeLoom(obj, "tcr_annotated.loom", overwrite = TRUE)
obj_loom <- readLoom("tcr_annotated.loom")

# Loom stores all metadata as column attributes
loom_tcr <- intersect(tcr_cols, colnames(obj_loom@meta.data))
cat("Loom -- TCR columns found:", paste(loom_tcr, collapse = ", "), "\n")
cat("Loom -- clonotype_id preserved:",
    all(as.character(obj$clonotype_id) == as.character(obj_loom$clonotype_id)),
    "\n")
```

### Summary of format support

| Format | TCR metadata preserved | Factor encoding | Notes |
|--------|----------------------|-----------------|-------|
| h5ad | Yes | HDF5 categorical | Native scanpy/scirpy support |
| h5Seurat | Yes | HDF5 levels/values | Native Seurat loading |
| h5mu | Yes | HDF5 categorical | Multi-assay support |
| Loom | Yes | String attributes | Column attributes in col_attrs |
| Zarr | Yes | Zarr categorical | AnnData-compatible |

## Workflow: scirpy h5ad to Seurat

If your immune repertoire analysis starts in Python with scirpy:

```{r scirpy_to_seurat, eval = FALSE}
# Load h5ad with TCR/BCR annotations from scirpy
obj <- readH5AD("scirpy_analyzed.h5ad")

# TCR metadata is in meta.data
head(obj[["IR_VJ_1_junction_aa"]])   # alpha chain CDR3
head(obj[["IR_VDJ_1_junction_aa"]])  # beta chain CDR3
head(obj[["clone_id"]])
head(obj[["receptor_type"]])         # TCR, BCR, etc.

# Continue analysis in Seurat
DimPlot(obj, group.by = "clone_id")
```

### scirpy column naming

scirpy stores TCR/BCR data using a standardized naming convention in `obs`:

| scirpy Column | Description |
|---------------|-------------|
| `IR_VJ_1_junction_aa` | Alpha/light chain CDR3 amino acid sequence |
| `IR_VDJ_1_junction_aa` | Beta/heavy chain CDR3 amino acid sequence |
| `IR_VJ_1_v_call` | V gene usage (alpha/light) |
| `IR_VDJ_1_v_call` | V gene usage (beta/heavy) |
| `clone_id` | Clonotype identifier |
| `receptor_type` | TCR, BCR, or multichain |
| `chain_pairing` | Single pair, extra VJ, etc. |

All these columns are loaded as `meta.data` columns by `readH5AD()`.

## BCR (B-cell receptor) data

BCR data follows the same patterns as TCR, with heavy (IGH) and light (IGK/IGL) chains. In a real workflow, you would use scRepertoire:

```{r real_bcr_workflow, eval = FALSE}
# Real workflow (requires Cell Ranger VDJ output):
library(scRepertoire)
bcr <- combineBCR(
  read.csv("filtered_contig_annotations.csv"),
  samples = "Sample1"
)
obj <- combineExpression(bcr, obj, cloneCall = "aa", group.by = "sample")
```

Here we add synthetic BCR metadata to demonstrate preservation:

```{r synthetic_bcr}
# Add BCR-specific columns
obj$IGH_V  <- sample(
  c("IGHV1-2", "IGHV3-23", "IGHV4-34", "IGHV1-69", "IGHV3-30"),
  n_cells, replace = TRUE
)
obj$IGH_D  <- sample(
  c("IGHD3-10", "IGHD2-2", "IGHD6-19", "IGHD1-26", "IGHD4-17"),
  n_cells, replace = TRUE
)
obj$IGH_J  <- sample(
  c("IGHJ4", "IGHJ6", "IGHJ5", "IGHJ3", "IGHJ1"),
  n_cells, replace = TRUE
)
obj$IGK_V  <- sample(
  c("IGKV1-39", "IGKV3-20", "IGKV1-5", "IGKV3-11", NA),
  n_cells, replace = TRUE
)
obj$IGK_J  <- sample(
  c("IGKJ1", "IGKJ2", "IGKJ4", NA, NA),
  n_cells, replace = TRUE
)
obj$IGL_V  <- sample(
  c(NA, NA, "IGLV2-14", "IGLV1-44", "IGLV3-1"),
  n_cells, replace = TRUE
)
obj$isotype <- factor(
  sample(c("IgM", "IgG1", "IgG2", "IgA1", "IgA2"),
         n_cells, replace = TRUE, prob = c(40, 25, 15, 12, 8)),
  levels = c("IgM", "IgG1", "IgG2", "IgA1", "IgA2")
)
obj$SHM_count <- rpois(n_cells, lambda = 8)
```

```{r bcr_roundtrip}
bcr_cols <- c("IGH_V", "IGH_D", "IGH_J", "IGK_V", "IGK_J",
              "IGL_V", "isotype", "SHM_count")

writeH5AD(obj, "bcr_annotated.h5ad")
obj_bcr <- readH5AD("bcr_annotated.h5ad")

# Verify BCR columns preserved
cat("BCR columns found:",
    paste(intersect(bcr_cols, colnames(obj_bcr@meta.data)), collapse = ", "),
    "\n")
cat("IGH_V match:",
    all(obj$IGH_V == obj_bcr$IGH_V, na.rm = TRUE), "\n")
cat("isotype match:",
    all(as.character(obj$isotype) == as.character(obj_bcr$isotype)), "\n")
cat("SHM_count match:",
    all(obj$SHM_count == obj_bcr$SHM_count), "\n")

# NA handling -- NAs in light chain genes should survive roundtrip
cat("IGK_V NA count (original):", sum(is.na(obj$IGK_V)), "\n")
cat("IGK_V NA count (loaded):", sum(is.na(obj_bcr$IGK_V)), "\n")
```

### BCR-specific columns

| Column | Description |
|--------|-------------|
| `IGH_V`, `IGH_D`, `IGH_J` | Heavy chain V/D/J gene segments |
| `IGK_V`, `IGK_J` | Kappa light chain gene segments |
| `IGL_V` | Lambda light chain gene segments |
| `isotype` | IgM, IgG1, IgG2, IgA1, IgA2, etc. |
| `SHM_count` | Somatic hypermutation count |

## Multimodal: GEX + VDJ + ADT via h5mu

For experiments combining gene expression, VDJ, and surface protein (CITE-seq), h5mu captures all modalities:

```{r multimodal_vdj, eval = FALSE}
# With a multimodal Seurat object (RNA + ADT assays + TCR in meta.data):
writeH5MU(obj, "gex_adt_tcr.h5mu")
obj_loaded <- readH5MU("gex_adt_tcr.h5mu")

# Both assays + TCR metadata preserved
Assays(obj_loaded)                       # RNA, ADT
head(obj_loaded[["clonotype_id"]])       # TCR annotations in meta.data
```

### MuData and scirpy

In the muon/scirpy ecosystem, TCR data can be stored as a separate modality in h5mu:

| h5mu Location | Content |
|---------------|---------|
| `/mod/rna/` | Gene expression (AnnData) |
| `/mod/prot/` | Surface protein (AnnData) |
| `/mod/airr/` | TCR/BCR receptor data (AnnData) |
| `/obs` | Shared cell metadata |

scConvert's `readH5MU()` reads all modalities as separate Seurat assays. The `airr` modality becomes an assay containing receptor gene usage counts or binary chain presence matrices, depending on how the data was encoded.

## Limitations and considerations

1. **VDJ contigs are not stored in h5ad/h5mu**: Raw contig sequences and alignment information (BAM-level data) are not part of the h5ad/h5mu specification. Only summarized annotations (CDR3 sequences, gene calls, clonotype IDs) in cell metadata are preserved.

2. **AIRR format**: The [AIRR Community standard](https://docs.airr-community.org/) defines a tabular format for adaptive immune receptor repertoires. scConvert preserves AIRR-standard columns when they appear in `obs` metadata, but does not perform AIRR-specific validation.

3. **Clonotype graphs**: Some tools (scirpy, Dandelion) compute clonotype similarity networks stored in `obsp`. scConvert preserves these as Seurat `Graph` objects during h5ad loading.

4. **Large VDJ metadata**: TCR/BCR annotations can add 20+ columns per cell. All columns are preserved during conversion as categorical (factor) or character types.

## Data mapping summary

| Source | Destination | TCR/BCR Data Path |
|--------|-------------|-------------------|
| Seurat to h5ad | `meta.data` to `obs` | All clonotype columns preserved |
| h5ad to Seurat | `obs` to `meta.data` | All columns loaded, categoricals to factors |
| Seurat to h5mu | `meta.data` to `/obs` + per-modality | Shared metadata in global obs |
| h5mu to Seurat | `/obs` + `/mod/*/obs` to `meta.data` | All modality metadata merged |

## See Also

- [Multimodal: h5mu Format](multimodal-h5mu.html) -- storing GEX + VDJ + ADT in a single h5mu file

## Session Info

```{r sessionInfo}
sessionInfo()
```

```{r final_cleanup, echo=FALSE, results='hide', purl = FALSE}
outs <- c(
  "tcr_annotated.h5ad", "tcr_annotated.h5seurat", "tcr_annotated.h5mu",
  "tcr_annotated.loom", "bcr_annotated.h5ad"
)
existing_files <- outs[file.exists(outs)]
if (length(existing_files) > 0) file.remove(existing_files)
```
