---
title: "Immune Repertoire Metadata Preservation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Immune Repertoire Metadata Preservation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

## Introduction

Single-cell immune profiling experiments often pair gene expression with
TCR/BCR sequencing. Tools like scRepertoire add clonotype annotations --
clonotype IDs, V/D/J gene usage, CDR3 sequences, and clone frequencies -- as
cell-level metadata. scConvert preserves all of this metadata through format
conversions, so TCR/BCR annotations survive roundtrips between Seurat and h5ad.

## Load PBMC data and add synthetic TCR metadata

We start with the shipped PBMC demo (500 cells, 9 cell types) and add
synthetic TCR annotations to demonstrate metadata preservation.

```{r load-data, eval = TRUE}
pbmc_path <- system.file("extdata", "pbmc_demo.rds", package = "scConvert")
obj <- readRDS(pbmc_path)

cat("Cells:", ncol(obj), "\n")
cat("Genes:", nrow(obj), "\n")
cat("Cell types:", paste(levels(obj$seurat_annotations), collapse = ", "), "\n")
```

### Add TCR annotations

In a real workflow, these columns would come from scRepertoire's
`combineExpression()`. Here we create synthetic clonotype data to illustrate
the conversion.

```{r add-tcr, eval = TRUE}
set.seed(42)
n_cells <- ncol(obj)

# 8 clonotypes with power-law-like frequencies
clonotype_ids <- paste0("clonotype_", sprintf("%03d", 1:8))
weights <- c(30, 22, 16, 12, 8, 5, 4, 3)
assignments <- sample(seq_along(clonotype_ids), n_cells,
                       replace = TRUE, prob = weights)

obj$clonotype_id <- factor(clonotype_ids[assignments], levels = clonotype_ids)

# V/J gene usage
tra_genes <- c("TRAV1-2", "TRAV12-1", "TRAV38-2", "TRAV21",
               "TRAV8-6", "TRAV26-1", "TRAV13-1", "TRAV29")
trb_genes <- c("TRBV6-1", "TRBV28", "TRBV5-1", "TRBV7-2",
               "TRBV19", "TRBV12-3", "TRBV4-1", "TRBV20-1")
obj$tra_gene <- tra_genes[assignments]
obj$trb_gene <- trb_genes[assignments]

# Clone frequency
freq_tbl <- table(assignments)
obj$clone_frequency <- as.integer(freq_tbl[as.character(assignments)])

cat("TCR columns added:", paste(c("clonotype_id", "tra_gene",
    "trb_gene", "clone_frequency"), collapse = ", "), "\n")
```

### Visualize clonotypes

```{r plot-clonotypes, eval = TRUE, fig.width = 7, fig.height = 5}
DimPlot(obj, group.by = "clonotype_id", reduction = "umap",
        label = FALSE) +
  ggplot2::ggtitle("Clonotype distribution on UMAP")
```

## Convert to h5ad and read back

```{r roundtrip, eval = TRUE}
h5ad_file <- tempfile(fileext = ".h5ad")
writeH5AD(obj, h5ad_file, overwrite = TRUE)
obj_rt <- readH5AD(h5ad_file, verbose = FALSE)
```

### Verify TCR metadata preservation

```{r verify, eval = TRUE}
tcr_cols <- c("clonotype_id", "tra_gene", "trb_gene", "clone_frequency")

cat("All TCR columns present:",
    all(tcr_cols %in% colnames(obj_rt[[]])), "\n")
cat("clonotype_id match:",
    all(as.character(obj$clonotype_id) ==
        as.character(obj_rt$clonotype_id)), "\n")
cat("tra_gene match:", all(obj$tra_gene == obj_rt$tra_gene), "\n")
cat("trb_gene match:", all(obj$trb_gene == obj_rt$trb_gene), "\n")
cat("clone_frequency match:",
    all(obj$clone_frequency == obj_rt$clone_frequency), "\n")
```

All TCR metadata columns -- factors, strings, and integers -- are preserved
exactly through the h5ad roundtrip.

### Clonotype frequency barplot

```{r barplot, eval = TRUE, fig.width = 7, fig.height = 4}
freq_df <- as.data.frame(table(obj_rt$clonotype_id))
colnames(freq_df) <- c("Clonotype", "Cells")
freq_df <- freq_df[order(-freq_df$Cells), ]

barplot(freq_df$Cells, names.arg = freq_df$Clonotype,
        col = "steelblue", las = 2, cex.names = 0.8,
        main = "Clonotype frequencies (after roundtrip)",
        ylab = "Number of cells")
```

### Gene expression across cell types

Cell type annotations and expression values are also preserved alongside the
TCR metadata.

```{r vlnplot, eval = TRUE, fig.width = 7, fig.height = 5}
VlnPlot(obj_rt, features = "LYZ", group.by = "seurat_annotations",
        pt.size = 0.1) +
  ggplot2::ggtitle("LYZ expression by cell type (after roundtrip)") +
  ggplot2::theme(legend.position = "none")
```

## Python interop

```{python python-tcr, eval = FALSE}
# Requires Python with scanpy installed
import scanpy as sc

adata = sc.read_h5ad("tcr_pbmc.h5ad")
print(adata.obs[["clonotype_id", "tra_gene", "trb_gene"]].head(10))

# Clonotype distribution
adata.obs["clonotype_id"].value_counts().plot.bar()
```

## Cleanup

```{r cleanup, eval = TRUE, echo = FALSE, results = "hide"}
unlink(h5ad_file)
```

## Session Info

```{r session-info}
sessionInfo()
```
