---
title: "Immune Repertoire (TCR/BCR) Data with scConvert"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Immune Repertoire (TCR/BCR) Data with scConvert}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  out.width = "100%",
  dev = "png",
  purl = FALSE
)
```

## Introduction

Single-cell immune repertoire profiling (TCR/BCR sequencing) is often paired with gene expression (GEX) in multimodal experiments such as 10x Genomics 5' GEX + VDJ. This vignette explains how scConvert handles immune repertoire data during format conversion.

### Key concepts

- **TCR/BCR clonotype data** is stored as cell-level metadata (e.g., `clonotype_id`, `TRA_cdr3`, `TRB_cdr3`, `IGH_cdr3`)
- In scanpy/scirpy workflows, TCR/BCR annotations live in `obs` (cell metadata) and `obsm` (alignment matrices)
- In Seurat, clonotype data lives in `meta.data` after tools like [scRepertoire](https://github.com/ncborcherding/scRepertoire) or [Immcantation](https://immcantation.readthedocs.io/)
- scConvert preserves all cell metadata during conversion, so TCR/BCR annotations survive roundtrip

```{r packages, purl = FALSE}
library(Seurat)
library(scConvert)
```

## Workflow: Seurat + scRepertoire to h5ad

A typical immune repertoire workflow starts with 10x Cell Ranger output (both GEX and VDJ), then uses scRepertoire to add clonotype annotations:

```{r screpertoire_workflow, eval = FALSE, purl = FALSE}
# Step 1: Load GEX data into Seurat
gex <- Read10X_h5("filtered_feature_bc_matrix.h5")
obj <- CreateSeuratObject(counts = gex, project = "TCR_study")

# Step 2: Add TCR clonotype annotations via scRepertoire
# (scRepertoire reads the Cell Ranger VDJ output)
library(scRepertoire)
tcr <- combineTCR(
  read.csv("filtered_contig_annotations.csv"),
  samples = "Sample1"
)
obj <- combineExpression(tcr, obj,
  cloneCall = "aa",
  group.by = "sample"
)

# Step 3: Standard Seurat processing
obj <- NormalizeData(obj)
obj <- FindVariableFeatures(obj)
obj <- ScaleData(obj)
obj <- RunPCA(obj)
obj <- FindNeighbors(obj, dims = 1:30)
obj <- FindClusters(obj)
obj <- RunUMAP(obj, dims = 1:30)

# Step 4: Export to h5ad for scirpy analysis in Python
scConvert(obj, dest = "tcr_annotated.h5ad")
# All TCR metadata columns (clonotype_id, CDR3 sequences, frequency, etc.)
# are preserved in obs of the h5ad file
```

### What gets preserved

When converting a Seurat object with TCR/BCR annotations to h5ad:

| Seurat Location | h5ad Location | Example Columns |
|-----------------|---------------|-----------------|
| `meta.data` | `obs` | `clonotype_id`, `CTgene`, `CTnt`, `CTaa`, `CTstrict` |
| `meta.data` | `obs` | `Frequency`, `cloneType`, `Proportion` |
| `meta.data` | `obs` | `TCR_alpha_cdr3`, `TCR_beta_cdr3` |
| Reductions | `obsm` | PCA, UMAP embeddings |
| Graphs | `obsp` | SNN/KNN neighbor graphs |

All string and numeric columns in `meta.data` are written to the h5ad `obs` dataframe. Factor columns are encoded as HDF5 categorical (enum) types, which scirpy and scanpy read natively.

## Workflow: scirpy h5ad to Seurat

If your immune repertoire analysis starts in Python with scirpy:

```{r scirpy_to_seurat, eval = FALSE, purl = FALSE}
# Load h5ad with TCR/BCR annotations from scirpy
obj <- readH5AD("scirpy_analyzed.h5ad")

# TCR metadata is in meta.data
head(obj[["IR_VJ_1_junction_aa"]])  # alpha chain CDR3
head(obj[["IR_VDJ_1_junction_aa"]])  # beta chain CDR3
head(obj[["clone_id"]])
head(obj[["receptor_type"]])  # TCR, BCR, etc.

# Continue analysis in Seurat
DimPlot(obj, group.by = "clone_id")
```

### scirpy column naming

scirpy stores TCR/BCR data using a standardized naming convention in `obs`:

| scirpy Column | Description |
|---------------|-------------|
| `IR_VJ_1_junction_aa` | Alpha/light chain CDR3 amino acid sequence |
| `IR_VDJ_1_junction_aa` | Beta/heavy chain CDR3 amino acid sequence |
| `IR_VJ_1_v_call` | V gene usage (alpha/light) |
| `IR_VDJ_1_v_call` | V gene usage (beta/heavy) |
| `clone_id` | Clonotype identifier |
| `receptor_type` | TCR, BCR, or multichain |
| `chain_pairing` | Single pair, extra VJ, etc. |

All these columns are loaded as `meta.data` columns by `readH5AD()`.

## Multimodal: GEX + VDJ + ADT via h5mu

For experiments combining gene expression, VDJ, and surface protein (CITE-seq):

```{r multimodal_vdj, eval = FALSE, purl = FALSE}
# Create a multimodal Seurat object
# RNA assay: gene expression
# ADT assay: surface protein (CITE-seq)
# TCR annotations: in meta.data

# Save as h5mu (preserves both assays)
writeH5MU(obj, "gex_adt_tcr.h5mu")

# Load back
obj_loaded <- readH5MU("gex_adt_tcr.h5mu")

# Both assays + TCR metadata preserved
Assays(obj_loaded)  # RNA, ADT
head(obj_loaded[["clonotype_id"]])  # TCR annotations in meta.data
```

### MuData and scirpy

In the muon/scirpy ecosystem, TCR data can be stored as a separate modality in h5mu:

| h5mu Location | Content |
|---------------|---------|
| `/mod/rna/` | Gene expression (AnnData) |
| `/mod/prot/` | Surface protein (AnnData) |
| `/mod/airr/` | TCR/BCR receptor data (AnnData) |
| `/obs` | Shared cell metadata |

scConvert's `readH5MU()` reads all modalities as separate Seurat assays. The `airr` modality becomes an assay containing receptor gene usage counts or binary chain presence matrices, depending on how the data was encoded.

## BCR (B-cell receptor) data

BCR data follows the same patterns as TCR, with heavy (IGH) and light (IGK/IGL) chains:

```{r bcr_example, eval = FALSE, purl = FALSE}
# BCR annotations from scRepertoire
library(scRepertoire)
bcr <- combineBCR(
  read.csv("filtered_contig_annotations.csv"),
  samples = "Sample1"
)
obj <- combineExpression(bcr, obj,
  cloneCall = "aa",
  group.by = "sample"
)

# Export preserving BCR metadata
scConvert(obj, dest = "bcr_study.h5ad")
```

### BCR-specific columns

| Column | Description |
|--------|-------------|
| `IGH_cdr3` | Heavy chain CDR3 |
| `IGK_cdr3` / `IGL_cdr3` | Light chain CDR3 |
| `isotype` | IgM, IgG, IgA, etc. |
| `SHM_count` | Somatic hypermutation count |

## Limitations and considerations

1. **VDJ contigs are not stored in h5ad/h5mu**: Raw contig sequences and alignment information (BAM-level data) are not part of the h5ad/h5mu specification. Only summarized annotations (CDR3 sequences, gene calls, clonotype IDs) in cell metadata are preserved.

2. **AIRR format**: The [AIRR Community standard](https://docs.airr-community.org/) defines a tabular format for adaptive immune receptor repertoires. scConvert preserves AIRR-standard columns when they appear in `obs` metadata, but does not perform AIRR-specific validation.

3. **Clonotype graphs**: Some tools (scirpy, Dandelion) compute clonotype similarity networks stored in `obsp`. scConvert preserves these as Seurat `Graph` objects during h5ad loading.

4. **Large VDJ metadata**: TCR/BCR annotations can add 20+ columns per cell. All columns are preserved during conversion as categorical (factor) or character types.

## Data mapping summary

| Source | Destination | TCR/BCR Data Path |
|--------|-------------|-------------------|
| Seurat → h5ad | `meta.data` → `obs` | All clonotype columns preserved |
| h5ad → Seurat | `obs` → `meta.data` | All columns loaded, categoricals → factors |
| Seurat → h5mu | `meta.data` → `/obs` + per-modality | Shared metadata in global obs |
| h5mu → Seurat | `/obs` + `/mod/*/obs` → `meta.data` | All modality metadata merged |

## See Also

- [Multimodal: h5mu Format](multimodal-h5mu.html) -- storing GEX + VDJ + ADT in a single h5mu file

## Session Info

```{r sessionInfo, purl = FALSE}
sessionInfo()
```
