---
title: "Multimodal: CITE-seq and Multiome"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multimodal: CITE-seq and Multiome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  dev = "png",
  fig.width = 8,
  fig.height = 6
)
has_SeuratData <- requireNamespace("SeuratData", quietly = TRUE)
has_cbmc <- FALSE
has_multiome <- FALSE
if (has_SeuratData) {
  library(SeuratData)
  has_cbmc <- tryCatch("cbmc.SeuratData" %in% rownames(InstalledData()), error = function(e) FALSE)
  has_multiome <- FALSE
}
has_muon <- FALSE
has_scanpy <- FALSE
if (requireNamespace("reticulate", quietly = TRUE)) {
  scverse_path <- "/opt/homebrew/Caskroom/miniconda/base/envs/scverse"
  if (dir.exists(scverse_path)) {
    reticulate::use_condaenv("scverse", required = FALSE)
    Sys.setenv(
      OMP_NUM_THREADS = "1", OPENBLAS_NUM_THREADS = "1",
      MKL_NUM_THREADS = "1", NUMBA_THREADING_LAYER = "tbb"
    )
    has_muon <- reticulate::py_module_available("mudata")
    has_scanpy <- reticulate::py_module_available("scanpy")
    if (has_scanpy) {
      reticulate::py_run_string("import matplotlib; matplotlib.use('Agg')")
    }
  }
}
```

## Overview

Multimodal single-cell experiments measure multiple data types from the same cells. This vignette demonstrates scConvert's handling of:

1. **CITE-seq** (RNA + surface protein / ADT) via h5mu format
2. **Multiome** (RNA + ATAC) via h5Seurat format

The **h5mu** (MuData) format is the standard for multimodal data in the scverse ecosystem, storing each modality as a separate AnnData within a single file.

```{r packages}
library(Seurat)
library(scConvert)
library(ggplot2)
```

## CITE-seq: CBMC dataset

The CBMC (cord blood mononuclear cell) dataset contains RNA and ADT (antibody-derived tag) measurements from CITE-seq:

```{r load_cbmc, eval = has_cbmc}
library(SeuratData)
data("cbmc", package = "cbmc.SeuratData")
cbmc <- UpdateSeuratObject(cbmc)

cat("Cells:", ncol(cbmc), "\n")
cat("Assays:", paste(Assays(cbmc), collapse = ", "), "\n")
cat("RNA features:", nrow(cbmc[["RNA"]]), "\n")
cat("ADT features:", nrow(cbmc[["ADT"]]), "\n")
```

### Process and cluster

```{r process_cbmc, eval = has_cbmc}
cbmc <- NormalizeData(cbmc, verbose = FALSE)
cbmc <- FindVariableFeatures(cbmc, verbose = FALSE)
cbmc <- ScaleData(cbmc, verbose = FALSE)
cbmc <- RunPCA(cbmc, verbose = FALSE)
cbmc <- RunUMAP(cbmc, dims = 1:30, verbose = FALSE)
cbmc <- FindNeighbors(cbmc, dims = 1:30, verbose = FALSE)
cbmc <- FindClusters(cbmc, resolution = 0.8, verbose = FALSE)
```

### UMAP visualization

```{r cbmc_umap, eval = has_cbmc, fig.width = 14, fig.height = 6}
library(patchwork)

p1 <- DimPlot(cbmc, label = TRUE, pt.size = 0.3) + NoLegend() + ggtitle("RNA clusters")

# Normalize ADT
cbmc <- NormalizeData(cbmc, assay = "ADT", normalization.method = "CLR", verbose = FALSE)
p2 <- FeaturePlot(cbmc, features = "adt_CD3", pt.size = 0.3) + ggtitle("CD3 protein (ADT)")

p1 + p2
```

### Export to h5mu

```{r export_h5mu, eval = has_cbmc}
# Remove overlapping features (h5mu requires unique names per modality)
overlap <- intersect(rownames(cbmc[["ADT"]]), rownames(cbmc[["RNA"]]))
if (length(overlap) > 0) {
  adt_keep <- setdiff(rownames(cbmc[["ADT"]]), overlap)
  cbmc[["ADT"]] <- subset(cbmc[["ADT"]], features = adt_keep)
  cat("Removed", length(overlap), "overlapping features from ADT\n")
}

h5mu_path <- tempfile(fileext = ".h5mu")
writeH5MU(cbmc, h5mu_path, overwrite = TRUE)
cat("h5mu file size:", round(file.size(h5mu_path) / 1024^2, 1), "MB\n")
```

### Validate h5mu in Python

The exported h5mu file is directly compatible with the Python mudata/muon ecosystem:

```{python inspect_citeseq_h5mu, eval = has_cbmc && has_muon}
import mudata as md

mdata = md.read_h5mu(r.h5mu_path)
print(mdata)
print(f"\nModalities: {list(mdata.mod.keys())}")
for name, mod in mdata.mod.items():
    print(f"  {name}: {mod.n_obs} cells x {mod.n_vars} features")
    if mod.X is not None:
        import numpy as np
        x = mod.X.toarray() if hasattr(mod.X, 'toarray') else mod.X
        print(f"    X range: [{x.min():.2f}, {x.max():.2f}]")
```

```{python scanpy_on_citeseq, eval = has_cbmc && has_muon && has_scanpy}
import scanpy as sc
import matplotlib.pyplot as plt

rna = mdata.mod["rna"].copy()
sc.pp.normalize_total(rna)
sc.pp.log1p(rna)
sc.pp.highly_variable_genes(rna, n_top_genes=2000)
sc.pp.pca(rna)
sc.pp.neighbors(rna)
sc.tl.umap(rna)
sc.tl.leiden(rna, flavor="igraph")

fig, axes = plt.subplots(1, 2, figsize=(12, 5))
sc.pl.umap(rna, color="leiden", ax=axes[0], show=False, title="RNA Leiden clusters")
sc.pl.umap(rna, color="n_counts" if "n_counts" in rna.obs else rna.obs.columns[0],
           ax=axes[1], show=False, title="RNA QC metric")
plt.tight_layout()
plt.show()
```

### Load h5mu back

```{r load_h5mu, eval = has_cbmc}
cbmc_rt <- readH5MU(h5mu_path)

cat("Cells:", ncol(cbmc_rt), "\n")
cat("Assays:", paste(Assays(cbmc_rt), collapse = ", "), "\n")
cat("RNA features:", nrow(cbmc_rt[["RNA"]]), "\n")
cat("ADT features:", nrow(cbmc_rt[["ADT"]]), "\n")
```

### Verify preservation

```{r verify_preservation, eval = has_cbmc}
cat("=== Dimensions ===\n")
cat("Original:", ncol(cbmc), "cells,", length(Assays(cbmc)), "assays\n")
cat("Loaded:  ", ncol(cbmc_rt), "cells,", length(Assays(cbmc_rt)), "assays\n")
cat("Assay match:", identical(sort(Assays(cbmc)), sort(Assays(cbmc_rt))), "\n")

common_c <- intersect(colnames(cbmc), colnames(cbmc_rt))
cat("\n=== Barcodes ===\n")
cat("Shared:", length(common_c), "/", ncol(cbmc), "\n")

cat("\n=== RNA preservation ===\n")
common_g <- intersect(rownames(cbmc[["RNA"]]), rownames(cbmc_rt[["RNA"]]))
cat("Features shared:", length(common_g), "/", nrow(cbmc[["RNA"]]), "\n")
set.seed(42)
sc <- sample(common_c, min(300, length(common_c)))
sg <- sample(common_g, min(100, length(common_g)))
rna_orig <- as.numeric(GetAssayData(cbmc, assay = "RNA", layer = "counts")[sg, sc])
rna_rt <- as.numeric(GetAssayData(cbmc_rt, assay = "RNA", layer = "counts")[sg, sc])
cat("Values identical:", identical(rna_orig, rna_rt), "\n")
cat("Max abs diff:", max(abs(rna_orig - rna_rt)), "\n")

cat("\n=== ADT preservation ===\n")
common_adt <- intersect(rownames(cbmc[["ADT"]]), rownames(cbmc_rt[["ADT"]]))
cat("Features shared:", length(common_adt), "/", nrow(cbmc[["ADT"]]), "\n")
if (length(common_adt) > 0) {
  adt_orig <- as.numeric(GetAssayData(cbmc, assay = "ADT", layer = "counts")[common_adt, sc])
  adt_rt <- as.numeric(GetAssayData(cbmc_rt, assay = "ADT", layer = "counts")[common_adt, sc])
  cat("Values identical:", identical(adt_orig, adt_rt), "\n")
  cat("Max abs diff:", max(abs(adt_orig - adt_rt)), "\n")
}

cat("\n=== Metadata ===\n")
shared_meta <- intersect(colnames(cbmc[[]]), colnames(cbmc_rt[[]]))
cat("Columns preserved:", length(shared_meta), "/", ncol(cbmc[[]]), "\n")
```

```{r cbmc_cleanup, echo = FALSE, results = 'hide', eval = has_cbmc}
unlink(h5mu_path)
```

## Multiome: RNA + ATAC

10x Genomics Multiome measures RNA and chromatin accessibility (ATAC) from the same cells. scConvert supports roundtripping both modalities via h5mu.

```{r load_multiome, eval = has_multiome}
library(SeuratData)

# Load both RNA and ATAC from PBMC Multiome
pbmc.rna <- LoadData("pbmcMultiome", "pbmc.rna")
pbmc.atac <- LoadData("pbmcMultiome", "pbmc.atac")
pbmc.rna <- UpdateSeuratObject(pbmc.rna)
pbmc.atac <- UpdateSeuratObject(pbmc.atac)

cat("RNA cells:", ncol(pbmc.rna), "\n")
cat("RNA features:", nrow(pbmc.rna), "\n")
cat("ATAC cells:", ncol(pbmc.atac), "\n")
cat("ATAC features (peaks):", nrow(pbmc.atac), "\n")
```

### Combine RNA + ATAC into a single object

```{r combine_multiome, eval = has_multiome}
# Find shared cells
shared_cells <- intersect(colnames(pbmc.rna), colnames(pbmc.atac))
cat("Shared cells:", length(shared_cells), "\n")

# Create combined object
combined <- CreateSeuratObject(
  counts = GetAssayData(pbmc.rna, layer = "counts")[, shared_cells],
  assay = "RNA"
)
combined[["ATAC"]] <- CreateAssay5Object(
  counts = GetAssayData(pbmc.atac, layer = "counts")[, shared_cells]
)

cat("Combined object:\n")
cat("  Cells:", ncol(combined), "\n")
cat("  Assays:", paste(Assays(combined), collapse = ", "), "\n")
cat("  RNA features:", nrow(combined[["RNA"]]), "\n")
cat("  ATAC features:", nrow(combined[["ATAC"]]), "\n")
```

### Roundtrip via h5mu

The h5mu format stores each assay as a separate modality, making it ideal for multimodal data:

```{r multiome_h5mu, eval = has_multiome}
h5mu_path <- tempfile(fileext = ".h5mu")
writeH5MU(combined, h5mu_path, overwrite = TRUE)
cat("h5mu file size:", round(file.size(h5mu_path) / 1024^2, 1), "MB\n")

# Load back
combined_rt <- readH5MU(h5mu_path)
cat("Roundtrip assays:", paste(Assays(combined_rt), collapse = ", "), "\n")
cat("Roundtrip RNA features:", nrow(combined_rt[["RNA"]]), "\n")
cat("Roundtrip ATAC features:", nrow(combined_rt[["ATAC"]]), "\n")
```

### Verify preservation of both modalities

```{r multiome_verify, eval = has_multiome}
common_c <- intersect(colnames(combined), colnames(combined_rt))
cat("Cells preserved:", length(common_c), "/", ncol(combined), "\n")
set.seed(42)
sc <- sample(common_c, min(200, length(common_c)))

# RNA
rna_genes <- intersect(rownames(combined[["RNA"]]), rownames(combined_rt[["RNA"]]))
sg <- sample(rna_genes, min(100, length(rna_genes)))
rna_orig <- as.numeric(GetAssayData(combined, assay = "RNA", layer = "counts")[sg, sc])
rna_rt <- as.numeric(GetAssayData(combined_rt, assay = "RNA", layer = "counts")[sg, sc])
cat("RNA features preserved:", length(rna_genes), "/", nrow(combined[["RNA"]]), "\n")
cat("RNA values identical:", identical(rna_orig, rna_rt), "\n")

# ATAC
atac_peaks <- intersect(rownames(combined[["ATAC"]]), rownames(combined_rt[["ATAC"]]))
sp <- sample(atac_peaks, min(100, length(atac_peaks)))
atac_orig <- as.numeric(GetAssayData(combined, assay = "ATAC", layer = "counts")[sp, sc])
atac_rt <- as.numeric(GetAssayData(combined_rt, assay = "ATAC", layer = "counts")[sp, sc])
cat("ATAC features preserved:", length(atac_peaks), "/", nrow(combined[["ATAC"]]), "\n")
cat("ATAC values identical:", identical(atac_orig, atac_rt), "\n")
```

### Convert RNA component to h5ad

The RNA component can also be exported independently via h5ad:

```{r multiome_h5ad, eval = has_multiome}
h5ad_path <- tempfile(fileext = ".h5ad")
h5s_path <- tempfile(fileext = ".h5Seurat")

writeH5Seurat(pbmc.rna, h5s_path, overwrite = TRUE, verbose = FALSE)
scConvert(h5s_path, dest = "h5ad", overwrite = TRUE, verbose = FALSE)
h5ad_path <- sub("\\.h5Seurat$", ".h5ad", h5s_path)

rna_rt <- readH5AD(h5ad_path, verbose = FALSE)
cat("h5ad roundtrip cells:", ncol(rna_rt), "/", ncol(pbmc.rna), "\n")
```

```{r multiome_cleanup, echo = FALSE, results = 'hide', eval = has_multiome}
unlink(c(h5mu_path, h5ad_path, h5s_path))
```

## scATAC-seq notes

For standalone scATAC-seq data (not part of Multiome):

- **Peak matrices** are stored as standard sparse matrices and convert normally via h5ad/h5Seurat
- **Fragment files** (`.tsv.gz`) are not part of the h5ad/h5Seurat format and must be transferred separately
- **Signac ChromatinAssay** objects: The expression matrix (peaks x cells) is preserved; Signac-specific slots (fragments, motifs, annotations) require separate handling

```r
# Convert ATAC peak matrix
atac_obj <- CreateSeuratObject(counts = peak_matrix, assay = "ATAC")
writeH5Seurat(atac_obj, "atac.h5Seurat")
scConvert("atac.h5Seurat", dest = "h5ad")

# Fragment files must be copied separately
file.copy("fragments.tsv.gz", "output_dir/fragments.tsv.gz")
file.copy("fragments.tsv.gz.tbi", "output_dir/fragments.tsv.gz.tbi")
```

## Data mapping for multimodal formats

### h5mu modality mapping

| Seurat Assay | h5mu Modality | Standard for |
|-------------|---------------|-------------|
| RNA | rna | Gene expression |
| ADT | prot | Surface proteins (CITE-seq) |
| ATAC | atac | Chromatin accessibility |
| Spatial | spatial | Spatial gene expression |
| SCT | sct | SCTransform-normalized |
| Custom | lowercase name | Any other assay |

### What is preserved in h5mu roundtrip

| Component | Preserved | Notes |
|-----------|-----------|-------|
| Expression matrices | Yes | Per-modality |
| Cell metadata | Yes | Global obs + per-modality |
| Feature metadata | Yes | Per-modality var |
| Dimensional reductions | Partial | Stored in modality obsm |
| Neighbor graphs | Partial | Stored in modality obsp |

## Session Info

```{r sessionInfo}
sessionInfo()
```
