---
title: "Multimodal Data with MuData (h5mu)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multimodal Data with MuData (h5mu)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

## Introduction

The [MuData](https://muon.readthedocs.io/en/latest/io/mudata.html) format
(`.h5mu`) stores multimodal single-cell data in a single file. Each modality
(RNA, protein, ATAC) lives as a separate AnnData object, sharing a common set
of cell barcodes. This is the native format for
[muon](https://muon-tutorials.readthedocs.io/) in Python.
scConvert reads and writes h5mu natively -- no Python or MuDataSeurat required.

| Format | Best for |
|--------|----------|
| **h5ad** | Single-modality (RNA only) |
| **h5mu** | Multi-modal (RNA + ADT, RNA + ATAC, etc.) |

## Load CITE-seq demo data

The shipped `citeseq_demo.rds` contains 500 CITE-seq cells with two assays:
RNA (2000 genes) and ADT (10 surface protein antibodies).

```{r load-data, eval = TRUE}
obj <- readRDS(system.file("extdata", "citeseq_demo.rds", package = "scConvert"))

cat("Cells:", ncol(obj), "\n")
cat("Assays:", paste(Assays(obj), collapse = ", "), "\n")
cat("RNA features:", nrow(obj[["RNA"]]), "\n")
cat("ADT features:", nrow(obj[["ADT"]]), "\n")
```

```{r dimplot-original, eval = TRUE, fig.width = 7, fig.height = 5}
DimPlot(obj, group.by = "seurat_clusters", label = TRUE, pt.size = 0.8) +
  ggtitle("CITE-seq clusters (original)")
```

## Export to h5mu

`writeH5MU()` saves every assay as a separate modality. Seurat assay names are
automatically mapped to standard MuData conventions:

| Seurat Assay | h5mu Modality |
|-------------|---------------|
| RNA | rna |
| ADT | prot |
| ATAC | atac |
| Other | lowercase name |

```{r write-h5mu, eval = TRUE}
h5mu_path <- file.path(tempdir(), "citeseq.h5mu")
writeH5MU(obj, h5mu_path, overwrite = TRUE)
cat("Wrote:", h5mu_path, "\n")
cat("File size:", round(file.size(h5mu_path) / 1024^2, 1), "MB\n")
```

## Load h5mu back into R

`readH5MU()` reverses the mapping, restoring standard Seurat assay names:

| h5mu Modality | Seurat Assay |
|---------------|-------------|
| rna | RNA |
| prot | ADT |
| atac | ATAC |

```{r read-h5mu, eval = TRUE}
loaded <- readH5MU(h5mu_path)

cat("Cells:", ncol(loaded), "\n")
cat("Assays:", paste(Assays(loaded), collapse = ", "), "\n")
cat("RNA features:", nrow(loaded[["RNA"]]), "\n")
cat("ADT features:", nrow(loaded[["ADT"]]), "\n")
```

## Compare original and roundtripped data

### Side-by-side DimPlot

```{r dimplot-compare, eval = TRUE, fig.width = 12, fig.height = 5}
library(patchwork)

p1 <- DimPlot(obj, group.by = "seurat_clusters", label = TRUE, pt.size = 0.8) +
  ggtitle("Original") + NoLegend()
p2 <- DimPlot(loaded, group.by = "seurat_clusters", label = TRUE, pt.size = 0.8) +
  ggtitle("After h5mu roundtrip") + NoLegend()
p1 + p2
```

### ADT protein expression

```{r featureplot-adt, eval = TRUE, fig.width = 7, fig.height = 5}
DefaultAssay(loaded) <- "ADT"
loaded <- NormalizeData(loaded, normalization.method = "CLR", margin = 2, verbose = FALSE)
FeaturePlot(loaded, features = "CD3", pt.size = 0.8) +
  ggtitle("CD3 protein (ADT) -- loaded from h5mu")
DefaultAssay(loaded) <- "RNA"
```

### Verify expression values

```{r verify, eval = TRUE}
cat("=== Dimensions ===\n")
cat("Cells match:", ncol(obj) == ncol(loaded), "\n")
cat("Assays match:", identical(sort(Assays(obj)), sort(Assays(loaded))), "\n")

# RNA counts
common_cells <- intersect(colnames(obj), colnames(loaded))
common_genes <- intersect(rownames(obj[["RNA"]]), rownames(loaded[["RNA"]]))
orig_rna <- as.numeric(GetAssayData(obj, assay = "RNA", layer = "counts")[
  head(common_genes, 100), head(common_cells, 100)])
rt_rna <- as.numeric(GetAssayData(loaded, assay = "RNA", layer = "counts")[
  head(common_genes, 100), head(common_cells, 100)])
cat("\n=== RNA counts ===\n")
cat("Values identical:", identical(orig_rna, rt_rna), "\n")

# ADT counts
common_adt <- intersect(rownames(obj[["ADT"]]), rownames(loaded[["ADT"]]))
orig_adt <- as.numeric(GetAssayData(obj, assay = "ADT", layer = "counts")[
  common_adt, head(common_cells, 100)])
rt_adt <- as.numeric(GetAssayData(loaded, assay = "ADT", layer = "counts")[
  common_adt, head(common_cells, 100)])
cat("\n=== ADT counts ===\n")
cat("Values identical:", identical(orig_adt, rt_adt), "\n")
```

## Custom modality name mapping

You can override the default mapping when reading back with the `assay.names`
argument:

```{r custom-mapping, eval = TRUE}
custom <- readH5MU(h5mu_path, assay.names = c(rna = "RNA", prot = "Protein"))
cat("Assays with custom mapping:", paste(Assays(custom), collapse = ", "), "\n")
```

## Python interoperability

The exported h5mu file is directly compatible with muon in Python.

```{python inspect-python, eval = FALSE}
# Requires Python: pip install mudata
import mudata as md

mdata = md.read_h5mu("citeseq.h5mu")
print(mdata)
print("Modalities:", list(mdata.mod.keys()))
for name, mod in mdata.mod.items():
    print(f"  {name}: {mod.n_obs} cells x {mod.n_vars} features")
```

## Clean up

```{r cleanup, eval = TRUE}
unlink(h5mu_path)
```

## Session Info

```{r session-info}
sessionInfo()
```
