---
title: "Multimodal Data with MuData (h5mu)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multimodal Data with MuData (h5mu)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

## Introduction

The [MuData](https://muon.readthedocs.io/en/latest/io/mudata.html) format
(`.h5mu`) stores multimodal single-cell data in a single file. Each modality
(RNA, protein, ATAC) lives as a separate AnnData object, sharing a common set
of cell barcodes. This is the native format for
[muon](https://muon-tutorials.readthedocs.io/) in Python.
scConvert reads and writes h5mu natively -- no Python or MuDataSeurat required.

| Format | Best for |
|--------|----------|
| **h5ad** | Single-modality (RNA only) |
| **h5mu** | Multi-modal (RNA + ADT, RNA + ATAC, etc.) |

## Read h5mu directly

The shipped `citeseq_demo.h5mu` contains 500 CITE-seq cells with two
modalities: RNA (2,000 genes) and ADT (10 surface protein antibodies).
`readH5MU()` loads it into a Seurat object with both assays intact.

```{r read-h5mu, eval = TRUE}
h5mu_file <- system.file("extdata", "citeseq_demo.h5mu", package = "scConvert")
obj <- readH5MU(h5mu_file)

cat("Cells:", ncol(obj), "\n")
cat("Assays:", paste(names(obj@assays), collapse = ", "), "\n")
cat("RNA features:", nrow(obj[["RNA"]]), "\n")
cat("ADT features:", nrow(obj[["ADT"]]), "\n")
```

### UMAP of RNA clusters

```{r dimplot-h5mu, eval = TRUE, fig.width = 7, fig.height = 5}
DimPlot(obj, group.by = "seurat_clusters", label = TRUE, pt.size = 0.8) +
  ggtitle("CITE-seq clusters (loaded from h5mu)")
```

### Protein expression

After reading h5mu, the ADT assay contains only raw counts. Normalize with
CLR (centered log-ratio) before plotting protein markers.

```{r featureplot-adt, eval = TRUE, fig.width = 7, fig.height = 5}
DefaultAssay(obj) <- "ADT"
obj <- NormalizeData(obj, normalization.method = "CLR", margin = 2, verbose = FALSE)
FeaturePlot(obj, features = "CD3", pt.size = 0.8) +
  ggtitle("CD3 protein (ADT) -- loaded from h5mu")
DefaultAssay(obj) <- "RNA"
```

## Write and roundtrip

Load the same dataset from its Seurat `.rds` form, write to h5mu, then read
back and compare. This demonstrates that scConvert preserves both assays
through a full write/read cycle.

### Modality name mapping

`writeH5MU()` automatically maps Seurat assay names to standard MuData
conventions:

| Seurat Assay | h5mu Modality |
|-------------|---------------|
| RNA | rna |
| ADT | prot |
| ATAC | atac |
| Other | lowercase name |

`readH5MU()` reverses the mapping when loading.

```{r write-roundtrip, eval = TRUE}
orig <- readRDS(system.file("extdata", "citeseq_demo.rds", package = "scConvert"))

h5mu_path <- file.path(tempdir(), "citeseq_roundtrip.h5mu")
writeH5MU(orig, h5mu_path, overwrite = TRUE)
cat("Wrote:", round(file.size(h5mu_path) / 1024^2, 1), "MB\n")

loaded <- readH5MU(h5mu_path)
cat("Loaded:", ncol(loaded), "cells,", paste(names(loaded@assays), collapse = ", "), "\n")
```

### Side-by-side comparison

```{r dimplot-compare, eval = TRUE, fig.width = 12, fig.height = 5}
library(patchwork)

p1 <- DimPlot(orig, group.by = "seurat_clusters", label = TRUE, pt.size = 0.8) +
  ggtitle("Original (.rds)") + NoLegend()
p2 <- DimPlot(loaded, group.by = "seurat_clusters", label = TRUE, pt.size = 0.8) +
  ggtitle("After h5mu roundtrip") + NoLegend()
p1 + p2
```

### Verify data integrity

```{r verify, eval = TRUE}
cat("Cell count match:", ncol(orig) == ncol(loaded), "\n")
cat("Assays match:", identical(sort(names(orig@assays)), sort(names(loaded@assays))), "\n")

common_cells <- intersect(colnames(orig), colnames(loaded))
common_genes <- intersect(rownames(orig[["RNA"]]), rownames(loaded[["RNA"]]))
orig_rna <- as.numeric(GetAssayData(orig, assay = "RNA", layer = "counts")[
  head(common_genes, 100), head(common_cells, 100)])
rt_rna <- as.numeric(GetAssayData(loaded, assay = "RNA", layer = "counts")[
  head(common_genes, 100), head(common_cells, 100)])
cat("RNA counts identical:", identical(orig_rna, rt_rna), "\n")

common_adt <- intersect(rownames(orig[["ADT"]]), rownames(loaded[["ADT"]]))
orig_adt <- as.numeric(GetAssayData(orig, assay = "ADT", layer = "counts")[
  common_adt, head(common_cells, 100)])
rt_adt <- as.numeric(GetAssayData(loaded, assay = "ADT", layer = "counts")[
  common_adt, head(common_cells, 100)])
cat("ADT counts identical:", identical(orig_adt, rt_adt), "\n")
```

## Custom modality name mapping

You can override the default mapping when reading with the `assay.names`
argument:

```{r custom-mapping, eval = TRUE}
custom <- readH5MU(h5mu_file, assay.names = c(rna = "RNA", prot = "Protein"))
cat("Assays with custom mapping:", paste(names(custom@assays), collapse = ", "), "\n")
```

## Python interoperability

The h5mu file is directly compatible with muon in Python.

```{python inspect-python, eval = FALSE}
# Requires Python: pip install mudata
import mudata as md

mdata = md.read_h5mu("citeseq_demo.h5mu")
print(mdata)
print("Modalities:", list(mdata.mod.keys()))
for name, mod in mdata.mod.items():
    print(f"  {name}: {mod.n_obs} cells x {mod.n_vars} features")
```

## Clean up

```{r cleanup, eval = TRUE}
unlink(h5mu_path)
```

## Session Info

```{r session-info}
sessionInfo()
```
