---
title: "Multimodal Data with MuData (h5mu)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multimodal Data with MuData (h5mu)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  out.width = "100%",
  dev = "png",
  purl = FALSE
)
can_run <- requireNamespace("SeuratData", quietly = TRUE) &&
  tryCatch({
    library(SeuratData)
    "cbmc.SeuratData" %in% rownames(InstalledData())
  }, error = function(e) FALSE)
can_load <- can_run
has_muon <- FALSE
has_scanpy <- FALSE
if (requireNamespace("reticulate", quietly = TRUE)) {
  scverse_path <- "/opt/homebrew/Caskroom/miniconda/base/envs/scverse"
  if (dir.exists(scverse_path)) {
    reticulate::use_condaenv("scverse", required = FALSE)
    Sys.setenv(
      OMP_NUM_THREADS = "1", OPENBLAS_NUM_THREADS = "1",
      MKL_NUM_THREADS = "1", NUMBA_THREADING_LAYER = "tbb"
    )
    has_muon <- reticulate::py_module_available("mudata")
    has_scanpy <- reticulate::py_module_available("scanpy")
    if (has_scanpy) {
      reticulate::py_run_string("import matplotlib; matplotlib.use('Agg')")
    }
  }
}
```

```{r cleanup, echo=FALSE, results='hide', purl = FALSE}
outs <- c("cbmc_multimodal.h5mu", "cbmc_roundtrip.h5mu")
existing_files <- outs[file.exists(outs)]
if (length(existing_files) > 0) file.remove(existing_files)
```

## Introduction

The [MuData](https://muon.readthedocs.io/en/latest/io/mudata.html) format (`.h5mu`) stores multimodal single-cell data — multiple modalities (e.g., RNA + protein) in a single file with shared cell annotations. This is the native format for [muon](https://muon-tutorials.readthedocs.io/) in Python.

scConvert provides native h5mu read/write with no external dependencies (no MuDataSeurat or Python required):

- **`writeH5MU()`**: Export a multi-assay Seurat object to h5mu, with automatic assay-to-modality name mapping. Works with Seurat V5 Assay5.
- **`readH5MU()`**: Load an h5mu file as a Seurat object, with automatic modality-to-assay name mapping

### When to use h5mu vs h5ad

| Format | Use case |
|--------|----------|
| **h5ad** | Single-modality data (RNA only) |
| **h5mu** | Multi-modal data (RNA + ADT, RNA + ATAC, etc.) |

```{r packages, eval = can_run, purl = FALSE}
library(Seurat)
library(scConvert)
library(ggplot2)
library(patchwork)
```

## Create a multimodal Seurat object

We use the `cbmc` dataset from SeuratData — a CITE-seq experiment with RNA and ADT (antibody-derived tag) assays:

```{r load_cbmc, eval = can_run, purl = FALSE}
library(SeuratData)
if (!"cbmc" %in% rownames(InstalledData())) {
  InstallData("cbmc")
}
data("cbmc", package = "cbmc.SeuratData")
cbmc <- UpdateSeuratObject(cbmc)

# Remove ADT features that overlap with RNA (h5mu requires unique feature names per modality)
overlap <- intersect(rownames(cbmc[["ADT"]]), rownames(cbmc[["RNA"]]))
if (length(overlap) > 0) {
  adt_keep <- setdiff(rownames(cbmc[["ADT"]]), overlap)
  cbmc[["ADT"]] <- subset(cbmc[["ADT"]], features = adt_keep)
  cat("Removed", length(overlap), "overlapping features from ADT\n")
}

cat("Assays:", paste(Assays(cbmc), collapse = ", "), "\n")
cat("Cells:", ncol(cbmc), "\n")
cat("RNA features:", nrow(cbmc[["RNA"]]), "\n")
cat("ADT features:", nrow(cbmc[["ADT"]]), "\n")
```

This object has two assays — RNA (gene expression) and ADT (surface protein) — measured on the same cells, making it ideal for h5mu.

> **Note**: h5mu requires unique feature names across modalities. If your CITE-seq ADT panel has features that share names with RNA genes (e.g., CD3D), remove duplicates before saving.

### Process and visualize

```{r process_cbmc, eval = can_run, purl = FALSE}
DefaultAssay(cbmc) <- "RNA"
cbmc <- NormalizeData(cbmc, verbose = FALSE)
cbmc <- FindVariableFeatures(cbmc, verbose = FALSE)
cbmc <- ScaleData(cbmc, verbose = FALSE)
cbmc <- RunPCA(cbmc, verbose = FALSE)
cbmc <- RunUMAP(cbmc, dims = 1:30, verbose = FALSE)
cbmc <- NormalizeData(cbmc, assay = "ADT", normalization.method = "CLR", verbose = FALSE)
```

```{r plot_cbmc, eval = can_run, purl = FALSE, fig.width = 14, fig.height = 6}
library(patchwork)

p1 <- DimPlot(cbmc, label = TRUE, pt.size = 0.3) + NoLegend() + ggtitle("RNA clusters")
DefaultAssay(cbmc) <- "ADT"
p2 <- FeaturePlot(cbmc, features = "adt_CD3", pt.size = 0.3) + ggtitle("CD3 protein (ADT)")
DefaultAssay(cbmc) <- "RNA"
p1 + p2
```

## Save with writeH5MU

`writeH5MU()` exports all assays to a single h5mu file. Assay names are automatically mapped to standard MuData modality names:

| Seurat Assay | h5mu Modality |
|-------------|---------------|
| RNA | rna |
| ADT | prot |
| ATAC | atac |
| Spatial | spatial |
| Other | lowercase name |

```{r save_h5mu, eval = can_run, purl = FALSE}
writeH5MU(cbmc, "cbmc_multimodal.h5mu", overwrite = TRUE)
```

### Inspect in Python

If you have [muon](https://muon-tutorials.readthedocs.io/) installed, you can inspect the h5mu file:

```{python inspect_h5mu, eval = can_run && has_muon}
import mudata as md
mdata = md.read_h5mu("cbmc_multimodal.h5mu")
print(mdata)
print("\nModalities:", list(mdata.mod.keys()))
for mod_name, mod in mdata.mod.items():
    print(f"  {mod_name}: {mod.n_obs} cells x {mod.n_vars} features")
```

### scanpy on RNA modality

We can run standard scanpy preprocessing on the RNA modality extracted from the h5mu file:

```{python scanpy_on_h5mu, eval = can_run && has_muon && has_scanpy}
import scanpy as sc

rna = mdata.mod["rna"].copy()
sc.pp.normalize_total(rna)
sc.pp.log1p(rna)
sc.pp.highly_variable_genes(rna, n_top_genes=2000)
sc.pp.pca(rna)
sc.pp.neighbors(rna)
sc.tl.umap(rna)
sc.tl.leiden(rna, flavor="igraph")

sc.pl.umap(rna, color="leiden", title="CBMC RNA (from h5mu) — Leiden clusters")
```

## Load with readH5MU

`readH5MU()` reads the h5mu file back into a Seurat object, reversing the modality name mapping:

| h5mu Modality | Seurat Assay |
|---------------|-------------|
| rna | RNA |
| prot | ADT |
| atac | ATAC |
| spatial | Spatial |
| Other | preserved |

```{r load_h5mu, eval = can_load, purl = FALSE}
cbmc_loaded <- readH5MU("cbmc_multimodal.h5mu")
cbmc_loaded
```

`readH5MU()` uses scConvert's native reader — no MuDataSeurat or Python required.

## Verify round-trip preservation

```{r verify_roundtrip, eval = can_load, purl = FALSE}
cat("=== Original ===\n")
cat("Assays:", paste(Assays(cbmc), collapse = ", "), "\n")
cat("Cells:", ncol(cbmc), "\n")

cat("\n=== Loaded ===\n")
cat("Assays:", paste(Assays(cbmc_loaded), collapse = ", "), "\n")
cat("Cells:", ncol(cbmc_loaded), "\n")

# Check metadata
cat("\nMetadata columns (original):", ncol(cbmc[[]]), "\n")
cat("Metadata columns (loaded):", ncol(cbmc_loaded[[]]), "\n")
```

### Verify expression values

```{r verify_expression, eval = can_load, purl = FALSE}
common_c <- intersect(colnames(cbmc), colnames(cbmc_loaded))
common_g <- intersect(rownames(cbmc[["RNA"]]), rownames(cbmc_loaded[["RNA"]]))
set.seed(42)
sc <- sample(common_c, min(200, length(common_c)))
sg <- sample(common_g, min(100, length(common_g)))

orig_vals <- as.numeric(GetAssayData(cbmc, assay = "RNA", layer = "counts")[sg, sc])
rt_vals <- as.numeric(GetAssayData(cbmc_loaded, assay = "RNA", layer = "counts")[sg, sc])
cat("RNA values identical:", identical(orig_vals, rt_vals), "\n")
cat("Max abs diff:", max(abs(orig_vals - rt_vals)), "\n")

# Check ADT assay too
adt_genes <- intersect(rownames(cbmc[["ADT"]]), rownames(cbmc_loaded[["ADT"]]))
if (length(adt_genes) > 0) {
  adt_orig <- as.numeric(GetAssayData(cbmc, assay = "ADT", layer = "counts")[adt_genes, sc])
  adt_rt <- as.numeric(GetAssayData(cbmc_loaded, assay = "ADT", layer = "counts")[adt_genes, sc])
  cat("ADT values identical:", identical(adt_orig, adt_rt), "\n")
}
```

## Custom modality name mapping

You can override the default mapping with `assays` (for `writeH5MU()`) or `assay.names` (for `readH5MU()`):

```{r custom_mapping, eval = can_load, purl = FALSE}
# Save with custom modality names (native writer)
writeH5MU(cbmc, filename = "cbmc_roundtrip.h5mu", overwrite = TRUE)

# Load with custom assay names
cbmc_custom <- readH5MU(
  "cbmc_roundtrip.h5mu",
  assay.names = c(rna = "RNA", prot = "Protein")
)

cat("Assays:", paste(Assays(cbmc_custom), collapse = ", "), "\n")
```

## Conflict resolution

When a Seurat object has multiple assays that would map to the same modality name, `writeH5MU()` automatically resolves the conflict. For example, if you have both RNA and SCT assays (both map to "rna" by default), SCT is remapped to "sct":

| Seurat Assays | Modality Mapping |
|--------------|-----------------|
| RNA, ADT | rna, prot (no conflict) |
| RNA, SCT, ADT | rna, sct, prot (SCT remapped) |

This ensures each modality has a unique name in the h5mu file.

## Data mapping reference

### writeH5MU: Seurat to h5mu

| Seurat Component | h5mu Location | Notes |
|-----------------|--------------|-------|
| Each assay | `/mod/{modality}/` | One AnnData per modality |
| Assay counts | `/mod/{modality}/X` | Expression matrix |
| Assay metadata | `/mod/{modality}/var` | Feature metadata |
| Cell metadata | `/obs` | Global cell annotations |
| Reductions | `/mod/{modality}/obsm/X_{name}` | Per-modality |
| Graphs | `/mod/{modality}/obsp/{name}` | Per-modality |

### readH5MU: h5mu to Seurat

| h5mu Location | Seurat Component | Notes |
|--------------|-----------------|-------|
| `/mod/{modality}/` | Assay (renamed) | Modality -> assay name mapping |
| `/mod/{modality}/X` | Assay data | Expression matrix |
| `/obs` | `meta.data` | Global cell metadata merged |
| `/mod/{modality}/obsm/spatial` | Spatial coordinates | If present |

## Session Info

```{r sessionInfo, purl = FALSE}
sessionInfo()
```

```{r cleanup-final, echo=FALSE, results='hide', purl = FALSE}
outs <- c("cbmc_multimodal.h5mu", "cbmc_roundtrip.h5mu")
existing_files <- outs[file.exists(outs)]
if (length(existing_files) > 0) file.remove(existing_files)
```
