---
title: "CITE-seq Analysis Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CITE-seq Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

## Introduction

CITE-seq (Cellular Indexing of Transcriptomes and Epitopes by Sequencing)
measures RNA transcription and surface protein abundance from the same cells.
This produces two assays per cell: RNA gene expression and ADT
(antibody-derived tag) protein counts. scConvert handles both assays during
format conversion, giving you two export options:

- **h5ad**: Exports a single assay (RNA or ADT) per file, ideal for scanpy
  workflows
- **h5mu**: Exports all assays in one file, ideal for muon/MuData workflows

## Load CITE-seq data

The shipped `citeseq_demo.rds` contains 500 cells with RNA (2000 genes) and
ADT (10 antibodies: CD3, CD4, CD8, CD45RA, CD56, CD16, CD11c, CD14, CD19,
CD34), pre-processed with PCA, UMAP, and clustering.

```{r load-data, eval = TRUE}
obj <- readRDS(system.file("extdata", "citeseq_demo.rds", package = "scConvert"))

cat("Cells:", ncol(obj), "\n")
cat("Assays:", paste(Assays(obj), collapse = ", "), "\n")
cat("RNA features:", nrow(obj[["RNA"]]), "\n")
cat("ADT features:", nrow(obj[["ADT"]]), "\n")
cat("ADT markers:", paste(rownames(obj[["ADT"]]), collapse = ", "), "\n")
```

## Visualize RNA clusters

```{r dimplot-rna, eval = TRUE, fig.width = 7, fig.height = 5}
DimPlot(obj, group.by = "seurat_clusters", label = TRUE, pt.size = 0.8) +
  ggtitle("RNA-based clusters")
```

## Visualize protein expression

ADT markers reveal cell-surface protein levels that complement the
transcriptomic clusters. CD3 marks T cells; CD14 marks monocytes.

```{r featureplot-adt, eval = TRUE, fig.width = 12, fig.height = 5}
library(patchwork)

DefaultAssay(obj) <- "ADT"
p1 <- FeaturePlot(obj, features = "CD3", pt.size = 0.8) + ggtitle("CD3 (T cells)")
p2 <- FeaturePlot(obj, features = "CD14", pt.size = 0.8) + ggtitle("CD14 (Monocytes)")
DefaultAssay(obj) <- "RNA"

p1 + p2
```

### ADT expression across clusters

```{r vlnplot-adt, eval = TRUE, fig.width = 8, fig.height = 5}
DefaultAssay(obj) <- "ADT"
VlnPlot(obj, features = "CD3", group.by = "seurat_clusters", pt.size = 0.1) +
  ggtitle("CD3 protein across clusters") +
  NoLegend()
DefaultAssay(obj) <- "RNA"
```

## Export to h5ad (single assay)

`writeH5AD()` writes one assay at a time. By default it writes the active
assay (RNA). This is the right choice for standard scanpy workflows.

```{r write-h5ad, eval = TRUE}
h5ad_path <- file.path(tempdir(), "citeseq_rna.h5ad")
writeH5AD(obj, h5ad_path, overwrite = TRUE)
cat("Wrote RNA to:", basename(h5ad_path), "\n")
cat("File size:", round(file.size(h5ad_path) / 1024^2, 1), "MB\n")
```

## Export to h5mu (multi-assay)

`writeH5MU()` writes all assays into a single file, each as a separate
modality. This preserves both RNA and ADT together.

```{r write-h5mu, eval = TRUE}
h5mu_path <- file.path(tempdir(), "citeseq.h5mu")
writeH5MU(obj, h5mu_path, overwrite = TRUE)
cat("Wrote RNA + ADT to:", basename(h5mu_path), "\n")
cat("File size:", round(file.size(h5mu_path) / 1024^2, 1), "MB\n")
```

## Reload and verify

### From h5ad (RNA only)

```{r reload-h5ad, eval = TRUE}
loaded_h5ad <- readH5AD(h5ad_path)
cat("h5ad loaded:", ncol(loaded_h5ad), "cells x", nrow(loaded_h5ad), "features\n")
cat("Assays:", paste(Assays(loaded_h5ad), collapse = ", "), "\n")
```

### From h5mu (RNA + ADT)

```{r reload-h5mu, eval = TRUE}
loaded_h5mu <- readH5MU(h5mu_path)
cat("h5mu loaded:", ncol(loaded_h5mu), "cells\n")
cat("Assays:", paste(Assays(loaded_h5mu), collapse = ", "), "\n")
cat("RNA features:", nrow(loaded_h5mu[["RNA"]]), "\n")
cat("ADT features:", nrow(loaded_h5mu[["ADT"]]), "\n")
```

### Verify expression preservation

```{r verify, eval = TRUE}
common_cells <- intersect(colnames(obj), colnames(loaded_h5mu))

# RNA counts
common_genes <- intersect(rownames(obj[["RNA"]]), rownames(loaded_h5mu[["RNA"]]))
orig <- as.numeric(GetAssayData(obj, assay = "RNA", layer = "counts")[
  head(common_genes, 100), head(common_cells, 100)])
rt <- as.numeric(GetAssayData(loaded_h5mu, assay = "RNA", layer = "counts")[
  head(common_genes, 100), head(common_cells, 100)])
cat("RNA counts identical:", identical(orig, rt), "\n")

# ADT counts
common_adt <- intersect(rownames(obj[["ADT"]]), rownames(loaded_h5mu[["ADT"]]))
orig_adt <- as.numeric(GetAssayData(obj, assay = "ADT", layer = "counts")[
  common_adt, head(common_cells, 100)])
rt_adt <- as.numeric(GetAssayData(loaded_h5mu, assay = "ADT", layer = "counts")[
  common_adt, head(common_cells, 100)])
cat("ADT counts identical:", identical(orig_adt, rt_adt), "\n")
```

## When to use h5ad vs h5mu

| Scenario | Recommended format | Why |
|----------|--------------------|-----|
| scanpy RNA analysis | h5ad | scanpy expects single-AnnData input |
| Share with Python collaborator (multi-assay) | h5mu | Keeps RNA + ADT together in one file |
| Archive a CITE-seq experiment | h5mu | Preserves all modalities |
| Feed into cellxgene | h5ad | cellxgene reads h5ad, not h5mu |
| Downstream muon/MOFA+ analysis | h5mu | muon operates on MuData objects |

## Python interoperability

Both exported files are directly readable in Python.

```{python python-h5ad, eval = FALSE}
# Requires Python: pip install scanpy
import scanpy as sc

adata = sc.read_h5ad("citeseq_rna.h5ad")
print(adata)
sc.pl.umap(adata, color="seurat_clusters")
```

```{python python-h5mu, eval = FALSE}
# Requires Python: pip install mudata
import mudata as md

mdata = md.read_h5mu("citeseq.h5mu")
print(mdata)
for name, mod in mdata.mod.items():
    print(f"  {name}: {mod.n_obs} cells x {mod.n_vars} features")
```

## Clean up

```{r cleanup, eval = TRUE}
unlink(c(h5ad_path, h5mu_path))
```

## Session Info

```{r session-info}
sessionInfo()
```
