---
title: "scATAC-seq Data Conversion with Signac"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scATAC-seq Data Conversion with Signac}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  out.width = "100%",
  dev = "png",
  purl = FALSE
)
has_scanpy <- FALSE
if (requireNamespace("reticulate", quietly = TRUE)) {
  scverse_path <- "/opt/homebrew/Caskroom/miniconda/base/envs/scverse"
  if (dir.exists(scverse_path)) {
    reticulate::use_condaenv("scverse", required = FALSE)
    Sys.setenv(
      OMP_NUM_THREADS = "1", OPENBLAS_NUM_THREADS = "1",
      MKL_NUM_THREADS = "1", NUMBA_THREADING_LAYER = "tbb"
    )
    has_scanpy <- reticulate::py_module_available("scanpy")
    if (has_scanpy) {
      reticulate::py_run_string("import matplotlib; matplotlib.use('Agg')")
    }
  }
}
```

```{r cleanup, echo=FALSE, results='hide', purl = FALSE}
outs <- c(
  "pbmc_atac.h5ad",
  "pbmc_atac_roundtrip.h5ad",
  "pbmc_atac_multiome.h5mu",
  "pbmc_atac_cli.h5ad"
)
existing_files <- outs[file.exists(outs)]
if (length(existing_files) > 0) file.remove(existing_files)
```

## Introduction

Single-cell ATAC-seq (scATAC-seq) measures chromatin accessibility at single-cell resolution. The primary data structure is a **peak-by-cell count matrix**, where each row represents a genomic region (peak) and each value indicates how many Tn5 insertion events were observed in that region for a given cell.

### scATAC-seq data components

| Component | Description | Format |
|-----------|-------------|--------|
| Peak matrix | Peaks x cells count matrix | Sparse matrix (HDF5, MEX) |
| Fragment file | Per-read genomic coordinates | `fragments.tsv.gz` + `.tbi` index |
| Cell metadata | Per-cell QC and annotations | Data frame |
| Gene annotations | Genomic coordinates for genes | GRanges (GTF/GFF) |
| Motif matrix | TF binding motif enrichment | Matrix in ChromatinAssay |

In R, the [Signac](https://stuartlab.org/signac/) package extends Seurat with the `ChromatinAssay` class, which wraps the peak matrix alongside fragment file paths, gene annotations, and motif information. In Python, scATAC-seq data is typically stored as an AnnData object (h5ad) with peaks as `var` and cells as `obs`, following the same conventions as scRNA-seq.

### What scConvert handles

scConvert converts the **peak count matrix** and associated **cell metadata** and **embeddings** between R and Python formats. Signac-specific annotations (fragment files, motif matrices, gene annotations) are not part of the h5ad specification and require separate handling.

```{r packages, purl = FALSE}
library(Seurat)
library(Signac)
library(scConvert)
library(Matrix)
```

## Create synthetic scATAC-seq data

We generate a realistic synthetic scATAC-seq dataset with 500 cells and 2000 peaks. Peak names follow the standard `chr-start-end` genomic coordinate format. The count matrix is 95% sparse, reflecting the typical sparsity of real scATAC-seq data.

```{r create_synthetic}
set.seed(42)

n_cells <- 500
n_peaks <- 2000
sparsity <- 0.95

# --- Generate realistic peak names across chromosomes ---
chrom_sizes <- c(
  chr1 = 249e6, chr2 = 243e6, chr3 = 198e6, chr4 = 191e6, chr5 = 181e6,
  chr6 = 171e6, chr7 = 159e6, chr8 = 145e6, chr9 = 138e6, chr10 = 134e6,
  chr11 = 135e6, chr12 = 133e6, chr13 = 114e6, chr14 = 107e6, chr15 = 102e6,
  chr16 = 90e6, chr17 = 83e6, chr18 = 80e6, chr19 = 59e6, chr20 = 64e6,
  chr21 = 47e6, chr22 = 51e6
)

# Distribute peaks proportionally across chromosomes
peaks_per_chrom <- round(n_peaks * chrom_sizes / sum(chrom_sizes))
peaks_per_chrom[1] <- peaks_per_chrom[1] + (n_peaks - sum(peaks_per_chrom))

peak_names <- character(0)
for (i in seq_along(peaks_per_chrom)) {
  chr <- names(peaks_per_chrom)[i]
  n <- peaks_per_chrom[i]
  if (n <= 0) next
  starts <- sort(sample(seq(10000, chrom_sizes[i] - 2000, by = 1000), n))
  widths <- sample(500:1500, n, replace = TRUE)
  peak_names <- c(peak_names, paste0(chr, "-", starts, "-", starts + widths))
}
peak_names <- peak_names[seq_len(n_peaks)]

# --- Generate cell barcodes ---
barcodes <- paste0("CELL_", sprintf("%04d", seq_len(n_cells)), "-1")

# --- Generate sparse count matrix (95% sparsity) ---
n_nonzero <- round(n_peaks * n_cells * (1 - sparsity))
rows <- sample(n_peaks, n_nonzero, replace = TRUE)
cols <- sample(n_cells, n_nonzero, replace = TRUE)
vals <- rpois(n_nonzero, lambda = 2) + 1L  # counts >= 1

counts <- sparseMatrix(
  i = rows, j = cols, x = vals,
  dims = c(n_peaks, n_cells),
  dimnames = list(peak_names, barcodes)
)

cat("Matrix dimensions:", nrow(counts), "peaks x", ncol(counts), "cells\n")
cat("Sparsity:", round(1 - nnzero(counts) / (nrow(counts) * ncol(counts)), 3), "\n")
```

## Load into Signac ChromatinAssay

We create a `ChromatinAssay` and wrap it in a Seurat object, then add metadata and embeddings to produce a complete scATAC-seq object.

```{r load_signac}
# Create a ChromatinAssay (without fragment file or annotations for portability)
chrom_assay <- CreateChromatinAssay(
  counts = counts,
  sep = c("-", "-"),
  min.cells = 0,
  min.features = 0
)

# Create Seurat object
pbmc_atac <- CreateSeuratObject(
  counts = chrom_assay,
  assay = "ATAC"
)

cat("Cells:", ncol(pbmc_atac), "\n")
cat("Peaks:", nrow(pbmc_atac), "\n")
cat("Assay class:", class(pbmc_atac[["ATAC"]])[1], "\n")

# Inspect peak names (genomic coordinates)
head(rownames(pbmc_atac), 5)
```

### Add QC metadata and cell annotations

```{r add_metadata}
set.seed(123)

# QC-like metadata
pbmc_atac$nCount_peaks <- colSums(GetAssayData(pbmc_atac, layer = "counts"))
pbmc_atac$nucleosome_signal <- runif(n_cells, 0.2, 2.0)
pbmc_atac$TSS_enrichment <- runif(n_cells, 2, 15)
pbmc_atac$pct_reads_in_peaks <- runif(n_cells, 0.3, 0.9)

# Cell-type annotations (as factor)
cell_types <- factor(
  sample(c("CD4_T", "CD8_T", "B_cell", "NK", "Monocyte", "DC"),
         n_cells, replace = TRUE,
         prob = c(0.30, 0.15, 0.20, 0.10, 0.20, 0.05)),
  levels = c("CD4_T", "CD8_T", "B_cell", "NK", "Monocyte", "DC")
)
pbmc_atac$cell_type <- cell_types
pbmc_atac$seurat_clusters <- factor(as.integer(cell_types) - 1L)

# Add fake UMAP coordinates (cluster-structured for visualization)
umap_centers <- matrix(c(
  -5, 3,   # CD4_T
   5, 3,   # CD8_T
  -3, -4,  # B_cell
   4, -3,  # NK
   0, 0,   # Monocyte
  -6, -1   # DC
), ncol = 2, byrow = TRUE)

umap_coords <- matrix(0, nrow = n_cells, ncol = 2)
for (i in seq_len(n_cells)) {
  ct <- as.integer(cell_types[i])
  umap_coords[i, ] <- umap_centers[ct, ] + rnorm(2, sd = 0.8)
}
colnames(umap_coords) <- c("umap_1", "umap_2")
rownames(umap_coords) <- barcodes

pbmc_atac[["umap"]] <- CreateDimReducObject(
  embeddings = umap_coords,
  key = "umap_",
  assay = "ATAC"
)

cat("Metadata columns:", paste(colnames(pbmc_atac[[]]), collapse = ", "), "\n")
cat("Reductions:", paste(Reductions(pbmc_atac), collapse = ", "), "\n")
```

## Convert Signac to h5ad

The `writeH5AD()` function writes the peak matrix, cell metadata, and embeddings directly to h5ad format. The `ChromatinAssay` peak matrix is extracted as a standard sparse matrix during conversion.

```{r signac_to_h5ad}
writeH5AD(pbmc_atac, filename = "pbmc_atac.h5ad", overwrite = TRUE)
cat("h5ad file size:", round(file.size("pbmc_atac.h5ad") / 1024^2, 1), "MB\n")
```

The resulting h5ad file can be loaded in Python with scanpy or episcanpy:

```{python load_atac_scanpy, eval = has_scanpy && file.exists("pbmc_atac.h5ad")}
import scanpy as sc

adata = sc.read_h5ad("pbmc_atac.h5ad")
print(adata)
print(f"\nPeak names (first 5): {list(adata.var_names[:5])}")
print(f"obsm keys: {list(adata.obsm.keys())}")
print(f"obs columns: {list(adata.obs.columns)[:10]}")

if "X_umap" in adata.obsm:
    sc.pl.umap(adata, color="cell_type", title="scATAC UMAP — scanpy (from scConvert h5ad)")
```

### CLI equivalent

The following examples show how to invoke scConvert from the command line. These use placeholder filenames and are not executed here.

```{r signac_to_h5ad_cli, eval = FALSE}
scConvert_cli("pbmc_atac.rds", "pbmc_atac_cli.h5ad")
```

Or from the command line:

```bash
scconvert pbmc_atac.rds pbmc_atac_cli.h5ad
```

## Convert h5ad back to Seurat with Signac

When loading an h5ad file that contains a peak matrix, `readH5AD()` returns a standard Seurat object. You can then upgrade the assay to a Signac `ChromatinAssay` if needed.

```{r h5ad_to_signac}
# Load the h5ad as a standard Seurat object
atac_loaded <- readH5AD("pbmc_atac.h5ad")

cat("Cells:", ncol(atac_loaded), "\n")
cat("Peaks:", nrow(atac_loaded), "\n")
cat("Default assay:", DefaultAssay(atac_loaded), "\n")

# Inspect: peak names should be chr-start-end format
head(rownames(atac_loaded), 5)
```

### Upgrade to ChromatinAssay

If you need Signac functionality (e.g., motif analysis, coverage plots), upgrade the loaded assay to a `ChromatinAssay`. This requires re-attaching any Signac-specific annotations.

```{r upgrade_chromatin}
# Extract the counts matrix from the loaded object
peak_counts <- GetAssayData(atac_loaded, layer = "counts")

# Create a ChromatinAssay from the loaded data
chrom_assay_rt <- CreateChromatinAssay(
  counts = peak_counts,
  sep = c("-", "-"),
  min.cells = 0,
  min.features = 0
)

# Replace the assay in the loaded object
atac_loaded[["ATAC"]] <- chrom_assay_rt
cat("Upgraded assay class:", class(atac_loaded[["ATAC"]])[1], "\n")

# Optional: re-attach fragment file if available locally
# Fragments(atac_loaded) <- CreateFragmentObject("fragments.tsv.gz")

# Optional: add gene annotations
# library(EnsDb.Hsapiens.v86)
# annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)
# Annotation(atac_loaded) <- annotations
```

### CLI equivalent

```{r h5ad_to_seurat_cli, eval = FALSE}
scConvert_cli("pbmc_atac.h5ad", "pbmc_atac_roundtrip.rds")
```

## Round-trip fidelity verification

After converting Signac -> h5ad -> Seurat, we verify that key data structures are preserved.

```{r roundtrip_verify}
# Reload from h5ad
atac_rt <- readH5AD("pbmc_atac.h5ad")

# --- Dimension check ---
cat("=== Dimensions ===\n")
cat("Original:  ", ncol(pbmc_atac), "cells x", nrow(pbmc_atac), "peaks\n")
cat("Roundtrip: ", ncol(atac_rt), "cells x", nrow(atac_rt), "peaks\n")
stopifnot(ncol(pbmc_atac) == ncol(atac_rt))
stopifnot(nrow(pbmc_atac) == nrow(atac_rt))
cat("Dimension check PASSED\n\n")

# --- Peak name preservation ---
cat("=== Peak Names ===\n")
orig_peaks <- sort(rownames(pbmc_atac))
rt_peaks <- sort(rownames(atac_rt))
cat("Peaks match:", identical(orig_peaks, rt_peaks), "\n")
stopifnot(identical(orig_peaks, rt_peaks))
cat("Peak name check PASSED\n\n")

# --- Cell barcode preservation ---
cat("=== Cell Barcodes ===\n")
orig_cells <- sort(colnames(pbmc_atac))
rt_cells <- sort(colnames(atac_rt))
cat("Barcodes match:", identical(orig_cells, rt_cells), "\n")
stopifnot(identical(orig_cells, rt_cells))
cat("Barcode check PASSED\n\n")

# --- Count matrix identity ---
cat("=== Count Matrix Fidelity ===\n")
orig_counts <- GetAssayData(pbmc_atac, layer = "counts")
rt_counts <- GetAssayData(atac_rt, layer = "counts")
# Reorder to match
rt_counts <- rt_counts[rownames(orig_counts), colnames(orig_counts)]
cat("Count matrices identical:", identical(as(orig_counts, "dgCMatrix"),
                                          as(rt_counts, "dgCMatrix")), "\n")
stopifnot(all.equal(as.matrix(orig_counts), as.matrix(rt_counts),
                    check.attributes = FALSE))
cat("Count matrix check PASSED\n\n")

# --- Metadata preservation ---
cat("=== Cell Metadata ===\n")
orig_meta_cols <- sort(colnames(pbmc_atac[[]]))
rt_meta_cols <- sort(colnames(atac_rt[[]]))
shared_cols <- intersect(orig_meta_cols, rt_meta_cols)
cat("Original metadata columns:", length(orig_meta_cols), "\n")
cat("Roundtrip metadata columns:", length(rt_meta_cols), "\n")
cat("Shared columns:", length(shared_cols), "\n")

# Verify cell_type factor survived
if ("cell_type" %in% shared_cols) {
  orig_ct <- as.character(pbmc_atac$cell_type)
  names(orig_ct) <- colnames(pbmc_atac)
  rt_ct <- as.character(atac_rt$cell_type)
  names(rt_ct) <- colnames(atac_rt)
  cat("cell_type match:", identical(orig_ct[sort(names(orig_ct))],
                                    rt_ct[sort(names(rt_ct))]), "\n")
}

# --- Embedding preservation ---
cat("\n=== Embeddings ===\n")
if ("umap" %in% Reductions(atac_rt)) {
  orig_umap <- Embeddings(pbmc_atac, "umap")
  rt_umap <- Embeddings(atac_rt, "umap")
  rt_umap <- rt_umap[rownames(orig_umap), ]
  umap_rmse <- sqrt(mean((orig_umap - rt_umap)^2))
  cat("UMAP RMSE:", umap_rmse, "\n")
  stopifnot(umap_rmse < 1e-6)
  cat("UMAP check PASSED\n")
} else {
  cat("UMAP not found in roundtrip object (may need recomputation)\n")
}
```

## Multiome: RNA + ATAC to h5mu

For 10x Multiome experiments (simultaneous RNA + ATAC from the same cells), the h5mu format stores both modalities in a single file. Each assay becomes a separate modality.

```{r multiome_to_h5mu}
set.seed(42)

# --- Generate synthetic RNA data for the same 500 cells ---
n_genes <- 200
n_rna_peaks <- 500  # smaller ATAC assay for the multiome demo

# RNA counts (sparse)
rna_nonzero <- round(n_genes * n_cells * 0.10)  # 90% sparse
rna_counts <- sparseMatrix(
  i = sample(n_genes, rna_nonzero, replace = TRUE),
  j = sample(n_cells, rna_nonzero, replace = TRUE),
  x = rpois(rna_nonzero, lambda = 3) + 1L,
  dims = c(n_genes, n_cells),
  dimnames = list(
    paste0("Gene", seq_len(n_genes)),
    barcodes
  )
)

# Smaller ATAC counts for multiome demo
atac_multiome_counts <- counts[seq_len(n_rna_peaks), ]

# Build a multimodal Seurat object
multiome <- CreateSeuratObject(
  counts = CreateChromatinAssay(
    counts = atac_multiome_counts,
    sep = c("-", "-"),
    min.cells = 0,
    min.features = 0
  ),
  assay = "ATAC"
)
multiome[["RNA"]] <- CreateAssay5Object(counts = rna_counts)
multiome$cell_type <- cell_types

cat("Assays:", paste(Assays(multiome), collapse = ", "), "\n")
cat("ATAC peaks:", nrow(multiome[["ATAC"]]), "\n")
cat("RNA genes:", nrow(multiome[["RNA"]]), "\n")
cat("Cells:", ncol(multiome), "\n")

# Export both modalities to h5mu
writeH5MU(multiome, filename = "pbmc_atac_multiome.h5mu", overwrite = TRUE)
cat("h5mu file size:", round(file.size("pbmc_atac_multiome.h5mu") / 1024^2, 1), "MB\n")
```

### Load h5mu back

```{r load_multiome_h5mu}
multiome_rt <- readH5MU("pbmc_atac_multiome.h5mu")

cat("Assays:", paste(Assays(multiome_rt), collapse = ", "), "\n")
cat("ATAC peaks:", nrow(multiome_rt[["ATAC"]]), "\n")
cat("RNA genes:", nrow(multiome_rt[["RNA"]]), "\n")
cat("Cells:", ncol(multiome_rt), "\n")
```

### CLI equivalent

```{r multiome_cli, eval = FALSE}
scConvert_cli("multiome.rds", "multiome.h5mu")
```

## What is and is not preserved

### Preserved during conversion

| Component | Seurat/Signac | h5ad/h5mu | Notes |
|-----------|--------------|-----------|-------|
| Peak count matrix | `counts` layer | `X` | Sparse matrix, lossless |
| Normalized matrix | `data` layer | `X` or `layers` | TF-IDF values |
| Cell metadata | `meta.data` | `obs` | All columns preserved |
| Peak metadata | `meta.features` | `var` | Feature-level annotations |
| LSI embeddings | `reductions$lsi` | `obsm['X_lsi']` | Latent semantic indexing |
| UMAP coordinates | `reductions$umap` | `obsm['X_umap']` | Visualization coords |
| Variable features | `VariableFeatures()` | `var['highly_variable']` | Top accessible peaks |
| Cluster assignments | `meta.data$seurat_clusters` | `obs['seurat_clusters']` | Cell labels |

### NOT preserved (Signac-specific)

| Component | Why | Workaround |
|-----------|-----|------------|
| Fragment file paths | Local filesystem reference, not data | Copy `fragments.tsv.gz` + `.tbi` separately; re-attach with `CreateFragmentObject()` |
| Gene annotations | GRanges object, no h5ad equivalent | Re-add with `Annotation()` after loading (e.g., from EnsDb.Hsapiens.v86) |
| Motif matrices | Signac-specific slot | Re-compute with `AddMotifs()` using a PWM database (e.g., JASPAR) |
| ChromatinAssay class | R-specific S4 class | Upgrade with `CreateChromatinAssay()` after loading (see above) |
| Links (peak-gene) | Signac-specific slot | Re-compute with `LinkPeaks()` after re-attaching annotations |

### Practical recommendations

1. **Always transfer fragment files separately.** They are large (often >10 GB) and are not embedded in h5ad. Copy both `fragments.tsv.gz` and `fragments.tsv.gz.tbi`.

2. **Re-attach annotations after loading.** Gene annotations and motif databases are organism-specific and best loaded from canonical sources (Ensembl, JASPAR) rather than serialized.

3. **Peak names encode genomic coordinates.** The `chr:start-end` format is preserved as `var_names` in h5ad, so peak identity survives round-trip without loss.

4. **For Multiome data, prefer h5mu.** It keeps RNA and ATAC in separate modalities within a single file, matching the muon/MuData convention in Python.

## See Also

- [Multimodal: CITE-seq and Multiome](multimodal-citeseq.html) -- combining RNA + ATAC or RNA + ADT in a single object

## Session Info

```{r sessionInfo, purl = FALSE}
sessionInfo()
```

```{r cleanup-final, echo=FALSE, results='hide', purl = FALSE}
outs <- c(
  "pbmc_atac.h5ad",
  "pbmc_atac_roundtrip.h5ad",
  "pbmc_atac_multiome.h5mu",
  "pbmc_atac_cli.h5ad"
)
existing_files <- outs[file.exists(outs)]
if (length(existing_files) > 0) file.remove(existing_files)
```
