---
title: "scATAC-seq Data Conversion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scATAC-seq Data Conversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

## Introduction

Single-cell ATAC-seq measures chromatin accessibility at single-cell
resolution. The core data structure is a **peak-by-cell count matrix** --
structurally identical to a gene-by-cell RNA matrix. scConvert converts this
matrix along with cell metadata, embeddings, and cluster labels between
formats, enabling interoperability between R (Seurat/Signac) and Python
(scanpy/episcanpy) workflows.

This vignette demonstrates the h5ad roundtrip using the shipped PBMC demo data
as a stand-in. The matrix structure and conversion mechanics are the same
regardless of whether rows represent genes or peaks.

## Load demo data

```{r load-data, eval = TRUE}
obj <- readRDS(system.file("extdata", "pbmc_demo.rds", package = "scConvert"))

cat("Cells:", ncol(obj), "\n")
cat("Features:", nrow(obj), "\n")
cat("Reductions:", paste(Reductions(obj), collapse = ", "), "\n")
cat("Metadata columns:", paste(colnames(obj[[]]), collapse = ", "), "\n")
```

```{r dimplot-original, eval = TRUE, fig.width = 7, fig.height = 5}
DimPlot(obj, group.by = "seurat_annotations", label = TRUE, pt.size = 0.8) +
  ggtitle("PBMC clusters (original)")
```

## Roundtrip through h5ad

Write the Seurat object to h5ad and read it back.

```{r roundtrip, eval = TRUE}
h5ad_path <- file.path(tempdir(), "pbmc_atac_demo.h5ad")

writeH5AD(obj, h5ad_path, overwrite = TRUE)
cat("Wrote:", basename(h5ad_path), "\n")
cat("File size:", round(file.size(h5ad_path) / 1024^2, 1), "MB\n")

loaded <- readH5AD(h5ad_path)
cat("Loaded:", ncol(loaded), "cells x", nrow(loaded), "features\n")
```

## Verify preservation

### Dimensions and barcodes

```{r verify-dims, eval = TRUE}
cat("Cells match:", ncol(obj) == ncol(loaded), "\n")
cat("Features match:", nrow(obj) == nrow(loaded), "\n")

orig_cells <- sort(colnames(obj))
rt_cells <- sort(colnames(loaded))
cat("Barcodes identical:", identical(orig_cells, rt_cells), "\n")

orig_features <- sort(rownames(obj))
rt_features <- sort(rownames(loaded))
cat("Feature names identical:", identical(orig_features, rt_features), "\n")
```

### Count matrix

```{r verify-counts, eval = TRUE}
common_cells <- intersect(colnames(obj), colnames(loaded))
common_feats <- intersect(rownames(obj), rownames(loaded))

orig_vals <- as.numeric(GetAssayData(obj, layer = "counts")[
  head(common_feats, 200), head(common_cells, 200)])
rt_vals <- as.numeric(GetAssayData(loaded, layer = "counts")[
  head(common_feats, 200), head(common_cells, 200)])
cat("Count values identical:", identical(orig_vals, rt_vals), "\n")
```

### Metadata

```{r verify-meta, eval = TRUE}
shared_cols <- intersect(colnames(obj[[]]), colnames(loaded[[]]))
cat("Metadata columns preserved:", length(shared_cols), "/", ncol(obj[[]]), "\n")
cat("Columns:", paste(shared_cols, collapse = ", "), "\n")

if ("seurat_annotations" %in% shared_cols) {
  orig_ann <- as.character(obj$seurat_annotations[common_cells])
  rt_ann <- as.character(loaded$seurat_annotations[common_cells])
  cat("Cell annotations match:", identical(orig_ann, rt_ann), "\n")
}
```

### Embeddings

```{r verify-embeddings, eval = TRUE}
cat("Original reductions:", paste(Reductions(obj), collapse = ", "), "\n")
cat("Loaded reductions:", paste(Reductions(loaded), collapse = ", "), "\n")

if ("umap" %in% Reductions(loaded)) {
  orig_umap <- Embeddings(obj, "umap")[common_cells, ]
  rt_umap <- Embeddings(loaded, "umap")[common_cells, ]
  max_diff <- max(abs(orig_umap - rt_umap))
  cat("UMAP max absolute difference:", max_diff, "\n")
}
```

## Compare plots

```{r dimplot-compare, eval = TRUE, fig.width = 12, fig.height = 5}
library(patchwork)

p1 <- DimPlot(obj, group.by = "seurat_annotations", label = TRUE, pt.size = 0.8) +
  ggtitle("Original") + NoLegend()
p2 <- DimPlot(loaded, group.by = "seurat_annotations", label = TRUE, pt.size = 0.8) +
  ggtitle("After h5ad roundtrip") + NoLegend()
p1 + p2
```

## What is preserved vs. what needs separate handling

For real scATAC-seq data, scConvert preserves the core data but some
Signac-specific components require separate handling:

### Preserved by scConvert

| Component | Seurat/Signac | h5ad |
|-----------|--------------|------|
| Peak count matrix | `counts` layer | `X` |
| Cell metadata | `meta.data` | `obs` |
| Peak metadata | `meta.features` | `var` |
| LSI/PCA embeddings | `reductions` | `obsm` |
| UMAP coordinates | `reductions$umap` | `obsm['X_umap']` |
| Cluster labels | `meta.data$seurat_clusters` | `obs['seurat_clusters']` |

### Needs separate handling

| Component | Why | Workaround |
|-----------|-----|------------|
| Fragment files | Large tabix files, not part of h5ad | Copy `fragments.tsv.gz` + `.tbi` separately |
| Gene annotations | GRanges object (R-specific) | Reload from EnsDb after import |
| Motif matrices | Signac-specific slot | Recompute with `AddMotifs()` |
| ChromatinAssay class | R S4 class, not in h5ad | Upgrade with `CreateChromatinAssay()` |

## Upgrading to ChromatinAssay (if Signac is available)

After loading an h5ad file containing peak data, you can upgrade the standard
Seurat assay to a Signac `ChromatinAssay` for peak-aware analyses.

```{r signac-upgrade, eval = FALSE}
library(Signac)

# Load h5ad containing peak-by-cell matrix
atac <- readH5AD("peaks.h5ad")

# Upgrade to ChromatinAssay
peak_counts <- GetAssayData(atac, layer = "counts")
atac[["peaks"]] <- CreateChromatinAssay(
  counts = peak_counts,
  sep = c("-", "-"),
  min.cells = 0,
  min.features = 0
)

# Re-attach annotations if needed
# library(EnsDb.Hsapiens.v86)
# Annotation(atac) <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)
# Fragments(atac) <- CreateFragmentObject("fragments.tsv.gz")
```

## Python interoperability

The exported h5ad is directly readable by scanpy or episcanpy.

```{python python-load, eval = FALSE}
# Requires Python: pip install scanpy
import scanpy as sc

adata = sc.read_h5ad("peaks.h5ad")
print(adata)
print(f"Peak names: {list(adata.var_names[:5])}")

# Standard scanpy pipeline works on peak matrices
sc.pp.normalize_total(adata)
sc.pp.log1p(adata)
sc.pp.pca(adata)
sc.pp.neighbors(adata)
sc.tl.umap(adata)
sc.pl.umap(adata, color="seurat_clusters")
```

## Clean up

```{r cleanup, eval = TRUE}
unlink(h5ad_path)
```

## Session Info

```{r session-info}
sessionInfo()
```
