---
title: "Spatial Transcriptomics: Visium"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Transcriptomics: Visium}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,

  warning = FALSE,
  out.width = "100%",
  dev = "png",
  fig.width = 8,
  fig.height = 6
)
has_SeuratData <- requireNamespace("SeuratData", quietly = TRUE)
has_data <- FALSE
if (has_SeuratData) {
  has_data <- tryCatch({
    library(SeuratData)
    "stxBrain.SeuratData" %in% rownames(InstalledData()) &&
      InstalledData()["stxBrain.SeuratData", "Installed"]
  }, error = function(e) FALSE)
}
can_run <- has_data
has_scanpy <- FALSE
has_squidpy <- FALSE
if (requireNamespace("reticulate", quietly = TRUE)) {
  scverse_path <- "/opt/homebrew/Caskroom/miniconda/base/envs/scverse"
  if (dir.exists(scverse_path)) {
    reticulate::use_condaenv("scverse", required = FALSE)
    Sys.setenv(
      OMP_NUM_THREADS = "1", OPENBLAS_NUM_THREADS = "1",
      MKL_NUM_THREADS = "1", NUMBA_THREADING_LAYER = "tbb"
    )
    has_scanpy <- reticulate::py_module_available("scanpy")
    has_squidpy <- reticulate::py_module_available("squidpy")
    if (has_scanpy) {
      reticulate::py_run_string("import matplotlib; matplotlib.use('Agg')")
    }
  }
}
```

## Overview

This vignette demonstrates how scConvert handles **10x Genomics Visium** spatial transcriptomics data, including tissue images, scale factors, and spot coordinates. Visium data is stored in h5ad files with spatial-specific fields (`obsm/spatial`, `uns/spatial`) that scConvert automatically reconstructs into Seurat `VisiumV2` image objects.

We show that the exported h5ad is **natively usable** in scanpy and squidpy (spatial scatter plots, Moran's I), and that all key components — dimensions, metadata, barcodes, features, spatial coordinates, scale factors, and dimensionality reductions — are preserved.

```{r packages}
library(Seurat)
library(scConvert)
library(ggplot2)
```

## Load Visium data

We use the **stxBrain** dataset (mouse brain anterior1, ~2,700 spots x ~31,000 genes):

```{r load_brain, eval = can_run}
library(SeuratData)
brain <- UpdateSeuratObject(LoadData("stxBrain", type = "anterior1"))
brain <- NormalizeData(brain)

cat("Cells:", ncol(brain), "\n")
cat("Genes:", nrow(brain), "\n")
cat("Images:", Images(brain), "\n")
cat("Layers:", paste(Layers(brain), collapse = ", "), "\n")
cat("Metadata columns:", ncol(brain[[]]), "\n")
```

### Visualize the original spatial data

```{r plot_orig_spatial, eval = can_run, fig.width = 10, fig.height = 8}
SpatialFeaturePlot(brain, features = "Hpca", pt.size.factor = 1.5)
```

Hpca (Hippocalcin) is a marker for hippocampal neurons, showing strong regional expression.

## Export to h5ad

scConvert writes Visium spatial data to h5ad using scanpy/squidpy conventions:

| Seurat Component | h5ad Location | Description |
|-----------------|---------------|-------------|
| Tissue coordinates | `obsm/spatial` | Spot XY positions |
| Scale factors | `uns/spatial/{lib}/scalefactors` | `spot_diameter_fullres`, `tissue_hires_scalef`, etc. |
| Tissue image (lowres) | `uns/spatial/{lib}/images/lowres` | Low-resolution H&E image |
| Tissue image (hires) | `uns/spatial/{lib}/images/hires` | High-resolution H&E image |

```{r export_h5ad, eval = can_run}
h5ad_path <- tempfile(fileext = ".h5ad")
SeuratToH5AD(brain, h5ad_path, overwrite = TRUE)
cat("File size:", round(file.size(h5ad_path) / 1024^2, 1), "MB\n")
```

## Validate in Python with scanpy

The exported h5ad is immediately usable in the scverse ecosystem — no extra configuration needed:

```{python validate_scanpy, eval = can_run && has_scanpy}
import scanpy as sc
import numpy as np

adata = sc.read_h5ad(r.h5ad_path)
print(adata)
print(f"\nobsm keys: {list(adata.obsm.keys())}")
print(f"Spatial coords shape: {adata.obsm['spatial'].shape}")
print(f"Spatial coord range X: [{adata.obsm['spatial'][:,0].min():.1f}, {adata.obsm['spatial'][:,0].max():.1f}]")
print(f"Spatial coord range Y: [{adata.obsm['spatial'][:,1].min():.1f}, {adata.obsm['spatial'][:,1].max():.1f}]")
print(f"No NaN in coords: {not np.isnan(adata.obsm['spatial']).any()}")

# Verify uns/spatial structure matches scanpy convention
if "spatial" in adata.uns:
    lib_ids = list(adata.uns["spatial"].keys())
    print(f"\nSpatial library IDs: {lib_ids}")
    for lib_id in lib_ids:
        lib = adata.uns["spatial"][lib_id]
        if "images" in lib:
            for k, v in lib["images"].items():
                print(f"  Image '{k}': shape {v.shape}, dtype {v.dtype}")
        if "scalefactors" in lib:
            sf = lib["scalefactors"]
            print(f"  Scale factors: {sf}")
            # Verify standard keys exist
            for key in ["spot_diameter_fullres", "tissue_hires_scalef", "tissue_lowres_scalef"]:
                print(f"    {key}: {'present' if key in sf else 'MISSING'}")
```

### Spatial scatter plot with squidpy

Because scConvert writes scale factors with the correct scanpy-convention names (`spot_diameter_fullres`, `tissue_hires_scalef`, etc.), squidpy plotting works out of the box:

```{python squidpy_scatter, eval = can_run && has_squidpy}
import squidpy as sq

sq.pl.spatial_scatter(adata, color="Hpca", library_id=list(adata.uns["spatial"].keys())[0],
                      img_res_key="lowres", size=1.5, alpha=0.7,
                      title="Hpca — squidpy spatial_scatter (from scConvert h5ad)")
```

### Scanpy spatial plot (coordinate-based)

```{python scanpy_spatial_plot, eval = can_run && has_scanpy}
import matplotlib.pyplot as plt

coords = adata.obsm["spatial"]
expr = adata[:, "Hpca"].X.toarray().flatten() if hasattr(adata[:, "Hpca"].X, 'toarray') else adata[:, "Hpca"].X.flatten()

fig, ax = plt.subplots(1, 1, figsize=(10, 8))
sc_plot = ax.scatter(coords[:, 0], -coords[:, 1], c=expr, s=8, cmap="viridis", alpha=0.8)
plt.colorbar(sc_plot, ax=ax, label="Hpca expression")
ax.set_title("Hpca — scanpy/matplotlib (from scConvert h5ad)")
ax.set_aspect("equal")
ax.axis("off")
plt.show()
```

### Spatial analysis: Moran's I

```{python squidpy_morans, eval = can_run && has_squidpy}
import squidpy as sq

sq.gr.spatial_neighbors(adata, coord_type="generic")
genes = list(adata.var_names[:5])
sq.gr.spatial_autocorr(adata, mode="moran", genes=genes)
print("Moran's I results:")
print(adata.uns["moranI"].head())
```

## Load h5ad back into Seurat

```{r load_h5ad, eval = can_run}
brain_rt <- LoadH5AD(h5ad_path, verbose = FALSE)

cat("Dimensions:", ncol(brain_rt), "cells x", nrow(brain_rt), "genes\n")
cat("Images:", paste(Images(brain_rt), collapse = ", "), "\n")
cat("Dims match:", ncol(brain) == ncol(brain_rt) && nrow(brain) == nrow(brain_rt), "\n")
```

## Known limitations

- **Image reconstruction**: Tissue images stored in h5ad may require specific Seurat version compatibility for full `SpatialFeaturePlot()` support. Coordinates and expression values are always preserved.
- **Non-Visium spatial**: For Slide-seq, MERFISH, Xenium, and other technologies, see the [Spatial Technologies](spatial-technologies.html) vignette.
- **Large images**: High-resolution images increase file size substantially. Consider using `lowres` images for exploratory analysis.

## Session Info

```{r sessionInfo}
sessionInfo()
```

```{r cleanup, echo = FALSE, results = 'hide'}
unlink(h5ad_path)
```
