---
title: "Visium Spatial Data Conversion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visium Spatial Data Conversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

## Introduction

10x Genomics Visium captures spatially-resolved gene expression across tissue
sections. When converting Visium data between Seurat and h5ad, scConvert
preserves spot coordinates, tissue images, scale factors, and all associated
metadata so the data is immediately usable in both R (Seurat) and Python
(scanpy/squidpy) ecosystems.

## Load the demo Visium dataset

The shipped demo contains 400 mouse brain Visium spots with 1,500 genes,
PCA/UMAP reductions, and 15 clusters.

```{r load-data, eval = TRUE}
spatial_path <- system.file("extdata", "spatial_demo.rds", package = "scConvert")
brain <- readRDS(spatial_path)

cat("Spots:", ncol(brain), "\n")
cat("Genes:", nrow(brain), "\n")
cat("Image:", Images(brain), "\n")
cat("Clusters:", nlevels(brain$seurat_clusters), "\n")
```

### Visualize a brain marker gene

Hpca (Hippocalcin) marks hippocampal neurons and shows clear regional
expression in the mouse brain.

```{r plot-original, eval = TRUE, fig.width = 7, fig.height = 5}
SpatialFeaturePlot(brain, features = "Hpca", pt.size.factor = 1.6) +
  ggplot2::ggtitle("Hpca expression (original)")
```

## Convert to h5ad

Write the Visium object to h5ad format. scConvert automatically exports spot
coordinates to `obsm/spatial`, tissue images and scale factors to
`uns/spatial`, and all metadata to `obs`.

```{r write-h5ad, eval = TRUE}
h5ad_file <- tempfile(fileext = ".h5ad")
writeH5AD(brain, h5ad_file, overwrite = TRUE)
cat("Wrote:", round(file.size(h5ad_file) / 1024^2, 1), "MB\n")
```

## Read back into Seurat

```{r read-h5ad, eval = TRUE}
brain_rt <- readH5AD(h5ad_file, verbose = FALSE)

cat("Roundtrip spots:", ncol(brain_rt), "\n")
cat("Roundtrip genes:", nrow(brain_rt), "\n")
cat("Image:", Images(brain_rt), "\n")
```

## Compare original and roundtrip

The spatial expression pattern is preserved through the h5ad roundtrip.

```{r plot-roundtrip, eval = TRUE, fig.width = 7, fig.height = 5}
SpatialFeaturePlot(brain_rt, features = "Hpca", pt.size.factor = 1.6) +
  ggplot2::ggtitle("Hpca expression (after h5ad roundtrip)")
```

Cluster assignments also survive conversion:

```{r plot-clusters, eval = TRUE, fig.width = 7, fig.height = 5}
DimPlot(brain_rt, reduction = "umap", group.by = "seurat_clusters",
        label = TRUE, repel = TRUE) +
  ggplot2::ggtitle("UMAP clusters (after h5ad roundtrip)")
```

### Verify data integrity

```{r verify, eval = TRUE}
cat("Dimensions match:",
    ncol(brain) == ncol(brain_rt) && nrow(brain) == nrow(brain_rt), "\n")
cat("Barcodes match:", all(colnames(brain) == colnames(brain_rt)), "\n")
cat("Clusters match:",
    all(as.character(brain$seurat_clusters) ==
        as.character(brain_rt$seurat_clusters)), "\n")
```

## Use in Python (scanpy/squidpy)

The exported h5ad works directly with scanpy and squidpy. Images, coordinates,
and scale factors use standard scanpy conventions.

```{python python-spatial, eval = FALSE}
# Requires Python with scanpy and squidpy installed
import scanpy as sc

adata = sc.read_h5ad("brain.h5ad")
print(adata)
print(f"Spatial coords shape: {adata.obsm['spatial'].shape}")

# Spatial scatter plot
sc.pl.spatial(adata, color="seurat_clusters")

# Squidpy spatial analysis
import squidpy as sq
sq.gr.spatial_neighbors(adata, coord_type="generic")
sq.gr.spatial_autocorr(adata, mode="moran", genes=["Hpca", "Ttr"])
```

## Cleanup

```{r cleanup, eval = TRUE, echo = FALSE, results = "hide"}
unlink(h5ad_file)
```

## Session Info

```{r session-info}
sessionInfo()
```
