---
title: "Spatial Technologies Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Technologies Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

## Introduction

Spatial transcriptomics spans a range of technologies -- from spot-based
platforms like Visium and Slide-seq to subcellular-resolution methods like
Xenium and MERFISH. scConvert handles spatial data from any platform that
produces a Seurat object or h5ad file, preserving coordinates, images, and
metadata through format conversion.

## Supported spatial technologies

| Technology | Resolution | Coordinates | Images | h5ad roundtrip |
|------------|-----------|-------------|--------|----------------|
| **10x Visium** | ~55 um spots | Gridded | H&E tissue | Full support |
| **10x Visium HD** | 2--8 um bins | Dense grid | H&E tissue | Coordinates + image |
| **Slide-seq v2** | ~10 um beads | Continuous | None | Coordinates |
| **10x Xenium** | Subcellular | Molecule-based | DAPI/IF | Coordinates via FOV |
| **MERFISH (Vizgen)** | Subcellular | Molecule-based | DAPI | Coordinates via FOV |
| **CosMx (NanoString)** | Subcellular | Molecule-based | Morphology | Coordinates via FOV |
| **CODEX (Akoya)** | Cell-level | Centroids | Fluorescence | Coordinates |
| **Stereo-seq (BGI)** | ~0.5 um | Continuous | H&E | Coordinates |

All technologies store spatial coordinates in the h5ad `obsm/spatial` field.
Visium additionally stores tissue images and scale factors in `uns/spatial`.

## Visium example: mouse brain

We demonstrate a full roundtrip with the shipped Visium demo dataset (400 mouse
brain spots, 1,500 genes, 15 clusters).

```{r load-data, eval = TRUE}
spatial_path <- system.file("extdata", "spatial_demo.rds", package = "scConvert")
brain <- readRDS(spatial_path)

cat("Spots:", ncol(brain), "\n")
cat("Genes:", nrow(brain), "\n")
cat("Image:", paste(names(brain@images), collapse = ", "), "\n")
cat("Assay:", paste(names(brain@assays), collapse = ", "), "\n")
```

### Spatial gene expression

Ttr (Transthyretin) is expressed in the choroid plexus and shows strong spatial
localization.

```{r plot-spatial, eval = TRUE, fig.width = 7, fig.height = 5}
SpatialFeaturePlot(brain, features = "Ttr", pt.size.factor = 1.6) +
  ggplot2::ggtitle("Ttr expression (original)")
```

### Convert and load back

```{r roundtrip, eval = TRUE}
h5ad_file <- tempfile(fileext = ".h5ad")
writeH5AD(brain, h5ad_file, overwrite = TRUE)
brain_rt <- readH5AD(h5ad_file, verbose = FALSE)

cat("Roundtrip spots:", ncol(brain_rt), "\n")
cat("Roundtrip genes:", nrow(brain_rt), "\n")
cat("Image preserved:", length(brain_rt@images) > 0, "\n")
```

### Verify coordinate preservation

```{r check-coords, eval = TRUE}
cat("Barcodes match:", all(colnames(brain) == colnames(brain_rt)), "\n")
cat("Features match:", all(rownames(brain) == rownames(brain_rt)), "\n")
cat("Clusters match:",
    all(as.character(brain$seurat_clusters) ==
        as.character(brain_rt$seurat_clusters)), "\n")
```

### Clusters preserved through roundtrip

```{r plot-clusters, eval = TRUE, fig.width = 7, fig.height = 5}
DimPlot(brain_rt, reduction = "umap", group.by = "seurat_clusters",
        label = TRUE, repel = TRUE) +
  ggplot2::ggtitle("UMAP clusters (after h5ad roundtrip)")
```

### Another gene: Nrgn

Nrgn (Neurogranin) is a cortical neuron marker with a different spatial pattern
than Ttr.

```{r plot-nrgn, eval = TRUE, fig.width = 7, fig.height = 5}
FeaturePlot(brain_rt, features = "Nrgn", reduction = "umap") +
  ggplot2::ggtitle("Nrgn expression (UMAP, after roundtrip)")
```

## Working with other spatial platforms

For any h5ad file with spatial coordinates, the same workflow applies:

```r
# Load spatial h5ad from any platform
obj <- readH5AD("merfish_data.h5ad")

# Check what was detected
cat("Images:", paste(names(obj@images), collapse = ", "), "\n")

# Convert to another format
scConvert("merfish_data.h5ad", "merfish_data.h5seurat")
```

Technologies without tissue images (Slide-seq, MERFISH, CODEX) store only
coordinates in `obsm/spatial`. scConvert reads these into the Seurat object and
makes them available for plotting and downstream analysis.

## Python interop

```{python python-example, eval = FALSE}
# Requires Python with scanpy installed
import scanpy as sc

adata = sc.read_h5ad("brain.h5ad")
print(adata)

# Works with any spatial technology
sc.pl.spatial(adata, color="Ttr")  # Visium (with image)
sc.pl.embedding(adata, basis="spatial", color="cluster")  # Generic spatial
```

## Cleanup

```{r cleanup, eval = TRUE, echo = FALSE, results = "hide"}
unlink(h5ad_file)
```

## Session Info

```{r session-info}
sessionInfo()
```
