---
title: "Visium Spatial Data Conversion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visium Spatial Data Conversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(scConvert)
library(Seurat)
library(ggplot2)
```

## Introduction

10x Genomics Visium captures spatially-resolved gene expression across tissue
sections. When converting Visium data between Seurat and h5ad, scConvert
preserves spot coordinates, tissue images, scale factors, and all associated
metadata so the data is immediately usable in both R (Seurat) and Python
(scanpy/squidpy) ecosystems.

## Read spatial data from h5ad

The shipped `spatial_demo.h5ad` contains 400 mouse brain Visium spots with
1,500 genes, PCA/UMAP reductions, and cluster labels. `readH5AD()` rebuilds
the full Seurat spatial object, including the tissue image.

```{r read-h5ad, eval = TRUE}
h5ad_file <- system.file("extdata", "spatial_demo.h5ad", package = "scConvert")
brain <- readH5AD(h5ad_file, verbose = FALSE)

cat("Spots:", ncol(brain), "\n")
cat("Genes:", nrow(brain), "\n")
cat("Image:", Images(brain), "\n")
cat("Assay:", Assays(brain), "\n")
```

### Spatial expression of a brain marker

Hpca (Hippocalcin) marks hippocampal neurons and shows clear regional
expression in the mouse brain.

```{r spatialplot-hpca, eval = TRUE, fig.width = 7, fig.height = 5}
SpatialFeaturePlot(brain, features = "Hpca", pt.size.factor = 1.6) +
  ggtitle("Hpca expression (loaded from h5ad)")
```

### Cluster overview

```{r dimplot-clusters, eval = TRUE, fig.width = 7, fig.height = 5}
DimPlot(brain, reduction = "umap", group.by = "seurat_clusters",
        label = TRUE, repel = TRUE) +
  ggtitle("UMAP clusters (loaded from h5ad)")
```

## Write-and-read roundtrip

Write the spatial object to a new h5ad file, read it back, and confirm that
coordinates, images, scale factors, and expression values are all preserved.

```{r write-roundtrip, eval = TRUE}
h5ad_rt <- file.path(tempdir(), "spatial_roundtrip.h5ad")
writeH5AD(brain, h5ad_rt, overwrite = TRUE)
cat("Wrote:", round(file.size(h5ad_rt) / 1024^2, 1), "MB\n")

brain_rt <- readH5AD(h5ad_rt, verbose = FALSE)
cat("Roundtrip spots:", ncol(brain_rt), "\n")
cat("Roundtrip genes:", nrow(brain_rt), "\n")
cat("Image:", Images(brain_rt), "\n")
```

### Compare spatial plots

The spatial expression pattern should be identical after roundtrip.

```{r spatialplot-roundtrip, eval = TRUE, fig.width = 7, fig.height = 5}
SpatialFeaturePlot(brain_rt, features = "Hpca", pt.size.factor = 1.6) +
  ggtitle("Hpca expression (after h5ad roundtrip)")
```

### Verify data integrity

```{r verify, eval = TRUE}
cat("Dimensions match:",
    ncol(brain) == ncol(brain_rt) && nrow(brain) == nrow(brain_rt), "\n")
cat("Barcodes match:", all(colnames(brain) == colnames(brain_rt)), "\n")
cat("Clusters match:",
    all(as.character(brain$seurat_clusters) ==
        as.character(brain_rt$seurat_clusters)), "\n")
```

## What gets preserved

scConvert maintains full fidelity for Visium spatial data during conversion:

| Component | h5ad location | Preserved |
|-----------|---------------|-----------|
| Expression matrix | `X` / `raw/X` | Yes |
| Spot coordinates | `obsm/spatial` | Yes |
| Tissue image | `uns/spatial/*/images` | Yes |
| Scale factors | `uns/spatial/*/scalefactors` | Yes |
| Cell metadata | `obs` | Yes |
| Reductions (PCA, UMAP) | `obsm` | Yes |

## Use in Python (scanpy/squidpy)

The exported h5ad works directly with scanpy and squidpy. Images, coordinates,
and scale factors use standard scanpy conventions.

```{python python-spatial, eval = FALSE}
# Requires Python with scanpy and squidpy installed
import scanpy as sc

adata = sc.read_h5ad("spatial_demo.h5ad")
print(adata)
print(f"Spatial coords shape: {adata.obsm['spatial'].shape}")

# Spatial scatter plot
sc.pl.spatial(adata, color="seurat_clusters")

# Squidpy spatial analysis
import squidpy as sq
sq.gr.spatial_neighbors(adata, coord_type="generic")
sq.gr.spatial_autocorr(adata, mode="moran", genes=["Hpca", "Ttr"])
```

## Clean up

```{r cleanup, eval = TRUE, echo = FALSE, results = "hide"}
unlink(h5ad_rt)
```

## Session Info

```{r session-info}
sessionInfo()
```
