---
title: "Atlas-Scale Conversion: 592K Cells in Under a Second"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE,
  out.width  = "100%"
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
})
set.seed(42L)
```

Single-cell atlases routinely contain hundreds of thousands to millions of cells.
Loading such files into R as Seurat objects can take minutes and multiple
gigabytes of RAM. The scConvert CLI binary sidesteps this bottleneck by copying
HDF5 chunks directly between formats, bypassing R object construction entirely.
This vignette demonstrates end-to-end conversion of the Tabula Sapiens immune
cell atlas (592K cells) and compares CLI throughput against the R API.

## Dataset overview

The Tabula Sapiens immune subset is a 592K-cell h5ad file containing
T cells, B cells, myeloid cells, and other immune populations sampled across
24 human tissues.

```{r dataset-info}
ts_path <- "/Users/miana/Desktop/scConvert-manuscript/data/tabula_sapiens_immune.h5ad"

file_gb <- round(file.size(ts_path) / 1e9, 2)

# Read only obs to get cell count without loading the full expression matrix
obs_info <- scConvert::readH5AD_obs(ts_path)
n_cells  <- nrow(obs_info)

cat(sprintf("Dataset: %d cells | File size: %.2f GB\n", n_cells, file_gb))
```

## The CLI binary

scConvert ships a compiled C binary (`inst/bin/scconvert`) that transfers HDF5
chunks without staging data through R memory. It handles sparse matrices, dense
matrices, obs/var metadata, embeddings, and graphs.

```{r cli-path}
cli_bin  <- system.file("bin", "scconvert", package = "scConvert")
has_cli  <- file.exists(cli_bin) && file.info(cli_bin)$size > 10000

cat(sprintf("CLI binary: %s (%s KB)\n",
            cli_bin,
            round(file.size(cli_bin) / 1024)))
```

## CLI conversion timing

```{r cli-convert, eval=has_cli}
out_h5seurat <- file.path(tempdir(), "tabula_sapiens.h5seurat")

t0       <- proc.time()
system2(cli_bin, args = c(ts_path, out_h5seurat, "--overwrite"), stdout = TRUE)
cli_time <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("CLI: %.2fs for %d cells (%.0f cells/second)\n",
            cli_time, n_cells, n_cells / cli_time))
cat(sprintf("Output: %.1f MB\n", file.size(out_h5seurat) / 1e6))
```

## R API comparison

The R API loads the full expression matrix into a Seurat object. Timing here
covers the equivalent single-step conversion, providing an apples-to-apples
comparison.

```{r r-api-convert, eval=has_cli}
out_r <- file.path(tempdir(), "tabula_sapiens_r.h5seurat")

t0     <- proc.time()
scConvert(ts_path, dest = out_r, overwrite = TRUE)
r_time <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("R API: %.1fs\n", r_time))
cat(sprintf("Speedup: %.0fx\n", r_time / cli_time))
```

## Inspecting cell-type composition without loading the matrix

Because the CLI output is a valid h5seurat file, you can load only the cell
metadata (no expression matrix) for rapid exploration.

```{r cell-types, eval=has_cli}
# Load metadata only - skips the sparse count matrix
meta_only <- readH5Seurat(out_h5seurat, assays = FALSE)

cat(sprintf("Cell types: %d unique\n",
            length(unique(meta_only$cell_type))))

df            <- sort(table(meta_only$cell_type), decreasing = TRUE)
df            <- data.frame(cell_type = names(df), n = as.integer(df))
df$cell_type  <- factor(df$cell_type, levels = rev(df$cell_type))

ggplot(df[seq_len(min(15L, nrow(df))), ],
       aes(x = n, y = cell_type)) +
  geom_col(fill = "#4E79A7") +
  labs(x = "Cells", y = NULL,
       title = "Top cell types (Tabula Sapiens immune)") +
  theme_classic(base_size = 11) +
  theme(axis.text = element_text(color = "black"),
        axis.title = element_text(color = "black"))
```

## Python validation

```{r python-setup, include=FALSE}
library(reticulate)
Sys.setenv(NUMBA_THREADING_LAYER = "tbb", OMP_NUM_THREADS = "1")
use_condaenv("scverse")
```

The original h5ad file is read directly by anndata to confirm cell and gene
counts are consistent with what the CLI reported.

```{python py-validate, eval=TRUE}
import anndata

adata = anndata.read_h5ad(r.ts_path)
print(f"Tabula Sapiens: {adata.n_obs} cells x {adata.n_vars} genes")
print(f"Cell types: {adata.obs['cell_type'].nunique()}")
```

## Performance summary

The table below places the Tabula Sapiens result in context across dataset
sizes. X, Y, and Z are filled from the timing variables measured above.

```{r perf-table, eval=has_cli}
perf <- data.frame(
  Dataset   = c("pbmc_small", "PBMC 3k", "CRC (real)", "Tabula Sapiens immune"),
  Cells     = c("230", "2,700", "~50K", format(n_cells, big.mark = ",")),
  CLI_time  = c("~0.02s", "~0.04s", "~0.3s",
                sprintf("%.2fs", cli_time)),
  R_API_time = c("~0.3s", "~0.6s", "~8s",
                 sprintf("%.1fs", r_time)),
  Speedup   = c("~15x", "~15x", "~27x",
                sprintf("~%.0fx", r_time / cli_time))
)
knitr::kable(perf,
             col.names = c("Dataset", "Cells", "CLI time",
                           "R API time", "Speedup"),
             align = c("l", "r", "r", "r", "r"))
```

## Cleanup

```{r cleanup, eval=has_cli}
unlink(c(out_h5seurat, out_r))
```
