---
title: "Seurat <-> SingleCellExperiment: Bioconductor Interoperability"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE,
  out.width  = "100%"
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
})
set.seed(42L)
has_sce <- requireNamespace("SingleCellExperiment", quietly = TRUE) &&
  requireNamespace("SummarizedExperiment", quietly = TRUE)
```

## Introduction

Many Bioconductor workflows -- scran normalization, scater QC,
slingshot trajectory inference, MAST differential expression -- operate on the
[SingleCellExperiment](https://bioconductor.org/packages/SingleCellExperiment/)
(SCE) container. Seurat provides built-in converters (`as.SingleCellExperiment()`
and `as.Seurat()`) that move data between ecosystems. scConvert extends this by
letting you route SCE objects directly to h5ad, h5Seurat, Loom, or Zarr via
its hub architecture -- no manual intermediate step required.

This article walks through the full workflow on the PBMC 3k dataset.

---

## 1  Load real PBMC3k data

```{r load-pbmc}
t0      <- proc.time()
pbmc    <- readH5AD("../pbmc3k.h5ad", verbose = FALSE)
elapsed <- (proc.time() - t0)[["elapsed"]]
cat(sprintf("Loaded: %d cells x %d genes in %.2fs\n",
            ncol(pbmc), nrow(pbmc), elapsed))
```

---

## 2  Seurat analysis pipeline

We run the standard pipeline so the object carries PCA, UMAP, and cluster
labels for downstream inspection. Steps are skipped if reductions already
exist in the imported file.

```{r seurat-pipeline}
set.seed(42L)
pbmc <- NormalizeData(pbmc, verbose = FALSE)
pbmc <- FindVariableFeatures(pbmc, nfeatures = 2000L, verbose = FALSE)
pbmc <- ScaleData(pbmc, verbose = FALSE)
pbmc <- RunPCA(pbmc, npcs = 30L, verbose = FALSE)
pbmc <- RunUMAP(pbmc, dims = 1:20, verbose = FALSE)
pbmc <- FindNeighbors(pbmc, dims = 1:20, verbose = FALSE)
pbmc <- FindClusters(pbmc, resolution = 0.5, verbose = FALSE)
```

```{r dimplot, fig.cap="PBMC 3k UMAP coloured by Seurat clusters."}
DimPlot(pbmc, reduction = "umap", label = TRUE, label.size = 4) +
  ggtitle("PBMC 3k: Seurat clusters") +
  theme(plot.title = element_text(hjust = 0.5))
```

---

## 3  Seurat to SingleCellExperiment

`as.SingleCellExperiment()` transfers counts, normalized data, dimensional
reductions, and cell metadata into the SCE container.

```{r seurat-to-sce, eval=has_sce}
suppressPackageStartupMessages(library(SingleCellExperiment))

sce <- as.SingleCellExperiment(pbmc)
cat(sprintf("SCE assays: %s\n",
            paste(assayNames(sce), collapse = ", ")))
cat(sprintf("Reduced dims: %s\n",
            paste(reducedDimNames(sce), collapse = ", ")))
cat(sprintf("colData columns: %d\n", ncol(colData(sce))))
```

---

## 4  Bioconductor QC workflow on SCE

A typical Bioconductor entry point is computing per-cell QC metrics. Below we
use base R to calculate log-library-size and detected-feature counts -- the
same quantities that `scater::perCellQCMetrics()` would produce. The new
columns become available to any SCE-aware tool.

```{r sce-qc, eval=has_sce}
counts_mat <- assay(sce, "counts")
colData(sce)$log_library_size  <- log1p(Matrix::colSums(counts_mat))
colData(sce)$detected_features <- Matrix::colSums(counts_mat > 0)
cat(sprintf("Added QC columns: log_library_size, detected_features\n"))
cat(sprintf("Median library size (log1p): %.2f\n",
            median(colData(sce)$log_library_size)))
cat(sprintf("Median detected features: %.0f\n",
            median(colData(sce)$detected_features)))
```

```{r qc-violin, eval=has_sce, fig.width=10, fig.height=5, fig.cap="Per-cell QC metrics across Seurat clusters."}
qc_df <- data.frame(
  cluster         = as.character(pbmc$seurat_clusters),
  log_library     = colData(sce)$log_library_size,
  detected        = colData(sce)$detected_features
)

p1 <- ggplot(qc_df, aes(x = cluster, y = log_library, fill = cluster)) +
  geom_violin(scale = "width", trim = TRUE) +
  geom_boxplot(width = 0.1, outlier.size = 0.5, fill = "white") +
  labs(x = "Cluster", y = "log(1 + library size)") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 9))

p2 <- ggplot(qc_df, aes(x = cluster, y = detected, fill = cluster)) +
  geom_violin(scale = "width", trim = TRUE) +
  geom_boxplot(width = 0.1, outlier.size = 0.5, fill = "white") +
  labs(x = "Cluster", y = "Detected features") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 9))

suppressPackageStartupMessages(library(patchwork))
p1 + p2
```

---

## 5  SCE back to Seurat

`as.Seurat()` reconstructs the Seurat object from SCE. The `counts` and
`data` arguments map SCE assay names to the appropriate Seurat layers. Any
columns added to `colData` during the Bioconductor workflow travel with the
object.

```{r sce-to-seurat, eval=has_sce}
seurat_rt <- as.Seurat(sce, counts = "counts", data = "logcounts")
cat(sprintf("Converted back: %d cells\n", ncol(seurat_rt)))
cat(sprintf("Metadata columns: %s\n",
            paste(head(colnames(seurat_rt[[]]), 6L), collapse = ", ")))
```

---

## 6  SCE to h5ad via scConvert

To share an SCE dataset with Python collaborators, pass it directly to
`scConvert()`. The SCE is converted to a Seurat object internally, then
written to the target format -- no manual intermediate step is needed.

```{r sce-to-h5ad, eval=has_sce}
h5ad_path <- file.path(tempdir(), "pbmc3k_sce.h5ad")

t0      <- proc.time()
scConvert(sce, dest = h5ad_path, overwrite = TRUE)
elapsed <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("h5ad size: %.1f MB (%.2fs)\n",
            file.size(h5ad_path) / 1e6, elapsed))
```

---

## 7  What is preserved

### Seurat -> SCE -> Seurat round-trip

| Component | Preserved |
|-----------|-----------|
| Count matrix | Yes |
| Normalized data | Yes |
| Cell metadata | Yes |
| Gene metadata | Yes |
| PCA / UMAP | Yes (as reducedDims) |
| Neighbor graphs | Partial |

### Common Bioconductor workflows enabled by scConvert

| Workflow | Package | After scConvert |
|----------|---------|-----------------|
| QC filtering | scater | `perCellQCMetrics()` on SCE |
| Normalization | scran | `computeSumFactors()` on SCE |
| Trajectory | slingshot | `slingshot(sce, ...)` |
| Differential expression | MAST | `zlm(~ condition, sce)` |

---

## 8  Cleanup

```{r cleanup, eval=has_sce}
unlink(h5ad_path)
```
