---
title: "Loom Format: HDF5 Interop with loompy and scVelo"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE,
  out.width  = "100%"
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
})
set.seed(42L)
```

## Overview

Loom is an HDF5-based format developed by the Linnarsson lab and used natively by
[loompy](https://loompy.org), [RNA velocity tools](https://scvelo.readthedocs.io)
(scVelo, velocyto), and several single-cell atlases. It stores the expression matrix
at `/matrix`, cell metadata in `/col_attrs`, gene metadata in `/row_attrs`, and
additional layers (spliced/unspliced counts, normalized data) in `/layers`.

scConvert provides `readLoom()` and `writeLoom()` for full Seurat <-> Loom conversion
without a Python dependency.

This article demonstrates the conversion on PBMC 3k (2,638 cells, 13,714 genes) and
validates the written file with loompy and scanpy.

---

## 1  Load h5ad and write Loom

We first load the PBMC 3k h5ad into Seurat, then export to Loom.

```{r read-h5ad}
input_h5ad <- "../pbmc3k.h5ad"
pbmc_seurat <- readH5AD(input_h5ad, verbose = FALSE)
cat(sprintf("Loaded: %d cells x %d genes\n", ncol(pbmc_seurat), nrow(pbmc_seurat)))
```

```{r write-loom}
loom_path <- file.path(tempdir(), "pbmc3k.loom")

t0 <- proc.time()
writeLoom(pbmc_seurat, filename = loom_path, overwrite = TRUE, verbose = FALSE)
elapsed <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("Wrote Loom: %.2fs | %.1f MB\n", elapsed, file.size(loom_path) / 1e6))
```

---

## 2  Read Loom back into Seurat

```{r read-loom}
pbmc_loom <- readLoom(loom_path, verbose = FALSE)
cat(sprintf("Loaded: %d cells x %d genes\n", ncol(pbmc_loom), nrow(pbmc_loom)))
```

```{r inspect-loom}
pbmc_loom
head(pbmc_loom[[]], 4)
```

```{r dimplot-loom, fig.cap="PBMC 3k UMAP coloured by cell-type annotation after Loom round-trip."}
set.seed(42L)
pbmc_loom <- NormalizeData(pbmc_loom, verbose = FALSE)
pbmc_loom <- FindVariableFeatures(pbmc_loom, nfeatures = 2000L, verbose = FALSE)
pbmc_loom <- ScaleData(pbmc_loom, verbose = FALSE)
pbmc_loom <- RunPCA(pbmc_loom, npcs = 30L, verbose = FALSE)
pbmc_loom <- RunUMAP(pbmc_loom, dims = 1:20, verbose = FALSE)

DimPlot(
  pbmc_loom,
  reduction  = "umap",
  group.by   = "seurat_annotations",
  label      = TRUE,
  label.size = 3.5,
  repel      = TRUE
) +
  ggtitle("PBMC 3k: cell-type annotations (from Loom)") +
  theme(plot.title = element_text(hjust = 0.5))
```

---

## 3  What is preserved

| Component | Preserved | Loom path |
|-----------|:---------:|-----------|
| Expression matrix | Yes | `/matrix` |
| Raw counts | Yes | `/layers/counts` |
| Cell metadata | Yes | `/col_attrs` |
| Gene metadata | Yes | `/row_attrs` |
| PCA / UMAP embeddings | Yes | `/col_attrs/PC_1..n`, `UMAP_1..n` |
| Seurat cluster labels | Yes | `/col_attrs/seurat_clusters` |
| Neighbor graphs | No | Recompute with `FindNeighbors()` |
| `misc` / `uns` | No | Not part of the Loom spec |

---

## 4  Python validation with loompy

```{r python-setup}
library(reticulate)
Sys.setenv(NUMBA_THREADING_LAYER = "tbb", OMP_NUM_THREADS = "1")
use_condaenv("scverse")
```

```{python py-loompy, eval=TRUE}
import loompy
import matplotlib
matplotlib.use('Agg')
import matplotlib.pyplot as plt

with loompy.connect(r.loom_path, mode="r") as ds:
    print(f"Shape: {ds.shape[0]} genes x {ds.shape[1]} cells")
    print(f"Column attributes (cell metadata): {list(ds.ca.keys())[:8]}")
    print(f"Row attributes (gene metadata): {list(ds.ra.keys())[:5]}")
    print(f"Layers: {list(ds.layers.keys())}")
```

---

## 5  Read Loom with scanpy

scanpy can load a Loom file directly and treat it as an AnnData object.

```{python py-scanpy-loom, eval=TRUE}
import scanpy as sc
adata = sc.read_loom(r.loom_path, sparse=True, cleanup=False)
print(adata)
print(f"obs columns: {list(adata.obs.columns)[:6]}")
```

---

## 6  scVelo use case

Loom is the primary input format for RNA velocity analysis with
[scVelo](https://scvelo.readthedocs.io) and
[velocyto](https://velocyto.org). scVelo expects two layers: `spliced` and
`unspliced`, which are produced by velocyto or STARsolo during alignment.

A typical workflow:

1. Run STARsolo or velocyto on the BAM files to produce a `.loom` with
   spliced/unspliced counts.
2. Load that loom into R with `readLoom()` for QC and metadata annotation in Seurat.
3. Export the annotated object back with `writeLoom()`, which preserves
   the cluster labels and embeddings in `/col_attrs`.
4. Pass the annotated loom to scVelo in Python.

scConvert writes expression counts to `/layers/counts`. Users need to merge the
spliced/unspliced layers from the velocyto output separately before running scVelo.

```{python py-scvelo-note, eval=TRUE}
import loompy
with loompy.connect(r.loom_path, mode="r") as ds:
    layers = list(ds.layers.keys())
    print(f"Layers in scConvert output: {layers}")
    print("scVelo additionally requires: ['spliced', 'unspliced']")
    print("Merge these from the velocyto loom before calling scvelo.pp.filter_and_normalize()")
```

---

## 7  Cleanup

```{r cleanup}
unlink(loom_path)
```
