---
title: "Zarr Format: Cloud-Native Storage with Python zarr Validation"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE,
  out.width  = "100%"
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
})
set.seed(42L)
```

## Overview

Zarr is a directory-based format for chunked compressed arrays. Unlike HDF5, every chunk
is stored as an independent file, making Zarr cloud-native: individual chunks can be read
directly from S3, GCS, or Azure Blob Storage without downloading the full dataset.
Zarr is the storage format used by
[CELLxGENE](https://cellxgene.cziscience.com),
[Human Cell Atlas](https://www.humancellatlas.org), and
[scverse](https://scverse.org) streaming pipelines.

scConvert reads and writes Zarr v2 stores following the AnnData on-disk specification.
Three conversion paths are available:

- `scConvert()` dispatcher (hub path through Seurat)
- `readZarr()` / `writeZarr()` (load/save from R)
- `H5ADToZarr()` / `ZarrToH5AD()` (streaming, no Seurat intermediate)

This article demonstrates all three paths on the PBMC 3k dataset (2,638 cells,
13,714 genes) and validates the output with Python zarr and anndata.

---

## 1  h5ad to Zarr

`scConvert()` detects output format from the file extension and routes through the Seurat
hub. This path preserves all obs, var, obsm, obsp, and uns fields.

```{r h5ad-to-zarr}
input_h5ad <- "../pbmc3k.h5ad"
zarr_path  <- file.path(tempdir(), "pbmc3k.zarr")

t0 <- proc.time()
scConvert(input_h5ad, dest = zarr_path, overwrite = TRUE, verbose = FALSE)
elapsed <- (proc.time() - t0)[["elapsed"]]

n_files <- length(list.files(zarr_path, recursive = TRUE))
cat(sprintf("Converted to Zarr: %.2fs | %d chunk files\n", elapsed, n_files))
```

Each chunk file in the Zarr store corresponds to a tile of a compressed array. The
number of files reflects the chunking strategy applied to X, obsm embeddings, and
metadata arrays.

---

## 2  Read Zarr into Seurat

`readZarr()` reconstructs a Seurat object from the store, mapping AnnData conventions
back to Seurat slots (X->data layer, raw/X->counts layer, obsm->reductions, obsp->graphs).

```{r read-zarr}
pbmc <- readZarr(zarr_path, verbose = FALSE)
cat(sprintf("Loaded: %d cells x %d genes\n", ncol(pbmc), nrow(pbmc)))
cat(sprintf("Reductions: %s\n", paste(names(pbmc@reductions), collapse = ", ")))
```

```{r inspect-zarr}
pbmc
head(pbmc[[]], 4)
```

```{r dimplot-zarr, fig.cap="PBMC 3k UMAP coloured by cell-type annotation after Zarr round-trip."}
DimPlot(
  pbmc,
  reduction = "umap",
  group.by  = "seurat_annotations",
  label     = TRUE,
  label.size = 3.5,
  repel     = TRUE
) +
  ggtitle("PBMC 3k: cell-type annotations (from Zarr)") +
  theme(plot.title = element_text(hjust = 0.5))
```

---

## 3  Python validation

We confirm the Zarr store is readable with Python zarr and anndata.

```{r python-setup}
library(reticulate)
Sys.setenv(NUMBA_THREADING_LAYER = "tbb", OMP_NUM_THREADS = "1")
use_condaenv("scverse")
```

```{python py-zarr-inspect, eval=TRUE}
import zarr
import anndata
import matplotlib
matplotlib.use('Agg')
import matplotlib.pyplot as plt

store = zarr.open(r.zarr_path, mode="r")
print("Top-level Zarr groups:", list(store.keys()))
```

```{python py-anndata-read, eval=TRUE}
adata = anndata.read_zarr(r.zarr_path)
print(adata)
print(f"Validated: {adata.n_obs} cells x {adata.n_vars} genes")
print(f"Obs columns: {list(adata.obs.columns)[:5]}")
print(f"Obsm keys: {list(adata.obsm.keys())}")
```

anndata reads the store natively, confirming the encoding attributes, chunk
shapes, and dtype conventions are all compliant with the AnnData Zarr specification.

---

## 4  Streaming converters (no Seurat intermediate)

`H5ADToZarr()` streams fields from HDF5 directly into Zarr chunks without constructing
a Seurat object. This is the lowest-memory conversion path and is appropriate when
you need to reformat large datasets that do not fit in R memory.

```{r h5ad-to-zarr-stream}
zarr2_path <- file.path(tempdir(), "pbmc3k_stream.zarr")

t0 <- proc.time()
H5ADToZarr(input_h5ad, zarr2_path, overwrite = TRUE, verbose = FALSE)
elapsed2 <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("H5ADToZarr (streaming): %.2fs\n", elapsed2))
```

The symmetric streaming converter `ZarrToH5AD()` writes a valid h5ad from any Zarr
store, including stores downloaded from cloud sources.

---

## 5  Zarr to h5ad round-trip

```{r zarr-to-h5ad}
rt_h5ad <- file.path(tempdir(), "pbmc3k_from_zarr.h5ad")
ZarrToH5AD(zarr_path, rt_h5ad, overwrite = TRUE, verbose = FALSE)

rt <- readH5AD(rt_h5ad, verbose = FALSE)
cat(sprintf("Round-trip: %d cells x %d genes\n", ncol(rt), nrow(rt)))
```

---

## 6  Conversion path reference

| Path | Function | Seurat object built | Best for |
|------|----------|:-------------------:|---------|
| h5ad -> Zarr (hub) | `scConvert()` | Yes | Full metadata round-trip |
| h5ad -> Zarr (stream) | `H5ADToZarr()` | No | Large files, low memory |
| Zarr -> h5ad (stream) | `ZarrToH5AD()` | No | Cloud-to-HDF5 reformat |
| h5seurat -> Zarr | `H5SeuratToZarr()` | No | Seurat-native streaming |
| Zarr -> h5seurat | `ZarrToH5Seurat()` | No | Seurat-native streaming |

---

## 7  Cleanup

```{r cleanup}
unlink(zarr_path,  recursive = TRUE)
unlink(zarr2_path, recursive = TRUE)
unlink(rt_h5ad)
```
