---
title: "SpatialData Zarr: The scverse Spatial Standard"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE,
  out.width  = "100%"
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
})
set.seed(42L)
```

## Overview

[SpatialData](https://spatialdata.scverse.org) is the scverse standard for spatial
omics. A `.spatialdata.zarr` store is a Zarr directory that encodes expression tables,
spatial coordinates, shapes (polygons, circles), tissue images, and coordinate systems
in a unified hierarchy. All major spatial technologies now have SpatialData
representations via the
[spatialdata-io](https://spatialdata-io.scverse.org) package.

This article demonstrates the Python–R interop workflow: Python reads the native
SpatialData Zarr store, exports an h5ad, and scConvert loads it into Seurat for
downstream analysis.

---

## 1  Python: native SpatialData reading

The Python spatialdata library reads the store directly, preserving the full
multi-element hierarchy (tables, images, shapes, points).

```{r python-setup, include=FALSE}
library(reticulate)
Sys.setenv(NUMBA_THREADING_LAYER = "tbb", OMP_NUM_THREADS = "1")
use_condaenv("scverse")
```

```{r setup-paths}
sd_path   <- normalizePath("../blobs.spatialdata.zarr")
h5ad_path <- file.path(tempdir(), "blobs.h5ad")
```

```{python py-spatialdata, eval=TRUE}
import spatialdata
import matplotlib
matplotlib.use('Agg')
import matplotlib.pyplot as plt

sdata = spatialdata.read_zarr(r.sd_path)
print(sdata)
print(f"\nTables : {list(sdata.tables.keys())}")
print(f"Images : {list(sdata.images.keys())}")
print(f"Shapes : {list(sdata.shapes.keys())}")
print(f"Points : {list(sdata.points.keys())}")
```

---

## 2  Python: export table to h5ad

The `table` element of a SpatialData store is a standard AnnData object.
Exporting it to h5ad makes it immediately readable by scConvert.

```{python py-export-h5ad, eval=TRUE}
import os

adata = sdata.tables["table"]
print(f"Table: {adata.n_obs} cells x {adata.n_vars} features")
print(f"obs columns: {list(adata.obs.columns)}")

adata.write_h5ad(r.h5ad_path, compression="gzip")
print(f"Written: {os.path.getsize(r.h5ad_path) / 1e3:.1f} KB -> {r.h5ad_path}")
```

---

## 3  R: load h5ad with scConvert

`readH5AD()` loads the exported h5ad into a standard Seurat object. All cell
metadata and spatial coordinates are preserved.

```{r read-h5ad}
t0 <- proc.time()
blobs <- readH5AD(h5ad_path, verbose = FALSE)
elapsed <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("Loaded: %d cells x %d features in %.2fs\n",
            ncol(blobs), nrow(blobs), elapsed))
cat(sprintf("Metadata: %s\n", paste(names(blobs[[]]), collapse = ", ")))
```

```{r inspect-blobs}
head(blobs[[]], 6)
```

---

## 4  Spatial coordinates in Seurat

Spatial coordinates written by Python are stored in `obsm['spatial']` in the h5ad
and mapped to a Seurat `FOV` image when loading with `readH5AD()`. This allows
`SpatialFeaturePlot()` and related Seurat spatial functions to work without further
setup.

```{r check-coords}
cat(sprintf("Reductions: %s\n",
            paste(names(blobs@reductions), collapse = ", ")))
cat(sprintf("Images   : %s\n",
            paste(names(blobs@images), collapse = ", ")))
```

---

## 5  scConvert API reference

For zarr v2 SpatialData stores, scConvert provides direct R-to-R conversion
functions without requiring Python.

```{r direct-converters, eval=FALSE}
SpatialDataToH5AD(
  source    = "sample.spatialdata.zarr",
  dest      = "sample.h5ad",
  overwrite = TRUE
)

H5ADToSpatialData(
  source    = "sample.h5ad",
  dest      = "sample.spatialdata.zarr",
  overwrite = TRUE
)

SpatialDataToH5Seurat(
  source    = "sample.spatialdata.zarr",
  dest      = "sample.h5seurat",
  overwrite = TRUE
)

scConvert("sample.spatialdata.zarr", dest = "sample.h5ad")
```

---

## 6  Technology coverage

The following spatial technologies produce SpatialData stores via `spatialdata-io`.
scConvert interoperates with all of them through the h5ad bridge above.

| Technology | Vendor | Modality | spatialdata-io reader |
|------------|--------|----------|-----------------------|
| Visium / Visium HD | 10x Genomics | transcriptomics | `visium()` / `visium_hd()` |
| Xenium | 10x Genomics | transcriptomics (FISH) | `xenium()` |
| CosMx | NanoString | transcriptomics (FISH) | `cosmx()` |
| MERFISH | Vizgen | transcriptomics (FISH) | `merfish()` |
| Slide-seq | Broad Institute | transcriptomics | `slideseq()` |
| CODEX / PhenoCycler | Akoya Biosciences | proteomics | `codex()` |
| IMC | Fluidigm / Standard BioTools | proteomics | `imc()` |

---

## 7  Cleanup

```{r cleanup}
unlink(h5ad_path)
```
