---
title: "Seurat <-> AnnData (h5ad): Real-Data Workflow with Python Validation"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE,
  out.width  = "100%"
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
})
set.seed(42L)
```

## Overview

The h5ad format is the standard on-disk representation for
[scanpy](https://scanpy.readthedocs.io),
[scvi-tools](https://scvi-tools.org), and
[CELLxGENE](https://cellxgene.cziscience.com).
scConvert lets you move data between h5ad and Seurat objects without any
Python dependency: all I/O goes through
[hdf5r](https://cran.r-project.org/package=hdf5r).

This article walks through the full round-trip on the PBMC 3k dataset
(2,700 cells, 32,738 genes) and confirms the written h5ad is readable
and analysable in Python/scanpy.

---

## 1  Load h5ad into Seurat

`readH5AD()` reads the AnnData file and returns a `Seurat` object.
It maps `X` to the `data` layer (normalized counts when present) and
`raw/X` to the `counts` layer.

```{r read-h5ad}
h5ad_path <- "../pbmc3k.h5ad"

t0 <- proc.time()
pbmc <- readH5AD(h5ad_path, verbose = FALSE)
elapsed <- (proc.time() - t0)[["elapsed"]]
cat(sprintf("readH5AD: %.2fs | %d cells x %d genes\n",
            elapsed, ncol(pbmc), nrow(pbmc)))
```

```{r inspect-object}
pbmc
head(pbmc[[]], 4)
```

---

## 2  Standard Seurat analysis pipeline

We run the canonical single-cell workflow on the loaded object.
The h5ad may already contain embeddings from the source; re-running
the pipeline here ensures the Seurat object is fully self-consistent
and demonstrates that the imported data is analytically valid.

```{r seurat-pipeline}
set.seed(42L)
pbmc <- NormalizeData(pbmc, verbose = FALSE)
pbmc <- FindVariableFeatures(pbmc, nfeatures = 2000L, verbose = FALSE)
pbmc <- ScaleData(pbmc, verbose = FALSE)
pbmc <- RunPCA(pbmc, npcs = 30L, verbose = FALSE)
pbmc <- RunUMAP(pbmc, dims = 1:20, verbose = FALSE)
pbmc <- FindNeighbors(pbmc, dims = 1:20, verbose = FALSE)
pbmc <- FindClusters(pbmc, resolution = 0.5, verbose = FALSE)
```

```{r dimplot, fig.cap="PBMC 3k UMAP coloured by Seurat clusters after h5ad import."}
DimPlot(pbmc, reduction = "umap", label = TRUE, label.size = 4) +
  ggtitle("PBMC 3k: Seurat clusters (from h5ad import)") +
  theme(plot.title = element_text(hjust = 0.5))
```

---

## 3  Write Seurat back to h5ad

`writeH5AD()` serialises the Seurat object to a spec-compliant AnnData
file. Reductions are stored in `obsm`, graphs in `obsp`, and metadata in
`obs`.

```{r write-h5ad}
tmp_h5ad <- file.path(tempdir(), "pbmc3k_seurat.h5ad")

t0 <- proc.time()
writeH5AD(pbmc, filename = tmp_h5ad, overwrite = TRUE, verbose = FALSE)
elapsed <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("writeH5AD: %.2fs\n", elapsed))
cat(sprintf("Output: %.1f MB\n", file.size(tmp_h5ad) / 1e6))
```

---

## 4  Layer mapping reference

| Seurat slot | h5ad path | Notes |
|---|---|---|
| `counts` layer | `/raw/X` | raw integer counts |
| `data` layer | `/X` | normalised expression |
| `scale.data` layer | `/layers/scale.data` | scaled matrix if present |
| `meta.data` | `/obs` | cell-level annotations |
| `var` rownames + metadata | `/var` | feature annotations |
| `reductions` (e.g. PCA, UMAP) | `/obsm/X_pca`, `/obsm/X_umap` | cell embeddings |
| `graphs` (SNN, NN) | `/obsp/` | sparse adjacency matrices |
| `misc` | `/uns` | unstructured metadata |

---

## 5  Python validation

We read the written h5ad back in Python/scanpy to confirm the file is
correctly formed. This requires the `scverse` conda environment.

```{r python-setup}
library(reticulate)
Sys.setenv(NUMBA_THREADING_LAYER = "tbb", OMP_NUM_THREADS = "1")
use_condaenv("scverse")
```

```{python py-validate, eval=TRUE}
import anndata
import scanpy as sc

adata = anndata.read_h5ad(r.tmp_h5ad)
print(adata)
```

Run scanpy's standard pipeline on the imported object:

```{python py-pipeline, eval=TRUE, fig.cap="PBMC 3k UMAP from scanpy after round-trip through scConvert."}
sc.pp.normalize_total(adata, target_sum=1e4)
sc.pp.log1p(adata)
sc.pp.pca(adata, n_comps=30, random_state=42)
sc.pp.neighbors(adata, n_pcs=20, random_state=42)
sc.tl.leiden(adata, resolution=0.5, random_state=42)
sc.tl.umap(adata, random_state=42)
sc.pl.umap(adata, color="leiden", show=True)
```

```{python py-summary, eval=TRUE}
n_clusters = len(adata.obs["leiden"].unique())
print(
    f"Python validation: {adata.n_obs} cells x {adata.n_vars} genes, "
    f"{n_clusters} clusters"
)
```

---

## 6  Cleanup

```{r cleanup}
unlink(tmp_h5ad)
```
