---
title: "SpatialData Zarr: The scverse Spatial Standard"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE,
  out.width  = "100%"
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
})
set.seed(42L)
```

## Overview

[SpatialData](https://spatialdata.scverse.org) is the scverse standard for spatial
omics. A `.spatialdata.zarr` store is a Zarr directory that encodes expression tables,
spatial coordinates, shapes (polygons, circles), tissue images, and coordinate systems
in a unified hierarchy. All major spatial technologies now have SpatialData
representations via the [spatialdata-io](https://spatialdata-io.scverse.org) package.

scConvert reads and writes SpatialData Zarr stores natively in R, with no Python
dependency, supporting both zarr v2 and v3 formats.

---

## 1  Read a SpatialData store

`readSpatialData()` reads the `tables/table` AnnData from the store and returns a
Seurat object. Spatial coordinates, images, and coordinate-system metadata are
attached when present.

```{r read-spatialdata}
sd_path <- normalizePath("../blobs.spatialdata.zarr")

t0 <- proc.time()
blobs <- readSpatialData(sd_path, verbose = FALSE)
elapsed <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("Loaded: %d cells x %d features in %.2fs\n",
            ncol(blobs), nrow(blobs), elapsed))
cat(sprintf("Metadata columns: %s\n",
            paste(names(blobs[[]]), collapse = ", ")))
```

```{r inspect-spatialdata}
head(blobs[[]], 6)
```

The three features (`channel-0-sum`, `channel-1-sum`, `channel-2-sum`) correspond to
summarised intensity channels from synthetic image blobs. This structure is
representative of imaging-based spatial proteomics platforms (CODEX, IMC, CyCIF) where
channels map to protein markers rather than gene transcripts.

---

## 2  Export to h5ad

`SpatialDataToH5AD()` extracts the expression table and spatial coordinates and writes
a spec-compliant h5ad suitable for direct import into scanpy or squidpy.

```{r spatialdata-to-h5ad}
h5ad_out <- file.path(tempdir(), "blobs.h5ad")

t0 <- proc.time()
SpatialDataToH5AD(sd_path, h5ad_out, overwrite = TRUE, verbose = FALSE)
elapsed <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("SpatialData -> h5ad: %.2fs | %.1f KB\n",
            elapsed, file.size(h5ad_out) / 1e3))
```

---

## 3  Python validation with spatialdata

```{r python-setup, include=FALSE}
library(reticulate)
Sys.setenv(NUMBA_THREADING_LAYER = "tbb", OMP_NUM_THREADS = "1")
use_condaenv("scverse")
```

```{python py-spatialdata, eval=TRUE}
import spatialdata
import matplotlib
matplotlib.use('Agg')

sdata = spatialdata.read_zarr(r.sd_path)
print(sdata)
print(f"\nTables : {list(sdata.tables.keys())}")
print(f"Images : {list(sdata.images.keys())}")
print(f"Shapes : {list(sdata.shapes.keys())}")
print(f"Points : {list(sdata.points.keys())}")
```

```{python py-h5ad-validate, eval=TRUE}
import anndata
adata = anndata.read_h5ad(r.h5ad_out)
print(adata)
print(f"obs columns: {list(adata.obs.columns)}")
```

---

## 4  Format converters reference

All four SpatialData format converters are exported. The `scConvert()` dispatcher
also handles `.spatialdata.zarr` paths automatically.

```{r direct-converters, eval=FALSE}
SpatialDataToH5AD(
  source    = "sample.spatialdata.zarr",
  dest      = "sample.h5ad",
  overwrite = TRUE
)

H5ADToSpatialData(
  source    = "sample.h5ad",
  dest      = "sample.spatialdata.zarr",
  overwrite = TRUE
)

SpatialDataToH5Seurat(
  source    = "sample.spatialdata.zarr",
  dest      = "sample.h5seurat",
  overwrite = TRUE
)

scConvert("sample.spatialdata.zarr", dest = "sample.h5ad")
```

---

## 5  Technology coverage

The following spatial technologies produce SpatialData stores via `spatialdata-io`.
scConvert interoperates with all of them.

| Technology | Vendor | Modality | spatialdata-io reader |
|------------|--------|----------|-----------------------|
| Visium / Visium HD | 10x Genomics | transcriptomics | `visium()` / `visium_hd()` |
| Xenium | 10x Genomics | transcriptomics (FISH) | `xenium()` |
| CosMx | NanoString | transcriptomics (FISH) | `cosmx()` |
| MERFISH | Vizgen | transcriptomics (FISH) | `merfish()` |
| Slide-seq | Broad Institute | transcriptomics | `slideseq()` |
| CODEX / PhenoCycler | Akoya Biosciences | proteomics | `codex()` |
| IMC | Fluidigm / Standard BioTools | proteomics | `imc()` |

---

## 6  Cleanup

```{r cleanup}
unlink(h5ad_out)
```
