---
title: "Multimodal CITE-seq with MuData (h5mu): Real CBMC Data"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE,
  out.width  = "100%"
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
})
set.seed(42L)
```

CITE-seq measures RNA and surface protein (ADT, antibody-derived tags)
simultaneously from the same cells, enabling joint transcriptomic and
proteomic profiling. The MuData h5mu format stores both modalities in a
single file, making it straightforward to exchange multimodal datasets
between Seurat (R) and muon/mudata (Python) without losing either
modality or the shared cell metadata.

scConvert reads and writes h5mu without any Python dependency, relying
entirely on hdf5r for HDF5 I/O.

## Load real CITE-seq data (CBMC)

The `cbmc_rna.h5seurat` file contains the cord blood mononuclear cell
(CBMC) CITE-seq dataset from Stoeckius et al. (2017): 8,617 cells with
20,501 RNA features and 10 ADT surface markers.

```{r load-data}
cbmc_path <- "../cbmc_rna.h5seurat"

t0      <- proc.time()
cbmc    <- readH5Seurat(cbmc_path)
elapsed <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("Loaded: %d cells in %.2fs\n", ncol(cbmc), elapsed))
cat(sprintf("Assays: %s\n", paste(names(cbmc@assays), collapse = ", ")))
cat(sprintf("RNA features: %d\n", nrow(cbmc[["RNA"]])))

adt_name <- intersect(names(cbmc@assays), c("ADT", "CITE", "PROT", "prot"))
if (length(adt_name) > 0) {
  cat(sprintf("ADT features (%s): %d\n",
              adt_name[1], nrow(cbmc[[adt_name[1]]])))
  cat(sprintf("ADT markers: %s\n",
              paste(rownames(cbmc[[adt_name[1]]]), collapse = ", ")))
}
```

## RNA analysis

Normalize RNA counts with log-normalization, select variable features,
run PCA, build a neighbor graph, cluster with the Leiden algorithm, and
compute a UMAP projection.

```{r rna-pipeline}
DefaultAssay(cbmc) <- "RNA"

cbmc <- NormalizeData(cbmc, verbose = FALSE)
cbmc <- FindVariableFeatures(cbmc, nfeatures = 2000, verbose = FALSE)
cbmc <- ScaleData(cbmc, verbose = FALSE)
cbmc <- RunPCA(cbmc, verbose = FALSE)
cbmc <- FindNeighbors(cbmc, dims = 1:30, verbose = FALSE)
cbmc <- FindClusters(cbmc, resolution = 0.8, verbose = FALSE)
cbmc <- RunUMAP(cbmc, dims = 1:30, verbose = FALSE)

cat(sprintf("RNA clusters: %d\n", length(unique(cbmc$seurat_clusters))))
```

```{r rna-umap}
DimPlot(cbmc, group.by = "seurat_clusters", label = TRUE, pt.size = 0.3) +
  labs(title = "CBMC CITE-seq: RNA clusters") +
  theme(plot.title = element_text(size = 12),
        axis.text  = element_text(color = "black"),
        axis.title = element_text(color = "black"))
```

If cell-type annotations are available from the original dataset, they
can be overlaid alongside cluster labels:

```{r rna-annotations}
if ("rna_annotations" %in% colnames(cbmc@meta.data)) {
  DimPlot(cbmc, group.by = "rna_annotations", label = TRUE,
          pt.size = 0.3, repel = TRUE) +
    labs(title = "CBMC: RNA cell-type annotations") +
    theme(plot.title  = element_text(size = 12),
          axis.text   = element_text(color = "black"),
          axis.title  = element_text(color = "black"),
          legend.text = element_text(size = 8))
}
```

## ADT protein expression

CLR (centered log-ratio) normalization is the recommended approach for
ADT data. After normalization the protein counts are projected onto the
UMAP computed from RNA to visualize protein expression in the RNA
embedding space.

```{r adt-normalize}
if (length(adt_name) > 0) {
  DefaultAssay(cbmc) <- adt_name[1]
  cbmc <- NormalizeData(cbmc,
                        normalization.method = "CLR",
                        margin = 2,
                        verbose = FALSE)
}
```

```{r adt-feature-plot, fig.height=5}
if (length(adt_name) > 0) {
  adt_features <- rownames(cbmc[[adt_name[1]]])

  # Select two markers covering distinct lineages: T cells (CD3) and monocytes (CD14)
  plot_features <- intersect(c("CD3", "CD14"), adt_features)
  if (length(plot_features) < 2) {
    plot_features <- adt_features[seq_len(min(2L, length(adt_features)))]
  }

  DefaultAssay(cbmc) <- adt_name[1]
  p <- FeaturePlot(cbmc, features = plot_features, ncol = 2,
                   reduction = "umap", min.cutoff = "q05")
  print(p)
}
```

## Export to h5mu

`writeH5MU()` serializes all Seurat assays as separate modalities in a
single h5mu file. Shared cell metadata, cluster labels, and dimensionality
reduction embeddings are written to the root-level `obs` and `obsm` groups.

```{r write-h5mu}
h5mu_path <- file.path(tempdir(), "cbmc.h5mu")

t0      <- proc.time()
writeH5MU(cbmc, h5mu_path, overwrite = TRUE)
elapsed <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("Wrote h5mu: %.2fs | %.1f MB\n",
            elapsed, file.size(h5mu_path) / 1e6))
```

## Python validation with mudata

The exported h5mu is read by mudata to confirm that both modalities are
present with the correct cell and feature counts.

```{r python-setup, include=FALSE}
library(reticulate)
Sys.setenv(NUMBA_THREADING_LAYER = "tbb", OMP_NUM_THREADS = "1")
use_condaenv("scverse")
```

```{python py-validate, eval=TRUE}
import mudata
import matplotlib
matplotlib.use('Agg')
import matplotlib.pyplot as plt

mdata = mudata.read_h5mu(r.h5mu_path)
print(mdata)
print(f"\nModalities: {list(mdata.mod.keys())}")
for name, mod in mdata.mod.items():
    print(f"  {name}: {mod.n_obs} cells x {mod.n_vars} features")
```

## Modality name mapping

Seurat assay names are lowercased when written to h5mu to follow the
MuData convention. The standard ADT assay is mapped to `prot`.

| Seurat assay | h5mu modality | Notes |
|---|---|---|
| RNA | rna | Standard RNA assay |
| ADT | prot | Antibody-derived tags |
| ATAC | atac | Chromatin accessibility |
| Other | lowercase name | Custom modalities |

To read the h5mu back into Seurat:

```{r round-trip}
cbmc_rt <- readH5MU(h5mu_path)
cat(sprintf("Round-trip: %d cells, assays: %s\n",
            ncol(cbmc_rt),
            paste(names(cbmc_rt@assays), collapse = ", ")))
```

## Related vignettes

- **[CITE-seq protein-RNA integration](multimodal-citeseq.html)** --
  Weighted nearest neighbor (WNN) integration of RNA and ADT modalities
- **[Getting started](../../getting-started.html)** -- single-format
  conversions and the `scConvert()` dispatcher

## Cleanup

```{r cleanup, include=FALSE}
unlink(h5mu_path)
```
