---
title: "CITE-seq Workflow: RNA + Protein with Real CBMC Data"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE,
  out.width  = "100%"
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
})
set.seed(42L)
```

## Introduction

CITE-seq (Cellular Indexing of Transcriptomes and Epitopes by Sequencing)
simultaneously profiles RNA and surface protein (antibody-derived tags, ADT)
from the same cells, enabling a more complete picture of cell identity than
RNA alone. Surface protein levels complement transcriptomics by capturing
post-transcriptional regulation and providing sharper separation for
immunophenotyping.

This article shows the full CITE-seq workflow using the CBMC dataset -- cord
blood mononuclear cells profiled with 13 antibodies -- and demonstrates how
scConvert exports the multi-modal result to h5mu for downstream use in Python's
[muon](https://muon.readthedocs.io) ecosystem.

---

## 1  Load CBMC data

`readH5Seurat()` reads the CBMC h5Seurat file and reconstructs both the RNA
and ADT assays. The ADT assay name varies between datasets; the code below
detects it automatically.

```{r load-cbmc}
cbmc <- readH5Seurat("../cbmc_rna.h5seurat")
cat(sprintf("Cells: %d\n", ncol(cbmc)))
cat(sprintf("Assays: %s\n", paste(names(cbmc@assays), collapse = ", ")))
cat(sprintf("RNA features: %d\n", nrow(cbmc[["RNA"]])))

adt_name <- names(cbmc@assays)[names(cbmc@assays) != "RNA"][1]
if (!is.na(adt_name)) {
  cat(sprintf("ADT assay '%s': %d features\n", adt_name, nrow(cbmc[[adt_name]])))
}
```

---

## 2  RNA analysis pipeline

We run the standard Seurat single-cell workflow on the RNA assay to obtain
clusters and a UMAP layout before integrating protein data.

```{r rna-pipeline}
DefaultAssay(cbmc) <- "RNA"
set.seed(42L)
cbmc <- NormalizeData(cbmc, verbose = FALSE)
cbmc <- FindVariableFeatures(cbmc, nfeatures = 2000L, verbose = FALSE)
cbmc <- ScaleData(cbmc, verbose = FALSE)
cbmc <- RunPCA(cbmc, npcs = 30L, verbose = FALSE)
cbmc <- FindNeighbors(cbmc, dims = 1:20, verbose = FALSE)
cbmc <- FindClusters(cbmc, resolution = 0.5, verbose = FALSE)
cbmc <- RunUMAP(cbmc, dims = 1:20, verbose = FALSE)
```

```{r rna-dimplot, fig.cap="CBMC UMAP coloured by RNA-based Seurat clusters."}
DimPlot(cbmc, reduction = "umap", group.by = "seurat_clusters",
        label = TRUE, label.size = 4, pt.size = 0.5) +
  ggtitle("CBMC: RNA clusters") +
  theme(plot.title = element_text(hjust = 0.5))
```

---

## 3  Protein (ADT) analysis

CLR (centered log-ratio) normalization is the standard approach for ADT data.
It accounts for compositional effects and differences in antibody capture
efficiency.

```{r adt-normalize}
DefaultAssay(cbmc) <- adt_name
cbmc <- NormalizeData(cbmc, normalization.method = "CLR", margin = 2,
                      verbose = FALSE)
```

```{r adt-featureplot, fig.width=10, fig.height=8, fig.cap="ADT protein levels on the RNA UMAP."}
adt_features <- rownames(cbmc[[adt_name]])
plot_features <- head(adt_features, 4L)

FeaturePlot(cbmc, features = plot_features, ncol = 2,
            reduction = "umap", pt.size = 0.3,
            order = TRUE)
```

```{r adt-vlnplot, fig.cap="ADT expression of the first marker across RNA clusters."}
DefaultAssay(cbmc) <- "RNA"
first_adt <- plot_features[1]
DefaultAssay(cbmc) <- adt_name

VlnPlot(cbmc, features = first_adt, group.by = "seurat_clusters",
        pt.size = 0) +
  ggtitle(sprintf("%s protein across RNA clusters", first_adt)) +
  NoLegend()
```

---

## 4  WNN multi-modal integration

Weighted Nearest Neighbor (WNN) integrates RNA and ADT by learning per-cell
modality weights. Cells where protein provides additional discriminatory power
receive higher ADT weight; cells where RNA is more informative are
RNA-dominant. The resulting WNN UMAP captures cell state more accurately than
either modality alone.

```{r wnn-integration}
DefaultAssay(cbmc) <- "RNA"
set.seed(42L)
cbmc <- RunPCA(cbmc, verbose = FALSE)

DefaultAssay(cbmc) <- adt_name
VariableFeatures(cbmc) <- rownames(cbmc[[adt_name]])
cbmc <- ScaleData(cbmc, verbose = FALSE)
n_adt <- nrow(cbmc[[adt_name]])
adt_dims <- min(18L, n_adt - 1L)
cbmc <- RunPCA(cbmc, features = rownames(cbmc[[adt_name]]),
               reduction.name = "apca", verbose = FALSE)

cbmc <- FindMultiModalNeighbors(
  cbmc,
  reduction.list = list("pca", "apca"),
  dims.list      = list(1:30, seq_len(adt_dims)),
  verbose        = FALSE
)

cbmc <- RunUMAP(
  cbmc,
  nn.name       = "weighted.nn",
  reduction.name = "wnn.umap",
  reduction.key  = "wnnUMAP_",
  verbose        = FALSE
)
```

```{r wnn-dimplot, fig.cap="WNN UMAP integrating RNA and ADT modalities."}
DimPlot(cbmc, reduction = "wnn.umap", label = TRUE,
        label.size = 4, pt.size = 0.5) +
  ggtitle("CBMC: WNN UMAP (RNA + ADT)") +
  theme(plot.title = element_text(hjust = 0.5))
```

> **Note:** WNN UMAP integrates both modalities for a more accurate
> representation of cell state. Clusters that are indistinguishable by RNA
> alone may separate when surface protein levels are incorporated.

---

## 5  Export to h5mu

`writeH5MU()` writes the full multi-modal Seurat object to the MuData (.h5mu)
format, preserving both RNA and ADT assays in a single file. This is the
recommended format for sharing CITE-seq data with Python collaborators.

```{r write-h5mu}
DefaultAssay(cbmc) <- "RNA"
h5mu_path <- file.path(tempdir(), "cbmc_wnn.h5mu")

t0      <- proc.time()
writeH5MU(cbmc, h5mu_path, overwrite = TRUE)
elapsed <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("Wrote h5mu: %.2fs | %.1f MB\n",
            elapsed, file.size(h5mu_path) / 1e6))
```

---

## 6  Python validation

We read the exported h5mu in Python via mudata to confirm both modalities are
intact and correctly structured.

```{r python-setup}
library(reticulate)
Sys.setenv(NUMBA_THREADING_LAYER = "tbb", OMP_NUM_THREADS = "1")
use_condaenv("scverse")
```

```{python py-validate, eval=TRUE}
import mudata
import matplotlib
matplotlib.use('Agg')

mdata = mudata.read_h5mu(r.h5mu_path)
print(mdata)
for name, mod in mdata.mod.items():
    print(f"  {name}: {mod.n_obs} cells x {mod.n_vars} features")
```

---

## 7  Cleanup

```{r cleanup}
unlink(h5mu_path)
```
