---
title: "scATAC-seq: Converting Peak-by-Cell Matrices to h5ad"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE,
  out.width  = "100%"
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
})
set.seed(42L)
```

## Introduction

Single-cell ATAC-seq measures chromatin accessibility at single-cell
resolution. The core data structure -- a **peak-by-cell count matrix** -- is
structurally identical to a gene-by-cell RNA matrix. scConvert converts this
matrix alongside cell metadata, embeddings, and cluster labels, enabling
interoperability between R (Seurat/Signac) and Python (episcanpy, snapATAC2).

This article demonstrates the conversion mechanics using a CRC (colorectal
cancer) sample as a proxy. The matrix structure and I/O paths are the same
whether rows represent genes or ATAC peaks.

---

## 1  Load data

```{r load-crc}
t0   <- proc.time()
crc  <- readH5AD("../crc_sample.h5ad", verbose = FALSE)
elapsed <- (proc.time() - t0)[["elapsed"]]
cat(sprintf("Loaded: %d cells x %d features in %.2fs\n",
            ncol(crc), nrow(crc), elapsed))
```

---

## 2  Brief analysis pipeline

We run a standard Seurat workflow to produce clusters and a UMAP embedding
for visualization.

```{r pipeline}
set.seed(42L)
crc <- NormalizeData(crc, verbose = FALSE)
crc <- FindVariableFeatures(crc, nfeatures = 2000L, verbose = FALSE)
crc <- ScaleData(crc, verbose = FALSE)
crc <- RunPCA(crc, npcs = 30L, verbose = FALSE)
crc <- FindNeighbors(crc, dims = 1:20, verbose = FALSE)
crc <- FindClusters(crc, resolution = 0.5, verbose = FALSE)
crc <- RunUMAP(crc, dims = 1:20, verbose = FALSE)
```

```{r dimplot, fig.cap="CRC sample UMAP coloured by Seurat clusters."}
DimPlot(crc, reduction = "umap", label = TRUE, label.size = 4, pt.size = 0.5) +
  ggtitle("CRC sample: Seurat clusters") +
  theme(plot.title = element_text(hjust = 0.5))
```

---

## 3  Export to h5ad

`writeH5AD()` serialises the Seurat object to a spec-compliant AnnData file.
Embeddings are stored in `obsm`, sparse matrices in `X`, and metadata in `obs`.

```{r write-h5ad}
h5ad_out <- file.path(tempdir(), "crc_atac_demo.h5ad")

t0      <- proc.time()
writeH5AD(crc, h5ad_out, overwrite = TRUE)
elapsed <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("Wrote h5ad: %.2fs | %.1f MB\n",
            elapsed, file.size(h5ad_out) / 1e6))
```

---

## 4  Python validation

```{r python-setup}
library(reticulate)
Sys.setenv(NUMBA_THREADING_LAYER = "tbb", OMP_NUM_THREADS = "1")
use_condaenv("scverse")
```

```{python py-validate, eval=TRUE}
import anndata
import matplotlib
matplotlib.use('Agg')
import matplotlib.pyplot as plt

adata = anndata.read_h5ad(r.h5ad_out)
print(f"Validated: {adata.n_obs} cells x {adata.n_vars} features")
print(f"Obs columns: {list(adata.obs.columns[:6])}")
print(f"Embeddings: {list(adata.obsm.keys())}")
```

---

## 5  What scConvert preserves vs. what needs separate handling

For real scATAC-seq data, scConvert preserves the core data matrix and
annotations. Signac-specific components that live outside the standard
AnnData/h5ad schema require separate handling.

### Preserved by scConvert

| Component | Seurat / Signac | h5ad path |
|-----------|----------------|-----------|
| Peak count matrix | `counts` layer | `/X` (or `/raw/X`) |
| Cell metadata | `meta.data` | `/obs` |
| Peak metadata | `meta.features` | `/var` |
| LSI / PCA embeddings | `reductions` | `/obsm/X_lsi`, `/obsm/X_pca` |
| UMAP coordinates | `reductions$umap` | `/obsm/X_umap` |
| Cluster labels | `meta.data$seurat_clusters` | `/obs/seurat_clusters` |

### Needs separate handling

| Component | Why | Workaround |
|-----------|-----|------------|
| Fragment files | Large tabix files, not part of h5ad | Copy `fragments.tsv.gz` + `.tbi` separately |
| Gene annotations | GRanges object (R-specific) | Reload from EnsDb after import |
| Motif matrices | Signac-specific slot | Recompute with `AddMotifs()` |
| ChromatinAssay class | R S4 class, not encodable in h5ad | Upgrade with `CreateChromatinAssay()` |

---

## 6  Upgrading to ChromatinAssay after import

After loading an h5ad file that contains peak data, you can upgrade the
standard Seurat assay to a Signac `ChromatinAssay` for peak-aware analyses
such as motif enrichment and TF-IDF normalization.

```{r signac-upgrade, eval=FALSE}
suppressPackageStartupMessages(library(Signac))

atac <- readH5AD("peaks.h5ad")

peak_counts <- GetAssayData(atac, layer = "counts")
atac[["peaks"]] <- CreateChromatinAssay(
  counts      = peak_counts,
  sep         = c("-", "-"),
  min.cells   = 0,
  min.features = 0
)

# Optionally re-attach genome annotations:
# library(EnsDb.Hsapiens.v86)
# Annotation(atac) <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)
# Fragments(atac) <- CreateFragmentObject("fragments.tsv.gz")
```

---

## 7  Cleanup

```{r cleanup}
unlink(h5ad_out)
```
