---
title: "CELLxGENE Census: Real-world Atlas Query"
description: >
  Stream CD4 T cells from the 60M-cell CELLxGENE Census directly into
  Seurat, run a standard analysis pipeline, and export to h5ad — no
  intermediate download required.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CELLxGENE Census: Real-world Atlas Query}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
.eval <- !nzchar(Sys.getenv("CI"))
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = .eval,
  fig.width = 7,
  fig.height = 5,
  message   = FALSE,
  warning   = FALSE
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
})
set.seed(42L)
```

## Overview

[CELLxGENE Census](https://chanzuckerberg.github.io/cellxgene-census/)
hosts 60M+ cells across 900+ datasets as a single queryable
[TileDB-SOMA](https://github.com/single-cell-data/TileDB-SOMA) experiment
on public S3. This article demonstrates the two-step pattern that
scConvert enables:

1. Issue a Census slice query via `cellxgene.census`
2. Pass the URI directly to `readSOMA()` — Seurat object materialised on
   the fly, no intermediate file written

We then run a standard Seurat analysis pipeline and export to h5ad for
Python interoperability.

### Prerequisites

```r
install.packages("cellxgene.census",
  repos = c(
    "https://chanzuckerberg.r-universe.dev",
    "https://cloud.r-project.org"
  )
)
install.packages("tiledbsoma",
  repos = c(
    "https://tiledb-inc.r-universe.dev",
    "https://cloud.r-project.org"
  )
)
```

```{r pkg-check, include = FALSE}
has_census <- requireNamespace("cellxgene.census", quietly = TRUE) &&
              requireNamespace("tiledbsoma",        quietly = TRUE)
```

## Step 1 — Open Census and obtain the slice URI

```{r open-census}
library(cellxgene.census)
library(tiledbsoma)

census <- open_soma(census_version = "stable")
human_uri <- census$get("census_data")$get("homo_sapiens")$uri
cat("Human Census URI:", human_uri, "\n")
```

## Step 2 — Stream CD4 T cells into Seurat

`readSOMA()` opens the URI with tiledbsoma, applies the value filter
server-side (only matching chunks are fetched over the wire), and
materialises a Seurat object.

```{r read-soma}
obs_filter <- paste(
  "cell_type == 'CD4-positive, alpha-beta T cell'",
  "& tissue_general == 'blood'",
  "& is_primary_data == True"
)

cd4 <- readSOMA(
  uri         = human_uri,
  measurement = "RNA",
  obs_query   = obs_filter
)

census$close()
cat("Loaded:", ncol(cd4), "cells x", nrow(cd4), "genes\n")
cd4
```

```{r donor-summary}
donors <- table(cd4$donor_id)
cat("Donors:", length(donors), "\n")
cat("Datasets:", length(unique(cd4$dataset_id)), "\n")
```

## Step 3 — Standard Seurat pipeline

```{r seurat-pipeline}
cd4 <- NormalizeData(cd4, verbose = FALSE)
cd4 <- FindVariableFeatures(cd4, nfeatures = 2000L, verbose = FALSE)
cd4 <- ScaleData(cd4, verbose = FALSE)
cd4 <- RunPCA(cd4, npcs = 30L, verbose = FALSE)
cd4 <- RunUMAP(cd4, dims = 1:20, verbose = FALSE)
cd4 <- FindNeighbors(cd4, dims = 1:20, verbose = FALSE)
cd4 <- FindClusters(cd4, resolution = 0.4, verbose = FALSE)
```

## Step 4 — Visualisation

### UMAP coloured by cluster

```{r umap-cluster, fig.height = 5}
DimPlot(cd4, reduction = "umap", label = TRUE, pt.size = 0.4) +
  ggtitle("CD4 T cells (CELLxGENE Census, blood)") +
  NoLegend()
```

### Key marker expression

CD4 itself, the IL7R memory marker, and CCR7 (naive/central-memory).

```{r feature-plot, fig.height = 4}
FeaturePlot(
  cd4,
  features  = c("CD4", "IL7R", "CCR7", "FOXP3"),
  ncol      = 2L,
  pt.size   = 0.3,
  order     = TRUE
)
```

### Donor composition per cluster

```{r donor-heatmap, fig.height = 5}
donor_mat <- table(
  cluster = Idents(cd4),
  donor   = cd4$donor_id
)
donor_frac <- sweep(donor_mat, 1L, rowSums(donor_mat), "/")
pheatmap::pheatmap(
  donor_frac,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  main         = "Donor fraction per cluster",
  fontsize      = 8L
)
```

## Step 5 — Export to h5ad

```{r write-h5ad}
h5ad_path <- file.path(tempdir(), "census_cd4.h5ad")
writeH5AD(cd4, h5ad_path)
cat("Written:", h5ad_path, "\n")
cat("File size:", round(file.info(h5ad_path)$size / 1e6, 1), "MB\n")
```

### Verify with readH5AD

```{r roundtrip}
cd4_rt <- readH5AD(h5ad_path)
stopifnot(
  ncol(cd4_rt)  == ncol(cd4),
  nrow(cd4_rt)  == nrow(cd4)
)
cat("Round-trip OK:", ncol(cd4_rt), "cells x", nrow(cd4_rt), "genes\n")
```

```{r cleanup, include = FALSE}
unlink(h5ad_path)
```

## Key takeaways

- **No download required**: tiledbsoma streams only the requested cells
  via S3 byte-range requests; the full 60M-cell Census never touches disk.
- **Two-step pattern**: `open_soma()` + `readSOMA()` is the recommended
  route for Census queries; scConvert handles the SOMA-to-Seurat conversion.
- **Format portability**: `writeH5AD()` produces a scanpy-compatible h5ad
  from any Seurat object in one call, enabling downstream Python analysis
  without reloading from Census.

## Session Info

```{r session-info}
sessionInfo()
```
