---
title: "CELLxGENE Census: Real-world Atlas Query"
description: >
  Stream CD4 T cells from the 60M-cell CELLxGENE Census, run a standard
  Seurat analysis pipeline, and export to h5ad for Python interoperability
  -- no intermediate download required.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CELLxGENE Census: Real-world Atlas Query}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
})
set.seed(42L)
```

## Overview

[CELLxGENE Census](https://chanzuckerberg.github.io/cellxgene-census/)
hosts 60M+ cells across 900+ datasets as a single queryable
[TileDB-SOMA](https://github.com/single-cell-data/TileDB-SOMA) experiment
on public S3. This article shows a practical end-to-end workflow:

1. Query CD4 T cells from Census via `cellxgene.census::get_seurat()`
2. Run a standard Seurat analysis pipeline on the result
3. Export to h5ad with `scConvert::writeH5AD()` for Python colleagues

The Census query handles S3 streaming internally; no full-dataset download
occurs.

### Prerequisites

```r
install.packages("tiledb",
  repos = c("https://tiledb-inc.r-universe.dev", "https://cloud.r-project.org"))
install.packages("tiledbsoma",
  repos = c("https://tiledb-inc.r-universe.dev", "https://cloud.r-project.org"))
install.packages("cellxgene.census",
  repos = c("https://chanzuckerberg.r-universe.dev", "https://cloud.r-project.org"))
```

## Step 1 — Query Census

```{r open-census}
library(cellxgene.census)

census <- open_soma(census_version = "stable")
```

```{r query-census}
cd4_raw <- get_seurat(
  census,
  organism         = "Homo sapiens",
  obs_value_filter = paste(
    "cell_type == 'CD4-positive, alpha-beta T cell'",
    "& tissue_general == 'blood'",
    "& is_primary_data == True"
  ),
  obs_column_names = c(
    "cell_type", "donor_id", "dataset_id",
    "sex", "disease", "tissue_general"
  ),
  var_column_names = c("feature_id", "feature_name")
)

census$close()

# Census returns Ensembl IDs as rownames; rename to gene symbols for usability
counts <- GetAssayData(cd4_raw, assay = "RNA", layer = "counts")
rownames(counts) <- make.unique(cd4_raw[["RNA"]][[]]$feature_name)
cd4 <- CreateSeuratObject(counts = counts, meta.data = cd4_raw[[]])
rm(cd4_raw, counts)

cat("Loaded:", ncol(cd4), "cells x", nrow(cd4), "genes\n")
cat("Donors:", length(unique(cd4$donor_id)), "\n")
cat("Datasets:", length(unique(cd4$dataset_id)), "\n")
cd4
```

## Step 2 — Seurat pipeline

```{r seurat-pipeline}
cd4 <- NormalizeData(cd4, verbose = FALSE)
cd4 <- FindVariableFeatures(cd4, nfeatures = 2000L, verbose = FALSE)
cd4 <- ScaleData(cd4, verbose = FALSE)
cd4 <- RunPCA(cd4, npcs = 30L, verbose = FALSE)
cd4 <- RunUMAP(cd4, dims = 1:20, verbose = FALSE)
cd4 <- FindNeighbors(cd4, dims = 1:20, verbose = FALSE)
cd4 <- FindClusters(cd4, resolution = 0.4, verbose = FALSE)
```

## Step 3 — Visualisation

### UMAP coloured by cluster

```{r umap-cluster, fig.height = 5}
DimPlot(cd4, reduction = "umap", label = TRUE, pt.size = 0.3) +
  ggtitle(sprintf("CD4 T cells from CELLxGENE Census (%s cells)", ncol(cd4))) +
  NoLegend()
```

### Key marker expression

CD4, IL7R (memory), CCR7 (naive / central-memory), and FOXP3 (Treg).

```{r feature-plot, fig.height = 5}
FeaturePlot(
  cd4,
  features = c("CD4", "IL7R", "CCR7", "FOXP3"),
  ncol     = 2L,
  pt.size  = 0.2,
  order    = TRUE
)
```

### Disease composition

```{r disease-bar, fig.height = 4}
disease_counts <- sort(table(cd4$disease), decreasing = TRUE)
df <- data.frame(
  disease = names(disease_counts),
  n       = as.integer(disease_counts)
)
ggplot(df, aes(x = reorder(disease, n), y = n)) +
  geom_col(fill = "#4E79A7") +
  coord_flip() +
  labs(x = NULL, y = "Cells", title = "Disease annotation (Census metadata)") +
  theme_classic(base_size = 11)
```

## Step 4 — Export to h5ad

```{r write-h5ad}
h5ad_path <- file.path(tempdir(), "census_cd4.h5ad")
writeH5AD(cd4, h5ad_path)
cat("Written:", h5ad_path, "\n")
cat("File size:", round(file.info(h5ad_path)$size / 1e6, 1L), "MB\n")
```

### Verify round-trip

```{r roundtrip}
cd4_rt <- readH5AD(h5ad_path)
stopifnot(
  ncol(cd4_rt) == ncol(cd4),
  nrow(cd4_rt) == nrow(cd4)
)
cat("Round-trip OK:", ncol(cd4_rt), "cells x", nrow(cd4_rt), "genes\n")
cat("Clusters preserved:", all(cd4_rt$seurat_clusters == cd4$seurat_clusters), "\n")
```

```{r cleanup, include = FALSE}
unlink(h5ad_path)
```

## Key takeaways

- **No download**: `get_seurat()` streams only matching cells via S3
  byte-range reads across `r length(unique(cd4$dataset_id))` datasets.
- **Metadata preserved**: Census obs columns (`donor_id`, `disease`,
  `tissue_general`) survive the query and are written into the h5ad obs
  dataframe by `writeH5AD()`.
- **One-call export**: `writeH5AD()` converts any Seurat object to a
  scanpy-compatible h5ad, preserving counts, normalised expression,
  embeddings, and cluster labels.

## Session Info

```{r session-info}
sessionInfo()
```
