---
title: "CELLxGENE Census: Query 60M Cells and Export to scverse"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE,
  out.width  = "100%"
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
})
set.seed(42L)
```

## Introduction

[CELLxGENE Census](https://chanzuckerberg.github.io/cellxgene-census/)
provides programmatic access to over 60 million cells from hundreds of
publicly deposited datasets. The
[cellxgene.census](https://chanzuckerberg.github.io/cellxgene-census/r/index.html)
R package exposes this corpus through TileDB-SOMA, enabling cell-type- and
tissue-specific queries that return a ready-to-use Seurat object -- no manual
download of individual datasets required.

This article queries CD4+ T cells from peripheral blood, runs a standard
Seurat analysis, and exports the result to h5ad for use in Python's scverse
ecosystem via scConvert.

---

## 1  Prerequisites

The Census packages are not on CRAN; install them from their respective
R-universes.

```{r install-prereqs, eval=FALSE}
install.packages(
  "tiledb",
  repos = c("https://tiledb-inc.r-universe.dev",
            "https://cloud.r-project.org")
)
install.packages(
  "tiledbsoma",
  repos = c("https://tiledb-inc.r-universe.dev",
            "https://cloud.r-project.org")
)
install.packages(
  "cellxgene.census",
  repos = c("https://chanzuckerberg.r-universe.dev",
            "https://cloud.r-project.org")
)
```

---

## 2  Open Census and query CD4 T cells

`open_soma()` opens a connection to the stable Census release without
downloading data locally. `get_seurat()` executes a server-side filter and
streams only the matching cells and their expression values.

```{r census-query}
suppressPackageStartupMessages(library(cellxgene.census))

t0 <- proc.time()

census <- open_soma(census_version = "stable")

cd4_raw <- get_seurat(
  census,
  organism = "Homo sapiens",
  obs_value_filter = paste(
    "cell_type == 'CD4-positive, alpha-beta T cell'",
    "& tissue_general == 'blood'",
    "& is_primary_data == True"
  ),
  obs_column_names = c("cell_type", "donor_id", "dataset_id",
                       "sex", "disease", "tissue_general"),
  var_column_names = c("feature_id", "feature_name")
)

census$close()

counts    <- GetAssayData(cd4_raw, assay = "RNA", layer = "counts")
rownames(counts) <- make.unique(cd4_raw[["RNA"]][[]]$feature_name)
cd4       <- CreateSeuratObject(counts = counts, meta.data = cd4_raw[[]])
rm(cd4_raw, counts)

query_time <- (proc.time() - t0)[["elapsed"]]
cat(sprintf("Queried %d cells from Census in %.1fs\n", ncol(cd4), query_time))
```

---

## 3  Seurat analysis pipeline

```{r seurat-pipeline}
set.seed(42L)
cd4 <- NormalizeData(cd4, verbose = FALSE)
cd4 <- FindVariableFeatures(cd4, nfeatures = 2000L, verbose = FALSE)
cd4 <- ScaleData(cd4, verbose = FALSE)
cd4 <- RunPCA(cd4, npcs = 30L, verbose = FALSE)
cd4 <- RunUMAP(cd4, dims = 1:20, verbose = FALSE)
cd4 <- FindNeighbors(cd4, dims = 1:20, verbose = FALSE)
cd4 <- FindClusters(cd4, resolution = 0.5, verbose = FALSE)
```

---

## 4  Visualizations

```{r dimplot, fig.cap="UMAP of CD4+ T cells from CELLxGENE Census, coloured by cluster."}
DimPlot(cd4, label = TRUE, label.size = 4, pt.size = 0.3) +
  ggtitle("CD4+ T cells from CELLxGENE Census") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r featureplot, fig.width=10, fig.height=8, fig.cap="Canonical CD4 T cell marker genes."}
markers <- intersect(c("CD4", "IL7R", "CCR7", "FOXP3"), rownames(cd4))
if (length(markers) >= 1L) {
  FeaturePlot(cd4, features = markers, ncol = 2, pt.size = 0.3, order = TRUE)
}
```

```{r disease-barplot, fig.cap="Disease composition of the CD4 T cell query result."}
disease_counts <- as.data.frame(table(cd4$disease))
colnames(disease_counts) <- c("disease", "n_cells")
disease_counts <- disease_counts[order(-disease_counts$n_cells), ]
disease_counts$disease <- factor(disease_counts$disease,
                                  levels = disease_counts$disease)

ggplot(disease_counts, aes(x = disease, y = n_cells, fill = disease)) +
  geom_col(show.legend = FALSE) +
  labs(x = "Disease annotation", y = "Number of cells",
       title = "CD4 T cells by disease condition") +
  theme_classic() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1, size = 9),
        plot.title   = element_text(hjust = 0.5))
```

---

## 5  Export to h5ad

`writeH5AD()` writes the Census-derived Seurat object to a spec-compliant
AnnData file. All cluster labels, embeddings, and metadata columns are
preserved.

```{r write-h5ad}
h5ad_path <- file.path(tempdir(), "census_cd4.h5ad")

t0          <- proc.time()
writeH5AD(cd4, h5ad_path)
export_time <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("Exported to h5ad: %.2fs | %.1f MB\n",
            export_time, file.size(h5ad_path) / 1e6))
```

---

## 6  Python validation with scanpy

```{r python-setup}
library(reticulate)
Sys.setenv(NUMBA_THREADING_LAYER = "tbb", OMP_NUM_THREADS = "1")
use_condaenv("scverse")
```

```{python py-validate, eval=TRUE}
import anndata
import scanpy as sc
import matplotlib
matplotlib.use('Agg')
import matplotlib.pyplot as plt

adata = anndata.read_h5ad(r.h5ad_path)
print(f"Python reads Census export: {adata.n_obs} cells x {adata.n_vars} genes")
print(f"Cluster labels preserved: {'seurat_clusters' in adata.obs.columns}")
print(f"Disease annotations: {adata.obs['disease'].nunique()} unique values")

sc.pl.umap(adata, color="seurat_clusters", show=False)
plt.title("CD4 T cells from CELLxGENE Census (Python/scanpy)")
plt.tight_layout()
plt.show()
```

---

## 7  Key takeaways

- **No bulk download required.** `get_seurat()` streams only the cells matching
  the query filter from the Census server; the full 60M-cell corpus never
  touches local disk.
- **Metadata preserved end-to-end.** Donor IDs, sex, disease annotations,
  tissue labels, and Seurat cluster assignments all survive the
  Census -> Seurat -> h5ad -> Python round-trip.
- **One-call export.** `writeH5AD(cd4, path)` produces a scanpy-readable h5ad
  with no additional Python environment or manual reformatting needed.

---

## 8  Cleanup

```{r cleanup}
unlink(h5ad_path)
```
