---
title: "Spatial Transcriptomics: Visium Data with Seurat and squidpy"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE,
  out.width  = "100%"
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
})
set.seed(42L)
```

10x Genomics Visium captures spatially resolved gene expression by measuring
transcriptomes at defined spots arrayed across a tissue section. Each spot
retains its physical coordinates and is co-registered with a histology image,
enabling joint spatial-statistical analysis.

scConvert preserves every spatial component during conversion: spot coordinates,
high-resolution and low-resolution tissue images, and scale factors. The same
dataset can therefore be analyzed in Seurat (R) and squidpy (Python) without
any data loss or manual reformatting.

## Load real Visium data from h5ad

The `stxBrain.h5ad` file contains the 10x Genomics mouse brain anterior section
distributed with the Seurat spatial vignette (2,696 spots, 31,053 genes, ~70 MB).

```{r load-data}
h5ad_path <- "../stxBrain.h5ad"

t0      <- proc.time()
brain   <- readH5AD(h5ad_path, verbose = FALSE)
elapsed <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("Loaded: %d spots x %d genes in %.2fs\n",
            ncol(brain), nrow(brain), elapsed))
cat(sprintf("Images: %s\n",
            paste(names(brain@images), collapse = ", ")))
cat(sprintf("Reductions: %s\n",
            paste(names(brain@reductions), collapse = ", ")))
```

## Spatial expression plots

`Hpca` (hippocalcin) is a marker of hippocampal neurons; `Ttr` (transthyretin)
marks the choroid plexus. Plotting both on the tissue section confirms that
scConvert preserves coordinate registration with the histology image.

```{r spatial-feature-plot, fig.height=6}
SpatialFeaturePlot(brain, features = c("Hpca", "Ttr"), ncol = 2)
```

## Seurat analysis pipeline

A standard Seurat workflow — normalization, variable feature selection, PCA,
graph construction, clustering, and UMAP — is applied to the spatial data.
The spatial context does not require any modification of the pipeline.

```{r seurat-pipeline}
brain <- NormalizeData(brain, verbose = FALSE)
brain <- FindVariableFeatures(brain, verbose = FALSE)
brain <- ScaleData(brain, verbose = FALSE)
brain <- RunPCA(brain, verbose = FALSE)
brain <- FindNeighbors(brain, verbose = FALSE)
brain <- FindClusters(brain, verbose = FALSE)
brain <- RunUMAP(brain, dims = 1:30, verbose = FALSE)

cat(sprintf("Clusters: %d\n",
            length(unique(brain$seurat_clusters))))
```

### Spatial cluster map

Each spot is colored by its Leiden cluster identity, overlaid on the histology
image. Cluster boundaries align with anatomical structures visible in the tissue.

```{r spatial-dim-plot, fig.height=6}
SpatialDimPlot(brain, label = TRUE, label.size = 3)
```

### UMAP embedding

The UMAP projection separates clusters that map to distinct spatial regions,
consistent with region-specific transcriptional programs.

```{r umap-plot}
DimPlot(brain, reduction = "umap", label = TRUE, pt.size = 0.5) +
  labs(title = "UMAP: mouse brain anterior section") +
  theme(plot.title = element_text(size = 12))
```

## Export to h5ad for Python

`writeH5AD()` serializes the Seurat object including spatial coordinates, the
tissue image, scale factors, cluster labels, and dimensionality reduction
embeddings.

```{r write-h5ad}
out_h5ad <- file.path(tempdir(), "stxBrain_seurat.h5ad")

t0      <- proc.time()
writeH5AD(brain, out_h5ad, overwrite = TRUE)
elapsed <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("Wrote h5ad: %.2fs | %.1f MB\n",
            elapsed, file.size(out_h5ad) / 1e6))
```

## Python validation with squidpy

The exported h5ad is read by anndata and visualized with squidpy to confirm
that spatial coordinates and cluster labels survived the round-trip.

```{r python-setup, include=FALSE}
library(reticulate)
Sys.setenv(NUMBA_THREADING_LAYER = "tbb", OMP_NUM_THREADS = "1")
use_condaenv("scverse")
```

```{python py-validate, eval=TRUE}
import anndata
import matplotlib
matplotlib.use('Agg')
import matplotlib.pyplot as plt
import numpy as np

adata = anndata.read_h5ad(r.out_h5ad)
print(f"Loaded: {adata.n_obs} spots x {adata.n_vars} genes")
print(f"Has spatial: {'spatial' in adata.obsm}")
print(f"Clusters: {adata.obs['seurat_clusters'].nunique()}")

try:
    import squidpy as sq
    img_keys = [k for v in adata.uns.get("spatial", {}).values()
                for k in v.get("images", {}).keys()]
    chosen_key = "lowres" if "lowres" in img_keys else (img_keys[0] if img_keys else None)
    sq.pl.spatial_scatter(adata, color="seurat_clusters",
                          img_key=chosen_key, figsize=(6, 5))
except Exception:
    coords   = adata.obsm["spatial"]
    clusters = adata.obs["seurat_clusters"].astype("category")
    n_cl     = len(clusters.cat.categories)
    palette  = plt.cm.get_cmap("tab20", n_cl)
    fig, ax  = plt.subplots(figsize=(6, 5))
    for i, cl in enumerate(clusters.cat.categories):
        mask = clusters == cl
        ax.scatter(coords[mask, 0], coords[mask, 1],
                   c=[palette(i)], s=4, label=cl, alpha=0.8)
    ax.set_title("Spatial clusters: mouse brain anterior")
    ax.invert_yaxis()
    ax.axis("off")
    ax.legend(bbox_to_anchor=(1.02, 1), loc="upper left",
              fontsize=6, ncol=2, markerscale=2)
plt.tight_layout()
plt.show()
```

## What is preserved

Every spatial component required for downstream Python analysis is written
verbatim by scConvert.

| Component | h5ad location | Preserved by scConvert |
|---|---|:---:|
| Expression matrix | X / raw/X | Yes |
| Spot coordinates | obsm/spatial | Yes |
| Tissue image | uns/spatial/*/images | Yes |
| Scale factors | uns/spatial/*/scalefactors | Yes |
| Cell metadata | obs | Yes |
| Cluster labels | obs/seurat_clusters | Yes |
| PCA embedding | obsm/X_pca | Yes |
| UMAP embedding | obsm/X_umap | Yes |

Scale factors are written as HDF5 scalars (`shape=()`), matching the
convention expected by squidpy and scanpy for spot-size calculations.

## Other spatial technologies

scConvert handles a range of spatial platforms beyond Visium:

- **MERFISH / CosMx / Xenium**: sub-cellular resolution platforms with polygon
  or transcript-level coordinates stored as obs columns
- **Slide-seq v2**: bead-based spatial transcriptomics
- **Stereo-seq GEF**: BGI Genomics DNB-based spatial format, read via
  `readGEF()` and converted to h5ad
- **SpatialData Zarr**: the scverse cloud-native spatial container; see
  [Convert SpatialData](convert-spatialdata.html) for a dedicated walkthrough

All platforms share the same read/write API: `readH5AD()`, `writeH5AD()`, and
the `scConvert()` dispatcher.

## Cleanup

```{r cleanup, include=FALSE}
unlink(out_h5ad)
```
