---
title: "Spatial Transcriptomics: NanoString CosMx SMI"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 9,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE,
  out.width  = "100%"
)
suppressPackageStartupMessages({
  library(scConvert)
  library(Seurat)
  library(ggplot2)
  library(patchwork)
})
set.seed(42L)
```

NanoString CosMx SMI (Spatial Molecular Imager) is a sub-cellular resolution
imaging-based spatial transcriptomics platform. It combines in-situ hybridization
with cyclic fluorescence imaging to capture RNA and protein targets at single-cell
and sub-cellular resolution, with panels of up to 6,000+ genes across the full
tissue section.

CosMx data arrives as a flat-file bundle: an expression matrix, per-cell metadata
(cell morphology and co-registered immunofluorescence intensity), per-FOV position
coordinates, and optionally transcript-level positions and segmentation polygons.
Each FOV (field of view) is an independently imaged tile that is later assembled
into a whole-slide mosaic via the global pixel coordinates.

scConvert loads the full bundle in a single call — counts, centroids, and the
immunofluorescence metadata — and exports a squidpy-compatible h5ad without
manual reformatting.

---

## 1  Data

The demo dataset is the NanoString NSCLC Lung9 Rep1 sample from the public
NanoString data portal. It contains **91,972 cells** across **20 FOVs** profiled
with a **980-gene** panel, co-registered with membrane stain and four immune /
epithelial fluorescence channels (PanCK, CD45, CD3, DAPI).

Public data from the same study (Lung5 Rep2, 106K cells) is used in the
squidpy CosMx tutorial. Both samples share identical bundle layout and metadata
columns and are interchangeable as inputs to `LoadCosMx()`.

```{r data-download, eval=FALSE}
# Download and extract the flat-file bundle from NanoString's public S3
# (example shown for Lung5 Rep2, the squidpy tutorial dataset)
url <- paste0(
  "https://nanostring-public-share.s3.us-west-2.amazonaws.com/",
  "SMI-Compressed/Lung5_Rep2+SMI+Flat+data.tar.gz"
)
dest <- file.path(dirname(getwd()), "cosmx_lung5.tar.gz")
download.file(url, dest, method = "wget", extra = "-c")
untar(dest, exdir = file.path(dirname(getwd()), "cosmx_lung5"))
```

```{r data-path}
data_dir <- normalizePath(
  "../cosmx_lung9/Lung9_Rep1-Flat_files_and_images",
  mustWork = FALSE
)
```

---

## 2  Load with `LoadCosMx()`

`LoadCosMx()` validates the bundle layout, invokes `Seurat::ReadNanostring()`
for the count matrix and centroids, and automatically attaches the
`*metadata_file.csv` to the Seurat object. This is the step that
`Seurat::LoadNanostring()` itself skips when polygon files are absent.

```{r load-cosmx}
t0  <- proc.time()
obj <- LoadCosMx(data_dir, fov = "cosmx", verbose = FALSE)
elapsed <- (proc.time() - t0)[["elapsed"]]

cat(sprintf(
  "Loaded: %d cells x %d genes in %.1fs\n  FOVs: %d\n  Metadata cols: %d\n",
  ncol(obj), nrow(obj), elapsed,
  length(unique(obj$fov)),
  ncol(obj[[]])
))
```

```{r inspect-meta}
# First few rows of the per-cell metadata
head(obj[[]], 6)
```

The `fov` column identifies each cell's field of view (1–30 for this sample).
Columns with `Mean.*` / `Max.*` prefixes are co-registered immunofluorescence
intensities measured alongside the transcriptome:

| Column | Marker | Biology |
|---|---|---|
| `Mean.MembraneStain` | Membrane stain | Cell boundary reference |
| `Mean.PanCK` | Pan-cytokeratin | Epithelial / tumor cells |
| `Mean.CD45` | CD45 (PTPRC) | All immune cells |
| `Mean.CD3` | CD3 | T lymphocytes |
| `Mean.DAPI` | DAPI nuclear stain | Nuclear segmentation quality |

---

## 3  QC: transcript counts, gene counts, FOV coverage

```{r qc-plots, fig.height=4}
meta <- obj[[]]

p1 <- ggplot(meta, aes(x = nCount_Nanostring)) +
  geom_histogram(bins = 80, fill = "#3182bd", color = NA) +
  scale_x_log10() +
  labs(x = "Total transcripts per cell (log10)", y = "Cells") +
  theme_classic(base_size = 11)

p2 <- ggplot(meta, aes(x = nFeature_Nanostring)) +
  geom_histogram(bins = 80, fill = "#31a354", color = NA) +
  labs(x = "Unique genes per cell", y = "Cells") +
  theme_classic(base_size = 11)

fov_counts <- as.data.frame(table(fov = factor(
  meta$fov,
  levels = as.character(sort(unique(as.integer(meta$fov))))
)))
p3 <- ggplot(fov_counts, aes(x = fov, y = Freq)) +
  geom_col(fill = "#756bb1") +
  labs(x = "FOV", y = "Cells") +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1 + p2 + p3 + plot_layout(widths = c(1, 1, 2))
```

```{r qc-stats}
cat(sprintf(
  "Median transcripts/cell : %.0f\nMedian genes/cell       : %.0f\nCells per FOV (median)  : %.0f\n",
  median(meta$nCount_Nanostring),
  median(meta$nFeature_Nanostring),
  median(table(meta$fov))
))
```

---

## 4  Immunofluorescence channels on the spatial map

Co-registered protein markers are in `obj@meta.data`; plotting them on the
spatial map reveals tissue architecture without any clustering step.

```{r spatial-if, fig.height=5}
# Single FOV to keep rendering fast; FOV 1 is shown here
p_panck <- ImageFeaturePlot(obj, fov = "cosmx", features = "Mean.PanCK",
                             max.cutoff = "q95") +
  ggtitle("PanCK (epithelial/tumor)") +
  theme(plot.title = element_text(size = 11))

p_cd45 <- ImageFeaturePlot(obj, fov = "cosmx", features = "Mean.CD45",
                            max.cutoff = "q95") +
  ggtitle("CD45 (immune cells)") +
  theme(plot.title = element_text(size = 11))

p_panck | p_cd45
```

---

## 5  Export to h5ad

`scConvert()` serializes the Seurat object to a squidpy-compatible h5ad. The
output preserves:

- expression matrix in `X`
- cell centroids in `obsm/spatial` (global pixel coordinates)
- all per-cell metadata (FOV ID, morphology, IF intensities) in `obs`
- segmentation boundaries in `uns/spatial/cosmx/segmentation/` (when present)
- transcript-level positions in `uns/spatial/cosmx/molecules/` (when present)

```{r export-h5ad}
h5ad_out <- file.path(tempdir(), "cosmx_lung5.h5ad")

t0 <- proc.time()
scConvert(obj, h5ad_out, overwrite = TRUE)
elapsed <- (proc.time() - t0)[["elapsed"]]

cat(sprintf("Wrote h5ad: %.1fs  |  %.1f MB\n",
            elapsed, file.size(h5ad_out) / 1e6))
```

---

## 6  Python validation with squidpy

```{r python-setup, include=FALSE}
library(reticulate)
Sys.setenv(NUMBA_THREADING_LAYER = "tbb", OMP_NUM_THREADS = "1",
           KMP_DUPLICATE_LIB_OK = "TRUE")
use_condaenv("scverse")
```

```{python py-load}
import anndata
import scanpy as sc
import matplotlib
matplotlib.use("Agg")
import matplotlib.pyplot as plt

adata = anndata.read_h5ad(r.h5ad_out)
print(adata)
print(f"\nobs columns : {list(adata.obs.columns[:10])} ...")
print(f"obsm keys   : {list(adata.obsm.keys())}")
print(f"uns/spatial : {list(adata.uns.get('spatial', {}).keys())}")
```

The spatial coordinates live in `obsm["spatial"]` (global pixel coordinates,
matching the original `CenterX_global_px` / `CenterY_global_px` columns).

```{python py-spatial-fov, fig.height=5}
# Color all cells by FOV — each color is one tile of the tissue mosaic
adata.obsm["spatial"] = adata.obs[["CenterX_global_px",
                                    "CenterY_global_px"]].values

sc.pl.embedding(
    adata,
    basis    = "spatial",
    color    = "fov",
    s        = 2,
    frameon  = False,
    title    = "CosMx NSCLC: cells colored by FOV (20 tiles)",
    legend_loc = None
)
plt.tight_layout()
plt.show()
```

```{python py-panck, fig.height=5}
# PanCK intensity on the spatial map — highlights epithelial/tumor regions
sc.pl.embedding(
    adata,
    basis     = "spatial",
    color     = "Mean.PanCK",
    s         = 2,
    frameon   = False,
    color_map = "viridis",
    vmax      = "p95",
    title     = "PanCK intensity (epithelial / tumor cells)"
)
plt.tight_layout()
plt.show()
```

```{python py-cd45, fig.height=5}
# CD45 intensity — highlights immune infiltration
sc.pl.embedding(
    adata,
    basis     = "spatial",
    color     = "Mean.CD45",
    s         = 2,
    frameon   = False,
    color_map = "magma",
    vmax      = "p95",
    title     = "CD45 intensity (immune cells)"
)
plt.tight_layout()
plt.show()
```

---

## 7  What is preserved

| Component | h5ad location | Preserved |
|---|---|:---:|
| Expression matrix | X | Yes |
| Cell centroids (global) | obsm/spatial | Yes |
| FOV identifier | obs/fov | Yes |
| Cell morphology (Area, AspectRatio...) | obs | Yes |
| IF intensities (Mean.PanCK, Mean.CD45...) | obs | Yes |
| Global pixel coordinates | obs/CenterX\_global\_px, obs/CenterY\_global\_px | Yes |
| Segmentation polygons | uns/spatial/cosmx/segmentation/ | Yes (when present) |
| Transcript molecules | uns/spatial/cosmx/molecules/ | Yes (when present) |

Immunofluorescence columns are loaded from the `*metadata_file.csv` during
`LoadCosMx()` — they are not returned by `Seurat::ReadNanostring()` alone and
must be explicitly read and attached, which `LoadCosMx()` handles automatically.

---

## 8  Cleanup

```{r cleanup, include=FALSE}
unlink(h5ad_out)
```
